Orco regulates the circadian activity of pheromone-sensitive olfactory receptor neurons in hawkmoths

Version 1

Abstract

The mating behavior of nocturnal Manduca sexta hawkmoths is under strict temporal control. It is orchestrated via circadian- and ultradian oscillations in sex-pheromone stimuli as social zeitgeber. The extremely sensitive pheromone-detecting olfactory receptor neurons (ORNs) that innervate the long trichoid sensilla on the male’s antennae are peripheral circadian clocks. They express the transcriptional-translational feedback loop (TTFL) circadian clockwork, best characterized in Drosophila melanogaster. In hawkmoths, it is still unknown whether or how the ORN’s TTFL clockwork regulates the daily rhythms in pheromone-sensitivity and in temporal resolution of ultradian pheromone pulses as prerequisites to the temporal regulation of hawkmoth mating behavior.

We hypothesize that, rather than the slow TTFL clock, a more rapidly adaptive post-translational feedback loop (PTFL) clockwork, associated with the ORN’s plasma membrane, allows for temporal control of pheromone detection via generation of multiscale endogenous membrane potential oscillations. The potential oscillations of the PTFL clock could rapidly synchronize to oscillations of pheromone stimuli at different timescales, thus enable the prediction of stimulus patterns as a mechanism for active sensing. With in vivo long-term tip recordings of long trichoid sensilla of male hawkmoths, we analyzed the spontaneous spiking activity indicative of the ORNs’ endogenous membrane potential oscillations. Consistent with our hypothesis of a multiscale PTFL clock in hawkmoth ORNs, spontaneous spiking was modulated on ultradian and circadian timescales, with maximum activity at night. When we blocked the evolutionarily conserved olfactory receptor coreceptor (Orco), the circadian modulation was abolished but the ultradian frequency modulation of the spontaneous activity remained. Consistent with PTFL control, Orco was not under the circadian control of the TTFL clock. We could replicate the experimental data in a conductance-based computational model of Orco. In this model, Orco conductivity changed as a function of fluctuating 2nd messenger levels. This study demonstrates that a PTFL is sufficient to impose a circadian pattern on ORN sensitivity.

Introduction

Biological clocks generate endogenous physiological and behavioral oscillations on multiple timescales that persist in the absence of any external timing cues under constant conditions (Lamont and Amir, 2010). Such clocks are essential to generate internal rhythms and to permit their synchronization to external zeitgebers at timescales from milliseconds to years. These synchronized endogenous multiscale clocks thus provide both a continuous time frame for physiology and behavior - the organism’s “present time” - and enable organisms to predict and respond to both regular, periodic environmental changes (Stengl and Schneider, 2024). Even though endogenous clocks are widespread in prokaryotes and eukaryotes (Géron et al., 2024; Hall and Rosbash, 1993; Millar, 2016; Thaiss et al., 2014), current knowledge about how multiscale clocks are linked over widely diverging timescales remains sparse.

Insect circadian clock neurons, which are tightly connected to visual brain circuits, are amongst the best studied biological clocks (Hall and Rosbash, 1993; Nishiitsutsuji-Uwo and Pittendrigh, 1968; Reinhard et al., 2024). They regulate physiological and behavioral oscillations that underly sleep-wake rhythms tied to the 24 h light-dark cycles on Earth (Dubowy and Sehgal, 2017). Circadian clock neurons possess a molecular clockwork of transcriptional-translational feedback loops (TTFLs), which generate endogenous oscillations of clock gene mRNA and protein levels with a cycle period of ∼24 h (Hardin, 2011). It is assumed for both insects and mammals that this TTFL comprises the master clockwork that regulates all physiological and behavioral rhythms in the circadian time range through genetically controlled output pathways (Hastings et al., 2020).

Molecular TTFL clockworks were not only found in central brain clock neurons but also in sensory neurons, such as olfactory receptor neurons (ORNs) in insect antennae, where they play a role in odor-dependent circadian behaviors like feeding, pollination, reproduction, and navigation (Flecke et al., 2010; Flecke and Stengl, 2009; Krishnan et al., 1999; Merlin et al., 2009, 2006; Page and Koelling, 2003; Plautz et al., 1997; Sauman and Reppert, 1998; Schendzielorz et al., 2015; Schuckel et al., 2007; Tanoue et al., 2004; Zhou et al., 2005). Little is known about how the TTFL clockwork of insect ORNs regulates circadian rhythms in chemosensory sensitivity that underlie the circadian control of these different behaviors.

At night, male nocturnal Manduca sexta hawkmoths are very sensitive to the sex pheromone blend that their conspecific females release in an intermittent pattern (Itagaki and Conner, 1988). The intermittency is a prerequisite to maintain the male’s search flight, which is suspended at pulse frequencies >30 Hz (Baker et al., 1985; Baker and Vogt, 1988; Stengl, 2010). The pheromone pulses are resolved of up to 3 Hz by pheromone-specific ORNs in the antenna’s long trichoid sensilla (Lee and Strausfeld, 1990; Marion-Poll and Tobin, 1992). Pheromone-specificity is provided by odor receptor (OR) subunits in the ciliary membranes of ORNs that heteromerize with the conserved olfactory receptor co-receptor Orco (Nakagawa and Vosshall, 2009; Stengl and Funk, 2013; Wicher and Miazzi, 2021). While the various roles of Orco are still under debate, it is known that in heterologous expression systems Orco homomers form a slow, leaky, non-specific cation channel, which depolarizes the cells (Jones et al., 2011; Sargsyan et al., 2011; Sato et al., 2008; Wicher et al., 2008). In vivo tip recordings of long trichoid sensilla in M. sexta confirmed that Orco acts as spontaneously opening cation channel, acting as a pacemaker channel, because it promotes ultradian membrane potential oscillations that underly the spontaneous action potential activity (Nolte et al., 2013). These potential oscillations turn the ORNs into ultradian temporal filters that are sensitive to the frequency of the pheromone pulses they encounter. Pheromone pulses that arrive at a similar frequency as the endogenous sub-threshold membrane potential oscillations would synchronize these oscillations and would be more likely to elicit action potential trains. Therefore, this pacemaker channel-dependent mechanism would allow for active sensing if endogenous membrane potential oscillations are in the physiological range necessary for tracking intermittent pheromone pulses.

It is not known whether the Orco-dependent membrane potential oscillations also occur at circadian timescales, and whether or how they are linked to ultradian potential oscillations. One hypothesis is that the molecular TTFL clockwork of ORNs exclusively controls endogenous circadian rhythms of the membrane potential via circadian expression of the pacemaker channel Orco. Alternatively, the plasma membrane of ORNs could by itself constitute an autonomous clock that is entrained by but does not rely on the TTFL clock. Constant levels of ion channels that do not require daily degradation and daily transcription via the TTFL clock, but with modulation of open-time probability by posttranslational modifications, would constitute a more economical and faster adjustable mechanism to generate endogenous membrane potential oscillations (Stengl and Schneider, 2024). This would facilitate active sensing, the rapid tuning to detect behaviorally relevant zeitgeber signals in the insect’s environment.

Here, we performed a combination of experimental and computational modeling studies to investigate whether Orco plays a role in active sensing at the circadian and/or ultradian timescale via generation of endogenous membrane potential oscillations. We hypothesize that Orco is part of a post-transcriptional feedback loop (PTFL) clockwork in the plasma membrane, rather than being TTFL controlled. To capture naturally occurring, physiologically relevant processes in insect ORNs we performed in vivo long-term tip recordings of the spontaneous spiking activity of pheromone-sensitive trichoid sensilla of intact but restrained hawkmoth males.

We corroborated the experimental data with a Hodgkin-Huxley (HH) based computational ORN model. In general, the widely used deterministic HH models do not include circadian control of ion channels. Moreover, the highly dynamic, bursting firing patterns of ORNs cannot be described by the traditional deterministic modeling of neuronal activity. This constitutes a drawback because stochasticity plays a crucial role in shaping firing patterns which are influenced not only by external stimuli but also by the inherent noise involved in channel gating. Therefore, to address the above points realistically, we developed a Langevin formulation of the HH model, which explicitly accounts for the noise introduced by the random opening and closing of ion channels and circadian modulation of the Orco channel conductivity. By integrating channel noise, we could explore how circadian regulation of ionic currents contributes to the variability of firing patterns and how this noise interacts with external stimuli to influence olfactory signal processing.

Our model aims to highlight the advantages of using a stochastic framework for modeling the circadian activity of ORNs. The combined experimental and modeling efforts generate a more complete picture and predictions concerning these autonomous processes in sensory neurons, which are required for future exploration of physiologically relevant anticipatory behaviors that are driven by biological clocks.

Methods

Animals

Nocturnal M. sexta hawkmoths were bred and raised from eggs in the rearing facility at the University of Kassel. Males and females were separated during pupal stages. Males were housed isolated from the females to avoid exposure to female pheromones. All experiments were performed with adult males reared in a long-day photoperiod (L:D 17:7 h to prevent diapause) at 25 °C with relative humidity of about 55%. Larvae were fed with an artificial diet modified after Bell and Joachim (1976); adult moths could feed on sugar solution with added vitamins (Roth) ad libitum (Riffell et al., 2008).

Solutions

Ringer compositions were taken from (Pézier et al., 2007) and contained (in mM): 6.4 KCl, 12 MgCl2, 1 CaCl2, 12 NaCl, 10 HEPES, 340 glucose for hemolymph ringer (HLR) at 450 mosmol/kg, and 172 KCl, 3 MgCl2, 1 CaCl2, 25 NaCl, 10 HEPES, 22.5 glucose for sensillum lymph ringer (SLR) at 475 mosmol/kg, both at pH 6.5. The Orco antagonist OLC15 (N-(4-butylphenyl)-2-((4-ethyl-5-(2-pyridinyl)-4H-1,2,4-triazol-3-yl)thio)acetamide (Chen and Luetje, 2012); kindly provided by Dr. D. Wicher, Max Planck Institute for Chemical Ecology, Jena, Germany) was dissolved in DMSO to a concentration of 100 mM (stock solution stored at 4-8 °C) before being diluted in SLR to a final concentration of 50 µM immediately before the start of the recording. DMSO percentage in the final solution was 0.05%.

Electrophysiology

The spontaneous electrical activity of the unstimulated pheromone-sensitive ORNs in a single long trichoid sensillum was recorded extracellularly as in vivo tip recording (Kaissling et al., 1987). 1-2-d-old male M. sexta that were raised in isolation from females and pheromone exposure were fixed in a custom-made Teflon holder with adhesive tape. The right antenna used for recordings was oriented with its dorsal surface facing up and immobilized near the base with dental wax (Boxing wax strips, KerrHawe SA). Electrodes (Ag/AgCl wire in a glass pipette filled with SLR for the recording electrode and HLR for the reference electrode) were pulled (DMZ-Universal Puller, Zeitz Instruments) from borosilicate capillaries (OD:1.56mm, ID: 1.17mm, without filament; Science Products) to a tip diameter of about 2 μm. After cutting off 20 distal annuli of the antenna with micro scissors, the reference electrode was inserted into the antennal lumen up to the second annulus from the cut end. The cut was covered with electrode gel (GE Medical Systems Information Technology) to prevent desiccation. Electrodes were moved with manual micromanipulators (MMJ, Märzhäuser Wetzlar; and Leitz, Leica Microsystems) or with motorized micromanipulators (Microstar, Scientifica) with a joystick controller (Scientifica). Under microscopic control, some tips of the long trichoid sensilla of the distal second annulus were cut off with sharpened Dumont forceps, and the recording electrode slipped over a single sensillum of an upper sensillar row.

ORN activity was amplified 200-fold (custom built amplifier with 1012 Ω input impedance, or ELC-01 MX in combination with DPA-2FX, npi, or EXT 10-2F, npi), low pass filtered at 1.3 kHz, digitized at 20 kHz with a Digidata (1550A or 1550B, Molecular Devices), and saved for offline analysis with Clampex 10 (pCLAMP suite, Molecular Devices).

All tip recording experiments were performed at room temperature (21-27 °C) and a relative humidity of 35-60%. We recorded ORN activity either in long-day conditions (17:7 LD) or constant darkness (DD) to examine the free-running peripheral ORN clock. For LD experiments, the light regime continued as in the rearing facility during the experiment. For DD and OLC15 experiments, the lights were turned off at the beginning of the recording, i.e. all animals were exposed to the same zeitgeber until the beginning of the experiment. To examine the effect of Orco on the spontaneous action potential (AP) activity pattern, we used the recording electrode to infuse OLC15 in DD conditions.

Data analysis

To characterize the spontaneous ORN spiking activity, we first detected spikes with a self-written program in Python. Using the scipy.signal package and find_peaks function, we applied a Bessel filter to the raw voltage trace to facilitate spike detection within the following conditions: spike full width (3-15 ms), threshold search for amplitude of spike (0.5-2.5 mV) and threshold prominence (0.1). Two or more consecutive spikes with intervals ≤ 50 ms were considered a burst (Dolzer et al., 2001). We then calculated the following attributes in bins of either 10 min or 1 h for further analysis: instantaneous spike frequency (1 / inter-spike interval (ISI)), inter-burst interval (IBI, time between two consecutive bursts with no single spikes in between), inter-event interval (IEI, time between two events, i.e. two consecutive single spikes, two consecutive bursts, a burst and a consecutive spike, and vice versa), burst duration (time from first to last spike in a burst), percentage of all spikes in the experiment that are members of burst, and mean number of spikes per burst (Figure 1). All statistics were done with those binned values.

Attributes of pheromone-sensitive ORN spiking patterns.

Spikes were identified by their prominence in the raw ORN recording. The inter-spike interval (ISI) was defined as the time from one spike to the next. A train with ≥2 spikes with ISI ≤ 50 ms was considered a burst. The intra-burst interval was the ISIs within a burst, the inter-burst interval (IBI) the ISIs between two bursts, omitting any individual spikes in between. The inter-event interval (IEI) included the ISIs between bursts and individual spikes.

Hawkmoth behavioral and physiological rhythms desynchronize across the population in the absence of zeitgebers like pheromone stimulation and light-dark cycles. Therefore, it was necessary to phase align the recordings obtained in DD and OLC15. To achieve this, we used an optimization procedure to calculate the optimal temporal lags for each dataset, aiming to maximize the overall cross-correlation between the calculated binned spike frequency series. The optimization was performed using simulated annealing, which provided a set of time shifts (lags) for each series. These lags were applied to shift the frequency data in a manner that maximized the synchronization between the different subjects’ data, resulting in a “subjective ZT” with subjective ZT 0 at the first maximum of the binned ISI for each individual. The objective function used for optimization was the negative sum of pairwise correlations between all series after applying the shifts. The resulting alignment allowed us to compare the frequency patterns across the different conditions.

Once the data were aligned, we computed the mean frequency across all subjects for each bin and each condition. Then, the first high activity period was identified by finding the maximum in the mean frequency series. Subsequently, we identified the second high activity period as the time 24 hours after the first one and the first low activity period was found by locating the minimum in the mean frequency series between the two high activity periods.

Besides the mean frequency, we also searched for significant oscillations in IBI, IEI, burst duration, percentage of spikes that are part of burst, and mean number of spikes per burst. For this we found the maximum value of these statistics within a 3 h range centered in the first low activity period and compared it with the values at the second high activity period.

Continuous wavelet and Fourier analysis were performed using PyBoat (Mönke et al., 2020) to examine the time series of neuronal firing frequency. The continuous wavelet method allows us to detect and characterize rhythmic patterns within the data by identifying frequency components that change over time. Specifically, we used wavelet analysis to determine whether the neuron exhibits circadian and ultradian rhythms. Unlike traditional Fourier analysis, which provides an overall frequency composition but assumes stationarity, wavelet analysis is particularly useful for biological signals because it can reveal how these frequencies evolve dynamically. By applying this approach, we could assess not only the presence of ultradian rhythms but also whether their strength or occurrence fluctuated in a circadian manner, indicating modulation by the 24-hour cycle.

Real-time quantitative polymerase chain reaction (qPCR)

We followed the qPCR methodology detailed in (Schneider et al., 2025). Briefly, male moths were isolated and cultured until the second day after eclosion, and the antennae from four individuals were collected at each zeitgeber time (ZT) 1, ZT9, and ZT17 using liquid nitrogen, with three biological replicates per time point. Total RNA was extracted using TRIzol® Reagent (Invitrogen) and quantified for concentration and purity with the NanoDrop ND-1000 spectrophotometer (Thermo Fisher Scientific). First-strand complementary DNA (cDNA) was synthesized using the PrimeScript™ RT reagent Kit with gDNA Erase (Takara Bio Inc.). Primers (F: AAGCACGTGGTCAGATTGGT, R: GTAAGCCAGGAGCGTCAGAG) for Orco (Gene ID: LOC115452348) were designed using the Primer-BLAST online program of NCBI. The qPCR was conducted following the protocol of TB Green® Premix Ex Taq™ II (Tli RNase H Plus) (Takara Bio Inc.), with the program: 95 °C for 30 s, followed by 40 cycles of 95 °C for 5 s, and 60 °C for 30 s. Relative expression levels of Orco were calculated using the 2-ΔΔCt method (Livak and Schmittgen, 2001).

Statistics

All statistics were done with self-written python codes, SigmaPlot 12 (Systat Software), or JASP (version 0.19.03, University of Amsterdam) with a significance level α = 0.05. Unless noted otherwise, data were checked for normal distribution with Anderson-Darling or Shapiro-Wilk tests and for equal variance with Levene’s test. If one of these assumptions failed, we used non-parametric tests, otherwise we used parametric tests.

To assess the statistical significance of observed differences in neuronal activity of the aligned data between ZTs, we applied the Wilcoxon signed-rank test to pairwise comparisons between the first low activity period and the subsequent high activity period and discarded the transient high activity at the beginning of most recordings.

For other data, we used an ANOVA on ranks with either Tukey’s or Dunn’s post-hoc test for multiple comparisons as indicated.

ORN Model

In this study, we introduce a single-compartment Hodgkin-Huxley (HH) type model of the spontaneous activity of M. sexta ORNs. The model contains 5 sets of ion channels that were previously identified in electrophysiological recordings (Dolzer et al., 2021): the fast sodium (Na+) and potassium (K+) channels responsible for the generation of spikes, low voltage-activated calcium channels (LVA) responsible for the initiation of bursts, calcium-gated potassium channels (BK) responsible for the termination of bursts, and the Orco channel, which is a voltage-independent non-specific, cAMP-gated channel. A schematic of the model is provided in Figure 2A. We modeled Orco as a ZT-dependent conductance to represent the effect of the oscillations in cAMP concentration that were identified in these cells (Schendzielorz et al., 2014) (Figure 2B, C). For simplicity we chose to use a sine function to directly represent the oscillations in Orco conductivity, but similar results can be achieved by modeling it as an output of a Goodwin model of the circadian clock via cAMP concentration.

The conductance-based model of a simplified hawkmoth ORN.

(A) Schematic of the model containing a voltage gated sodium channel (Na+), a voltage gated potassium channel (K+), a low voltage activated calcium channel (LVA), a calcium and voltage activated potassium channel (BK), a leak channel, and a cAMP-gated leaky non-specific cation channel (Orco). The Orco channel is represented as a linear conductance whose value oscillates with a sinusoidal shape as a function of the ZT. The circadian rhythm in the Orco conductance is hypothesized to result from a cAMP-dependent increase in channel open-time probability (as found in Drosophila Orco: (Getahun et al., 2013)) via a circadian oscillation of antennal cAMP concentrations (Schendzielorz et al., 2015). This oscillation is driven by the circadian clock and serves as an input; the internal mechanism of the clock is not modeled explicitly. Arrows indicate the flow of ions through the channels. (B) Parametrized circadian oscillation in Orco conductivity, introduced as an external input to our model, representing the gating of Orco by cAMP whose concentration oscillates on a circadian timescale. The times of minimum and maximum cAMP concentration are indicated by colored arrows. (C) Parametrized IV curve of the Orco channel showing a linear behavior with the slope depending on the cAMP concentration. Line colors correspond to the arrows in panel B. (D) Markov-chain representing all possible states and transitions for each type of ion channel included in the model. The per capita transition rates (α and β) depend on membrane potential and in the case of the BK channel also on the intracellular Ca2+-concentration. The “N” state vectors contain the population of channels in each state “n”. Subscripts of “n” indicate how many activation and inactivation gates are in the open state. Directed edges are numbered in red. The only conducting state of each ion channel, representing the condition where all its gates are open, is shown in green. Panel A was drawn using pictures from Servier Medical Art. Servier Medical Art by Servier is licensed under a Creative Commons Attribution 3.0 Unported License.

Each type of ion channel possesses one or more gates that can open and close individually with transition rates α and β for opening and closing, respectively, of activation gates (n, m) and inactivation gates (h). As a result, each possible combination of open and closed gates defines a possible state in which each ion channel can be found. A schematic of the Markov chain depicting all possible states and transitions for each ion channel is shown in Figure 2D. Each ion channel can conduct current only in the state where all its gates are open.

First, we constructed a deterministic 22-dimensional model to serve as the mean field of the Langevin model following the methodology proposed by Pu and Thomas (2020). Here, one dimension tracks the membrane potential, another the intracellular Ca2+ concentration, and the remaining 20 dimensions track the number of channels that are in a specific state of the Markov chains. Considering the conservation of the total number of ion channels, each type of ion channel required one less dimension than the total possible states.

The complete Langevin model adds to the mean field an independent noise source for each transition in the Markov chain. These noise sources are biophysically motivated since the transition between ion channel states is a stochastic process, with the probability of transition determined by the kinetics of each kind of ion channel. The variability in the spiking behavior of a neuron emerges from this non-deterministic gating of individual ion channels. The detailed mathematical description of the model can be found in the Supplementary Material.

Results

We tested the hypothesis that sensory neurons, such as insect ORNs, perform active, anticipatory sensing based upon an endogenous, plastic plasma membrane clock with the pacemaker channel Orco as the core element. We focused on pheromone-sensitive long trichoid sensilla of male M. sexta hawkmoth antennae which exhibit daily rhythms in pheromone sensitivity and temporal resolution that are controlled by a circadian clock. We performed minimally invasive in vivo tip-recordings over the course of several days to search for daily rhythms in spontaneous spiking activity as a measure of endogenous membrane potential oscillations.

Pheromone-sensitive hawkmoth ORNs exhibit Orco-dependent circadian rhythms in spontaneous spiking activity

The ORNs that innervate the pheromone-sensitive long trichoid sensilla on male M. sexta antennae display spontaneous spiking patterns of irregular bursts with interspersed single action potentials in short-term recordings (Dolzer et al., 2001). In the present study we documented the same pattern in long-term recordings over the course of several days. This enabled us to search for daily and circadian rhythmicity in spontaneous ORN spiking activity in the absence of pheromone stimulation.

The spontaneous spiking patterns revealed clear daily rhythms in 5 of 11 animals under long-day conditions (17:7 h LD) (Figure 3A). The maxima in spontaneous activity occurred mostly during the dark phase when nocturnal hawkmoths are active (Figure 3A (top), Bi light blue arrow), and minima occurred during the light phase when hawkmoths rest/sleep (Figure 3A (bottom), Bi dark blue arrow). High spiking activity in the dark phase had a mean maximum frequency of ∼3 Hz (theta range: 0.5 – 4 Hz) (Figure 3Bii). In contrast, the minimum spiking frequencies during the hawkmoth’s resting phase were below 0.5 Hz (Figure 3A (bottom)). Since ORN firing patterns had a clear daily rhythm in LD both in individual animals (Figure 3Bi) and averaged across animals (Figure 3Bii, Biii), we examined whether these rhythms depended on cycling light cues or whether they persisted in constant darkness (DD), as expected for an endogenous, circadian clock-driven rhythm.

Spontaneous activity of pheromone-sensitive olfactory receptor neurons (ORNs) of male M. sexta shows Orco-dependent circadian modulation.

(A) Example tip recordings of the spontaneous activity of one long trichoid sensillum during a period of high (top, average frequency ∼ 3 Hz) and low (bottom, average frequency ∼ 0.2 Hz) activity. (Bi) Spontaneous spike frequency of one long-term ORN recording during 17:7 light-dark cycles (same experiment as in A, colored arrows indicate respective zeitgeber times (ZT)), indicated by the black-white bar at the top, increased during each activity phase and was low during the resting phases. Spike frequencies were averaged for each 1 h bin. (Bii) Mean spontaneous ORN spike frequency across all LD animals in 1 h bins. Data from the individual in Bi is highlighted in red. Recordings had different lengths, therefore the number of recordings used for the element-wise mean for each 1 h bin decreased with the time since the start of the recording, indicated in the top panel. (Biii) Mean spontaneous ORN spike frequency across all LD animals in 1 h bins during the first 48 hours revealed circadian activity. (Ci) The spontaneous spike frequency of one long-term ORN recording in constant darkness (DD, the expected times of lights on are represented by grey bars above) exhibited a circadian pattern. The peak activity shifted in DD due to the animal’s endogenous, free-running circadian period τ = 21.27 h. (Cii) Mean spontaneous ORN spike frequency across all DD animals in 1 h bins. Data from the individual in Ci is highlighted in red. (Ciii) Mean spontaneous ORN spike frequency across all DD animals in 1 h bins during the first 48 hours after aligning the time to the first maximum in spontaneous activity (see Methods) revealed circadian activity. (Di) The spontaneous spike frequency of one long-term ORN recording in constant darkness with infusion of the Orco antagonist OLC15 (orange frame) dissipated the circadian rhythm of spontaneous activity. (Dii) Mean spontaneous ORN spike frequency across all OLC15 animals in 1 h bins. Data from the individual in Di is highlighted in red. In contrast to LD and DD conditions, the spike frequency decreases over time with prolonged exposure to OLC15. (Diii) Mean spontaneous ORN spike frequency across all OLC15 animals in 1 h bins (mean ± SD) during the first 48 hours after aligning the time to the first maximum in spontaneous activity (see Methods). In contrast to LD and DD conditions, the circadian change in spontaneous spiking frequency disappeared. (E) The slopes of the linear fits to the binned spiking activity of each individual animal in the three different conditions (see Methods). Each dot indicates the slope for one experiment. The line indicates the mean. The slope for OLC15 is significantly different from control LD.

In DD, the spontaneous ORN firing pattern remained rhythmic (7 of 10 animals), albeit with less robustness: The single peak of high spiking activity that was observed at the same ZT across several days in LD divided into multiple bouts of high spiking activity in the activity phase in 7 of 10 animals (Figure 3Ci). Also, in DD conditions, the high spiking activity phase shifted into the subjective day due to the individual endogenous circadian periods of about 23.5 h +/-2.8h (21.27 h in Figure 3Ci). Thus, averaging the instantaneous spiking activity across all DD animals did not depict clear circadian rhythms (Figure 3Cii). Therefore, we phase-aligned the mean frequency of spontaneous activity (see Methods). After alignment, circadian rhythms were exposed across DD animals (Figure 3Ciii), comparable to the daily rhythms in LD (cf. Figure 3 Biii and Ciii).

Next, we examined whether Orco, a leaky, non-specific cation channel, is the dominant depolarizing pacemaker current of ORNs that drives circadian rhythms in spontaneous spiking activity. Infusion of the Orco antagonist OLC15 into the sensillum lymph obliterated circadian rhythms and attenuated the spontaneous activity in several, but not all experiments (N = 8 of 12) (Figure 3Di). This attenuation resulted in a linear decrease in spiking activity over several days (Figure 3Dii, Diii). Even after phase-aligning the individual experiments, no circadian rhythm was present in the average across all OLC15-treated animals (Figure 3Diii).

Because OLC15 is not membrane-permeable on its own it was infused with low concentrations (0.05%) of DMSO through the recording electrode into the sensillum lymph and, therefore, effectiveness increased over time. Hence, we fit the spiking activity of each animal and in each condition with linear regression lines over the whole recording time to compare the slopes of the fits (Figure 3E). In LD, the mean ± SD slope was 0.004 ± 0.006 Hz/h, indicating that the average spiking activity remained similar throughout the duration of the recording. In DD, the slope was −0.003 ± 0.018 Hz/h and was not different from LD. In contrast, the mean slope during OLC15 treatment was −0.007 ± 0.022 Hz/h, significantly smaller than in LD (ANOVA on ranks with Dunn’s post-hoc test; H(2) = 7.681, p = 0.021). The negative slope corroborated the finding that spiking activity decreased over time in the presence of the Orco blocker.

Activity rhythms between hawkmoths desynchronize and have only weak phase coupling in the absence of pheromone stimulation, even in LD cycles (Ghosh et al., 2024). Therefore, instead of comparing fixed ZT or circadian time (CT) intervals, we compared different attributes of spontaneous activity between periods of high and low activity, for DD conditions after phase alignment (see Methods; Figure 4). Despite the high variability in ORN spiking between individuals, the mean frequency of spontaneous ORN activity was significantly higher during the high activity phase than during low activity in both LD and DD recordings (Figure 4Ai, Aii). In LD but not in DD, the frequency of bursting increased significantly during the high activity phase (Figure 4Bi, Bii). The percentage of spikes that are part of bursts was less variable and lower during the high activity phase in LD but not significantly different in DD (Figure 4Ci, Cii). The mean burst duration was less variable during the high than the low activity phase in LD, but not significantly shorter in any condition (Figure 4Di, Dii). Furthermore, both in LD and DD, the mean inter-event frequencies were significantly higher during the high activity phase (Figure 4Ei,Eii), indicating a significantly higher overall spontaneous spiking activity during the hawkmoth’s activity phase. The mean number of spikes per burst, and intra-burst frequencies did not differ significantly between high or low activity periods in both LD and DD conditions (Figure 4Fi, Fii, Gi, Gii). The addition of Orco antagonist OLC15 deleted any significant differences of spontaneous activity found in LD and DD (Figure 4Aiii-Giii). In conclusion, the mean frequency of spontaneous spiking and the frequency of bursting expressed daily modulation, and are both most likely controlled via a circadian clock that targets the leak channel Orco.

Blocking Orco removed circadian modulation of spontaneous ORN spiking patterns.

Some attributes of the spontaneous ORN activity were significantly different (Wilcoxon signed-rank test, α = 0.05) between the time windows of low vs. high activity, mostly in LD (white boxes) and DD (gray boxes) conditions, but never when Orco was blocked with OLC15 (orange boxes). In both LD and DD, the mean spiking frequency was increased during high activity (Ai, Aii). Only in LD but not in DD, the mean inter-burst intervals (IBI; Bi, Bii), the relative number of spikes belonging to a burst (Ci, Cii) and the mean inter-event interval (IEI; Di, Dii) decreased significantly during the high activity period. Both in LD and DD the mean burst duration (Ei,Eii), mean number of spikes per bust (Fi,Fii), and mean intra-burst intervals (Gi, Gii) did not differ between low vs. high spiking activity. Also, with Orco blocking, the attributes for low vs. high activity did not differ for any of the attributes tested (Aiii-Giii). In DD and OLC15conditions, activity phases were aligned as described in the Methods; thus, low activity in DD was at subjective ZT 19 and in OLC15 at subjective ZT 13, high activity was in both cases at subjective ZT 24. ZTs for low and high activity in LD were 10 and 0, respectively.

Orco imposes circadian modulation on the ultradian rhythms in spontaneous ORN activity

After establishing that Orco influences the circadian changes in ORN spiking activity between the animal’s resting and activity phase we examined if Orco also influences ORN spiking rhythms on the largely different ultradian timescales in the frequency range between 0.01-1000 Hz. In all conditions, two broad bands of ultradian instantaneous frequencies were evident: one upper band of >10 to ∼100 Hz and a lower band between 0.1 to ≤10 Hz (Figure 5). The upper band primarily represents the frequencies of spikes within a burst, while the lower band represents single spikes and frequencies between bursts.

Blocking Orco removed circadian regulation of ultradian rhythms in the spontaneous ORN spiking pattern.

Heat map of the instantaneous frequencies (1/ISI) over two consecutive days of long-term tip-recordings of one pheromone-sensitive long trichoid sensillum under different conditions (LD (A), DD (B), OLC15 (C)) in each panel. Pixel color indicates the counts of instantaneous frequencies in that respective bin. The points fall mostly in a band of high frequencies (>10 – ∼100 Hz) and low frequencies (0.01 – ∼10 Hz), with the high-frequency band representing the instantaneous frequencies of spikes within a burst and the low-frequency band the instantaneous frequency between bursts. (A, B) The high frequency band indicates daily and circadian modulation of frequency prevalence. In addition, the low frequency band (<0.01 Hz to ∼ 10 Hz) also displayed circadian modulation of the frequency composition. (C) Infusion of OLC15 deleted the circadian modulation in both frequency bands and enhanced the ultradian rhythm of frequency prevalence.

In LD, two types of daily/circadian modulation of the ultradian frequencies occurred (Figure 5A). The upper higher frequency band was mostly modulated in prevalence (its frequency of occurrence) but not in its frequency range. The lower band was modulated both in prevalence and frequency range, which resulted in a sinusoidal change in the mean of the frequency band over 24 h. The low and high frequency bands appeared to merge during the activity phase around ZT 0 in the animals that showed clear circadian rhythms (N = 5 of 11 in LD), when both bands had maximum prevalence. During the hawkmoth’s activity phase, higher frequencies of up to ∼10 Hz occurred, while frequencies dropped below 1 Hz during the resting phase.

The circadian modulation of both frequency bands was maintained in constant darkness (Figure 5B) but phase-shifted with respect to the projected ZT due to the endogenous circadian period of each individual. Again, this indicates that a clock is modulating the spontaneous ORN spiking activity over the course of a day.

The addition of the Orco antagonist OLC15 deleted any circadian modulation of either frequency band. Interestingly, the instantaneous frequencies in both bands still showed correlated changes in prevalence and merging, but now at ultradian periods only.

Additional Fourier analysis (see Methods) of spontaneous spiking activity over the course of several days revealed faster ultradian and slower infradian rhythms, ranging from 0.6 h to 38 h, in addition to the circadian (24 ± 4 h) rhythm (Figure 6Ai, Bi, Ci). Some rhythms (factorials of 24 h, like 12 h, 6 h, etc.) appeared to be harmonics of the circadian rhythm, which were not excluded in our analysis. Most animals expressed ultradian rhythms of 2-4 h in LD as well as DD conditions. With OLC15 the occurrence of circadian rhythmicity decreased from 70% to 33.3 %, while the prevalence of infradian periods increased from 0% to 33.3 % and ultradian periods remained the same. These results were corroborated by continuous wavelet analysis, which provided time-resolved spectral power across different rhythmic periods (Figure 6Aii, Bii, Cii).In control LD and DD conditions, wavelet power for ultradian periods increased during times of elevated firing rate and decreased during periods of low activity (Figure 6Aii and Bii, middle and bottom panels). This dynamic pattern indicates that ultradian rhythms were not constant over time but were instead modulated by the circadian rhythm, becoming more prominent during circadian peaks in neuronal activity. Such circadian gating of ultradian power was clearly visible in the wavelet spectrograms as alternating bands of higher and lower ultradian power aligned with daily activity cycles. Following Orco blockade with OLC15, this modulation was disrupted (Figure 6Cii middle panel): although ultradian components persisted (Figure 6Cii bottom panel), they no longer showed systematic variation with the circadian cycle, suggesting a decoupling of ultradian rhythms from circadian control.

Blocking Orco affected ultradian and infradian frequencies in the spontaneous ORN firing patterns.

The Fourier analysis of spontaneous spiking activity revealed rhythms with ultradian (<20 h; dark grey), circadian (20-28 h; white), and infradian periods (>28 h; light grey) in LD (Ai), DD (Bi), and OLC15 (Ci). Each line represents one animal where each triangular marker is at the local maximum in the frequency spectra obtained for that animal. The histograms above illustrate how often specific periods of spiking activity rhythms occurred averaged over all animals. Circadian rhythms were detected in LD (N = 5 of 11), DD (N = 7 of 10), and OLC15 (N = 4 of 12). Wavelet analysis of LD (Aii), DD (Bii), and OLC15 (Cii) confirmed the occurrence of multiscale periods. Example plots for one animal each, the same animals as in Figure 5. For each panel, the top plot depicts the mean firing frequency, the middle plot the wavelet power in the same period range as panels i, and the bottom plot the wavelet power for infradian periods up to 7 h on a log scale to highlight infradian time scales.

Orco expression is not under the control of the molecular TTFL clockwork

So far, our results indicated a prominent role of Orco in the circadian modulation of ORN spiking activity. Since it is the general opinion that the master circadian clock that dominates all physiological and behavioral circadian rhythms functions via transcriptional control we examined whether Orco transcription shows a circadian rhythm, as previously found for clock genes in hawkmoth antennae (Schneider et al., 2025; Schuckel et al., 2007). However, qPCR of whole male hawkmoth antennae revealed no significant circadian rhythm of Orco expression in DD (Figure 7; ANOVA on ranks, H(2) = 5.60, p = 0.061). Thus, it is possible that Orco is controlled via a posttranslational circadian clockwork mechanism that is associated directly with the plasma membrane (Stengl and Schneider, 2024; Stengl and Schröder, 2021).

Orco is not under control of the TTFL-based molecular circadian clock.

Relative expression levels of Orco mRNA from male hawkmoth antennae raised under LD conditions did not differ significantly (ANOVA on ranks, H(2) = 5.60, p = 0.061) between ZT 1 (end of activity phase), ZT 9 (resting phase), and ZT 17 (beginning of the activity phase). timeless mRNA levels from male hawkmoth antenna served as control and were published in Schneider et al. (2025).

Circadian changes in Orco conductance are sufficient to modulate spiking activity in a computational model of a hawkmoth ORN

Although Orco is a leaky ion channel, its maximum conductivity is modulated by cAMP (Getahun et al., 2013; Martín et al., 2001; Sargsyan et al., 2011; Wicher et al., 2008). In hawkmoth antennae, cAMP levels oscillate throughout the day (Dolzer et al., 2021; Flecke et al., 2010; Flecke and Stengl, 2009; Schendzielorz et al., 2014, 2015), which makes cAMP modulation of Orco a putative mechanism for the Orco-dependent circadian modulation of ORN spiking activity, integrating Orco in a circadian PTFL based membrane clockwork (Stengl and Schneider, 2024).

To test the hypothesis that Orco is a cAMP-gated, leaky ion channel, we constructed an ORN model (Methods and Supplementary Material) and tuned the parameters to provide a quantitative match with the biological data. The developed model provides a unified description of the experimental observations. For instance, the ORN model yields irregular spiking activity with a bimodal distribution, consisting of both bursts and isolated spikes, successfully reproducing the experimental recordings (Figure 8A). This suggests a unified mechanism underlying the initiation of both phenomena: If the pacemaker-driven depolarization reaches the Na+ threshold a firing process begins, and at this point randomness determines if other slower ion channels responsible for bursting are sufficiently activated to sustain a burst of several spikes.

The model adequately reproduces the firing pattern of the biological ORN.

(A) The model activity (bottom trace) displays both isolated spikes and bursts of variable length as in the original recording (top trace). The model represents an intracellular recording whereas the biological recordings were done with extracellular electrodes. (B) Semi-logarithmic plot of burst length distribution of an experimental LD recording (top) and simulated model (bottom) of spontaneous action potential activity of pheromone-sensitive neurons. Both the model and the recordings display exponential decrease of the count’s density as a function of the burst length, shown as a linear decrease in the logarithmic scale.

Additionally, the ORN model replicates the distribution of the number of spikes per burst, which follows a decreasing exponential trend in both experimental and simulated results (Figure 8B).The model also successfully reproduces the circadian modulation of spiking activity: the maximum firing rate occurs when the Orco conductance is maximal, whereas the maximum percentage of spikes that are part of bursts correlates with the lowest Orco conductance, resulting in antiphase oscillations, as were observed in the experimental recordings (Figure 9).

Comparison of the circadian modulation of model and ORN spiking attributes.

Mean spike frequency and percentage of spikes that belong to bursts show anti-phasic circadian oscillations both in the simulated model (left) and in the biological recording (right).

The frequency prevalences in the model over several days clustered in two distinct frequency bands: a top band that corresponds to the intra-burst frequencies and a bottom band, corresponding to inter-burst frequencies (Figure 10), in agreement with the biological experiments (Figure 5A, B). When these circadian oscillations in Orco conductance were included, inter-burst frequencies showed significant circadian variations, whereas intra-burst frequencies remained relatively constant, reminiscent of the biological data. Notably, the simulation captured the experimentally observed “merging” effect, where the low frequency band approaches the top band during high-activity periods (Figure 10).

The model captures the Orco-dependent circadian modulation of ultradian spiking rhythms.

Heat map of the simulated model, pixel color indicates the density of instantaneous frequencies in that respective bin. The points can be roughly divided in a band of high frequencies (top, ∼50 –80 Hz) and low frequencies (bottom, ∼0.1 – 5 Hz), with the high-frequency band representing the instantaneous frequencies of spikes within a burst and the low-frequency band the instantaneous frequency between bursts. (A) Simulated spikes with circadian regulation by the Orco channel; the inter-burst frequencies varied widely with circadian rhythmicity while the frequencies within burst remained approximately constant. The simulated results also reproduced the “merging” effect between the two bands when the bottom one approaches the top one. (B) Simulated spikes without circadian regulation of Orco had no circadian oscillations.

This merging of spiking and bursting frequencies suggests that the system approached a bifurcation, transitioning from a bursting regime to tonic spiking activity. Further evidence for the near bifurcation condition of the neurons is the distribution of number of spikes per burst that followed a decreasing exponential trend in contrast to a constant value, or a bounded range of spikes per burst, which would be expected from a standard bursting neuron model (Figure 8B). The observation that the circadian modulation of spiking activity requires only one parameter, namely changes in the Orco conductivity, to describe the circadian effects on the ORN spiking pattern indicates the crucial role of Orco channel in circadian control.

Discussion

We searched for a membrane-bound clockwork component in insect ORNs that allows for sensing of regular pheromone patterns to guide mating flights. For this, we recorded interlinked circadian and ultradian frequencies in the spontaneous spiking activity of the pheromone-sensitive long trichoid sensilla of male Manduca sexta hawkmoths with in vivo long-term tip recordings. We found that Orco, which serves several important functions in insect olfaction (Stengl and Funk, 2013; Wicher and Miazzi, 2021), acts as a pacemaker channel that controls the circadian modulation of the spiking activity without affecting the frequency ranges that are relevant for the primary fast, phasic pheromone response (Dolzer et al., 2003; Nolte et al., 2016; Schneider et al., 2025). While circadian rhythmicity disappeared in the presence of the Orco blocker OLC15, ultradian activity rhythms remained. We confirmed with computational modeling that the circadian modulation of open-time probability of Orco as a leaky pacemaker channel was sufficient to account for our experimental results. Since we observed no evidence for circadian transcriptional control of Orco, our findings are consistent with the hypothesis that the plasma membrane constitutes an endogenous post-translational feedback loop (PTFL) clockwork. In this clockwork, circadian changes in cyclic nucleotide levels (Schendzielorz et al., 2015) modulate Orco conductance rather than transcription and insertion of additional Orco channels in the membrane.

The expression of a pacemaker channel that does not require daily degradation and transcription would constitute an economic and rapidly adjustable mechanism to generate endogenous membrane potential oscillations that are tuned to detect behaviorally relevant zeitgeber signals in the insect’s environment (Stengl and Schneider, 2024; Stengl and Schröder, 2021). The present study provides the first evidence for a systems view of timing that involves a PTFL membrane clock in hawkmoth ORNs, instead of the prevailing hierarchical view that assumes all endogenous circadian oscillations are outputs of the genetic TTFL clock (Stengl and Schneider, 2024; Stengl and Schröder, 2021).

Orco acts as a pacemaker channel to drive the circadian rhythm in spontaneous ORN spiking activity

A common motif of spontaneously active neurons is the expression of pacemaker channels that open at hyperpolarized potentials and thus depolarize the neuron (Bose et al., 2014; Cochet-Bissuel et al., 2014; Das et al., 2016; Golowasch et al., 2017; Lüthi and McCormick, 1998; Ratliff et al., 2021; Robinson and Siegelbaum, 2003; Sharma et al., 2023). The resting membrane potential of most neurons lies close to the negative equilibrium potential for potassium ions due to potassium leak channels (Goldstein et al., 2001; Patel and Honoré, 2001; Talley et al., 2001). When the pacemaker-driven depolarization reaches spike thresholds, voltage-gated Na+, K+ and Ca2+ channels promote spiking. Voltage- and Ca2+-gated K+ channels then repolarize the neuron to negative membrane potentials, re-starting the cycle

The core element of the different frequencies of endogenous oscillations that needs to be identified is the respective pacemaker channel (Stengl and Schneider, 2024). The finding of distinct rhythms in spontaneous spiking activity under constant conditions are indicative of underlying endogenous membrane potential oscillations. Therefore, different sets of voltage- and Ca2+-dependent ion channels with antagonistic effects on the membrane potential appear to underlie different ultradian oscillation frequencies of spontaneous activity that still need to be further elucidated (Dolzer et al., 2021; Stengl, 1994, 1993; Zufall et al., 1991). While this study did not aim to unravel the overall generation of spontaneous activity in the ORNs, we could demonstrate that Orco acts as a circadian pacemaker channel that drives the circadian modulation of the spiking activity.

The constitutively expressed Orco is a core element of a circadian PTFL clockwork in hawkmoth ORNs

In hawkmoth ORNs, at least two types of pacemaker channels exist: the highly conserved, slow, leaky Orco, which we focus on in this study, and an HCN type channel (Butterwick et al., 2018; Dolzer et al., 2021; Nolte et al., 2016, 2013; Sato et al., 2008; Stengl and Funk, 2013; Stengl and Schneider, 2024; Stengl and Schröder, 2021; Wicher et al., 2008; Wicher and Miazzi, 2021).

When expressed in vitro in heterologous expression systems, Drosophila Orco homomers assemble into leaky, non-specific cation channels with a reversal potential around 0 mV, constitutively leaking Na+, K+, and Ca2+ (Sato et al., 2008; Wicher et al., 2008; Wicher and Miazzi, 2021). In Drosophila, Orco is regulated by Ca2+/calmodulin and phosphorylation/de-phosphorylation at multiple sites (Cao et al., 2016; Getahun et al., 2016; Guo and Smith, 2017; Mukunda et al., 2016, 2014; Sargsyan et al., 2011). In Drosophila, the Ca2+/calmodulin binding site controls the localization of the OR-Orco complex in the ciliary membrane of ORNs and depends on previous odor stimulation (Cao et al., 2016; Guo and Smith, 2022; Jain et al., 2021). After phosphorylation of Orco by protein kinase C (PKC), a cyclic nucleotide binding site becomes available, and subsequent binding of cAMP or cGMP increases the open-time probability (Getahun et al., 2013; Wicher et al., 2008). This indicates two complementary mechanisms by which Orco conductance can be regulated: by controlling the number of Orco channels in the membrane or by modulating the open-time probability of existing channels through posttranslational modifications.

Since blocking Orco in M. sexta decreased the spontaneous activity and deleted its circadian modulation, Orco is apparently the main target for modulation of circadian activity in hawkmoth ORNs. Since Orco transcription levels were constant, posttranslational modifications of Orco that modulate the open-time probability provide a mechanistic explanation. Increases in cAMP and cGMP levels sensitize or adapt the ORNs to pheromone, respectively (Dolzer et al., 2021, 2008; Flecke et al., 2010, 2006; Redkozubov, 2000; Schendzielorz et al., 2015, 2012; Ziegelberger et al., 1990). While cAMP levels are under circadian control and peak during the activity phase, cGMP levels oscillate in a daily rhythm and peak during the resting phase in insects (Schendzielorz et al., 2014, 2015). We propose that cAMP and cGMP binding modulate Orco with opposing effects, consistent with the physiological needs of the hawkmoth during its sleep-wake cycle: cAMP binding increases Orco conductance while cGMP binding decreases it.

Our hypothesis that a circadian PTFL membrane clock in hawkmoth ORNs controls circadian activity rhythms predicts that the pacemaker channel Orco is part of a membrane signalosome. The protein complexes of this signalosome would comprise the pacemaker channel that generates the specific circadian frequency of endogenous potential oscillations, linked to oscillations of second messenger cascades that determine physiological setpoints of the circadian sleep-wake cycles. Thus, the circadian cAMP oscillations could result from a PTFL membrane clock comprising of a signalosome that minimally consists of voltage-gated Ca2+-channels, Orco, and membrane-anchored adenylyl cyclases (ACs). Since the membrane potential directly influences intracellular Ca2+ levels, the AC activity could be periodically upregulated by the circadian oscillations of the endogenous membrane potential oscillations. Alternatively, AC expression could be under the control of the molecular TTFL clock, which would put Orco under indirect TTFL control. This possibility should be further explored in hawkmoth antennae. Instead of, or in addition to, the voltage-gated Ca2+ channels the signalosome could also work if it contained octopamine receptors that cause circadian rises in cAMP levels through activation of Gαs, as found in hawkmoths (Dacks et al., 2006; Flecke et al., 2010). Since octopamine levels in the hemolymph exhibit circadian oscillations and octopamine sets a physiological state of heightened alertness in insects, the circadian PTFL clock in ORNs would synchronize with endogenous physiological rhythms via octopamine receptor activation (Flecke and Stengl, 2009; Schendzielorz et al., 2015).

Circadian and ultradian rhythms in spontaneous ORN activity resemble the physiologically relevant timescales for active sensing

Moths are more sensitive to pheromone stimulation during their activity phase than at rest, apparently due to antennal circadian clocks (Flecke et al., 2010; Flecke and Stengl, 2009; Merlin et al., 2007, 2006; Schendzielorz et al., 2015; Schuckel et al., 2007; Silvegren et al., 2005). The greater the distance from the female, the lower the pheromone concentration. The increased pheromone sensitivity during the activity phase enhances the male’s chance to detect a calling female.

Pheromone concentration is encoded in the frequency of the first, phasic burst of spikes of the characteristic triphasic spiking sequence of ORNs in response to brief pheromone pulses, which is independent of Orco (Dolzer et al., 2003; Nolte et al., 2013; Schneider et al., 2025).

However, through the cAMP-dependent increase in Orco conductance during the activity phase, the ORN membrane potential is closer to spike threshold and fewer pheromone molecules are necessary to produce a receptor current that brings the membrane voltage over that threshold. Thus, larger Orco conductance increases the likelihood of an ORN response to pheromone stimulation without changing the typical triphasic spiking pattern of the response.

In the odor transduction of Drosophila, OR-Orco activation drives the initial receptor current response. In contrast, Orco is not part of an OR-Orco odor-gated receptor ion channel complex in hawkmoths but controls the spontaneous ORN activity and the late, long-lasting pheromone response (LLPR). The LLPR begins several hundred milliseconds after the initiation of the pheromone-dependent transduction current and correlates with the time course of pheromone-induced rises in cGMP levels (Boekhoff et al., 1993; Nolte et al., 2016, 2013; Stengl, 2010; Stengl et al., 2001; Wicher and Miazzi, 2021; Ziegelberger et al., 1990). Thus, in hawkmoths, Orco is not part of the primary transduction channel but rather controls membrane potential oscillations and thereby the temporal resolution and general pheromone sensitivity of ORNs. The higher spontaneous burst frequency during the hawkmoth’s activity phase improves the resolution of the anticipated pheromone pulses during its mating flight in search for females.

In the mammalian suprachiasmatic nucleus (SCN), the clock that controls the sleep-wake cycle, cAMP levels oscillate at an ultradian frequency, and cAMP and cGMP induce opposing phase shifts in SCN spiking activity (Prosser et al., 1989; Prosser and Gillette, 1991). In M. sexta antennae, cAMP levels express daily oscillations and peak during the activity phase. In contrast, cGMP levels do not oscillate but are elevated by pheromone exposure, which adapts the spiking activity (Boekhoff et al., 1993; Dolzer et al., 2021; Flecke et al., 2006; Schendzielorz et al., 2015; Ziegelberger et al., 1990). This implies that cAMP and cGMP may have opposing effects on Orco conductivity, with results similar to what is seen in the SCN.

The upregulation of neural activity during wakefulness/alertness seems to be ubiquitous among animals. For example, the many synchronously recorded neurons in human EEG show high frequency oscillations in the beta and gamma ranges during alertness and low frequency oscillations in the delta and theta ranges during rest and sleep. Similarly, the spontaneous spiking activity of hawkmoth ORNs is upregulated during their activity phase, albeit only from below delta to the delta/theta range. The increased spontaneous spiking activity in the theta range during the activity phase of the hawkmoth matches the maximal temporal resolution of pheromone pulses of up to 10 Hz (Marion-Poll and Tobin, 1992), which supports a role for Orco as part of a membrane-bound PTFL clockwork that tunes the ORNs for active sensing. Even through oscillations in the field potential of electroantennogram responses to odor stimulation can reach 100 Hz in different insects (Szyszka et al., 2014), these frequencies most likely do not encode physiological odor responses. Apparently, they reflect the astounding adaptive ability of antennal ORNs to synchronize their endogenous membrane potential oscillations across the antenna to a broad range of stimulation frequencies. Pheromone pulse resolution of about 100 Hz was never reported in single olfactory sensilla recordings, and in behavioral experiments male moths stopped their upwind mate search if pheromone pulses exceeded 30 Hz (Baker et al., 1985; Bau et al., 2005, 2002; Tripathy et al., 2010).

Circadian oscillations in the ORN membrane potential would be ideally suited to tune sensitivity and temporal resolution of the pheromone-sensitive ORNs to the hawkmoth’s rest-activity cycle, synchronizing male and female mating behavior. Further experiments are needed to test our hypothesis that ORN signalosomes are highly plastic endogenous PTFL clocks comprising receptors for circadian and ultradian Zeitgebers that allow to tune into internal physiological and external environmental rhythms as basis for active sensing.

Stochastic ion channel dynamics and circadian modulation of Orco conductance explain modeled ORN firing patterns

Our modeling results provide mechanistic insight into the irregular spontaneous activity of pheromone-sensitive ORNs in M. sexta, highlighting how stochastic ion channel dynamics and circadian modulation of ion conductance together shape spiking patterns. By employing a Langevin approximation of the stochastic HH formalism, we demonstrate that the intrinsic noise arising from the random gating of ion channels is sufficient to generate both isolated spikes and bursts of spikes, as observed experimentally. Our model effectively recapitulates the spike distributions found in vivo and accounts for the role of channel noise as a critical determinant of ORN firing patterns.

A key contribution of our model is the incorporation of the Orco ion channel as a pacemaker component, with a conductance that varies linearly with cAMP concentration. We show that circadian modulation of Orco conductance alone is sufficient to produce the observed time-of-day dependent changes in spike distribution. This finding suggests that the rhythmic expression of cAMP levels across the circadian cycle may act as a primary driver of diurnal changes in ORN excitability, without requiring additional external inputs or modulation of other ion channels.

Importantly, our results reveal that the same set of ion channels can support the emergence of both single action potentials and complex bursting behaviors purely through the stochastic variability of their gating kinetics. This unifying explanation reduces the need for invoking separate biophysical mechanisms or circuit-level inputs to explain these distinct firing modes.

For simplicity, the circadian variation in cAMP concentration was modeled using a sinusoidal function. While this approximation is sufficient for capturing the qualitative features of the rhythmic modulation, future work could enhance the biological realism of the model by coupling Orco conductance dynamics to a molecular circadian oscillator, such as a Goodwin-type transcriptional-translational feedback loop. Incorporating this layer would allow for a more detailed exploration of how molecular clock components regulate cAMP synthesis and degradation. Additionally, further extensions could include downstream regulatory mechanisms such as phosphorylation and dephosphorylation events mediated by kinases and phosphatases under circadian control, potentially contributing to fine-tuned modulation of ion channel function over the daily cycle.

Together, these findings demonstrate the value of a biophysically grounded, noise-inclusive modeling framework for understanding complex temporal patterns in sensory neuron activity and open the door for future integrative models linking molecular circadian clocks to cellular excitability.

eLife Assessment

In this manuscript, the authors used in vivo long-term tip recordings of the long trichoid sensilla of male hawkmoths to analyze spontaneous spiking activity indicative of the ORNs' endogenous membrane potential oscillations. The authors combine extracellular electrophysiology of the hawkmoth antennae with computational modeling to predict that Orco receptor neuron (ORN) activity is required for circadian, not ultradian, firing patterns. The work provides valuable support for the hypothesis that a posttranslational feedback loop regulates daily and ultradian rhythms in neuronal excitability. Nevertheless, the evidence reported provides only incomplete support for their conclusions, especially with regard to the biological implications of their assumption-heavy models.

Joint Public Review:

This manuscript puts forward the provocative idea that a posttranslational feedback loop regulates daily and ultradian rhythms in neuronal excitability. The authors used in vivo long-term tip recordings of the long trichoid sensilla of male hawkmoths to analyze spontaneous spiking activity indicative of the ORNs' endogenous membrane potential oscillations. This firing pattern was disrupted by pharmacological blockade of the Orco receptor. They then use these recordings together with computational modeling to predict that Orco receptor neuron (ORN) activity is required for circadian, not ultradian, firing patterns. Orco did not show a circadian expression pattern in a qPCR experiment, and its conductance was proposed to be regulated by cyclic nucleotide levels. This evidence led the authors to conclude that a post-translational feedback loop (PTFL) clockwork, associated with the ORN plasma membrane, allows for temporal control of pheromone detection via the generation of multi-scale endogenous membrane potential oscillations. The findings will interest researchers in neurophysiology, circadian rhythms, and sensory biology. However, the manuscript has limited experimental evidence to support its central hypothesis and is undermined by several questionable assumptions that underlie their data analysis and model builds, as well as insufficient biological data, including critical controls to validate and/or fully justify the model the authors are proposing.

Strengths:

The study is notable for its combination of long-term in vivo tip recordings with computational modeling, which is technically challenging and adds weight to the authors' claims. The link between Orco, cyclic nucleotides, and circadian regulation is potentially important for sensory neuroscience, and the modeling framework itself - a stochastic Hodgkin-Huxley formulation that explicitly incorporates channel noise - is a solid and forward-looking contribution. Together, these elements make the study conceptually bold and of clear interest to circadian and olfactory biologists.

Major weaknesses:

At the same time, several limitations temper the conclusions. The pharmacological evidence relies on a single antagonist and concentration, without key controls. The circadian analysis is based on relatively small numbers of neurons, with rhythms detected only in subsets, and the alignment procedure used in constant darkness raises concerns of bias. The molecular evidence is sparse, with only three qPCR timepoints, and the model, while creative, rests on assumptions that are not yet fully supported by in vivo data.

Detailed comments are provided below:

(1) The role for Orco proposed in the authors' model largely stems from the effects seen following the administration of (a single dose) of the Orco antagonist, OLC15. However, this hypothesis is undercut by the lack of adequate pharmacological controls, including a basic multipoint OLC15 dose-response series in addition to the administration of blockers for the other channels that are embedded in their model, but which were ruled out as being involved in the modulation of biological rhythms. In addition, these studies would (ideally) also benefit from the inclusion of the same concentration (series) of an inactive OLC15 analog to better control for off-target effects.

(2) The expression pattern of Orco was assessed using qPCR at only three timepoints. Rhythmic transcripts can easily be missed with such sparse sampling (Hughes et al., 2017). A minimum of six evenly spaced timepoints across a 24-hour cycle would be required to confidently rule out circadian transcriptional regulation. In addition, the use of the timeless mRNA control from another study is not acceptable. Furthermore, qPCR analysis measures transcript abundance, not transcription, as the authors repeatedly state. Transcriptional studies would require nuclear run-off or, more recently, can be done with snRNAseq analysis. Taken together, these concerns undermine the authors' desire to rule out TTFL-based control that directly led them to implicate a PTTF-based model.

(3) The modelling presented is based on Orco as a ZT-dependent conductance tied to the cAMP oscillations that were reported by this group in the cockroach and from the presence and functionality in Manduca of homomeric Orco complexes that are devoid of tuning ORs. While these complexes have been generated in cell culture and other heterologous expression systems, as well as presumably exist in vivo in the Drosophila empty neuron and other tuning OR mutants, there is no evidence that these complexes exist in wild-type Manduca ORNs. While this doesn't necessarily undermine every aspect of their models, the authors should note the presence of Orco/OR complexes rather than Orco homomeric complexes.

(4) Some aspects of the authors' models, most notably the decision to phase align/optimize their DD and OLC15 recordings, are likely to bias their interpretations.

(5) The tip recordings from long trichoid sensilla are critical aspects of this study. These recordings were carried out on upper sensillar tips located on the distal-most second annulus. Since there are approximately 80 annuli on the Manduca antennae, it is unclear whether the recordings are representative of the antennal response.

(6) The authors do not provide any data in support of their cAMP/cGMP-based Orco gating, and the PTTF model proposed is somewhat disappointing. The model seems to be influenced by their long-held proposal that insect olfactory signaling has a critical metabotropic component involving cyclic nucleotides, PKC, etc, a view that may be influenced by the use of Orco homomeric complexes generated in HEK cells. Nevertheless, structural studies on Orco do not support a cyclic nucleotide binding site, although PKC-based phosphorylation has been implicated in the fine-tuning/adaptation of olfactory signaling.

(7) Because only 5/11 LD and 7/10 DD animals showed daily rhythms, with averages lacking clear daily modulation, the methods are not sufficiently reliable enough to reveal novel underlying mechanisms of circadian rhythm generation. The reported results are therefore not yet reliable or quantifiable. To quantify their results, the authors should apply tests for circadian rhythmicity using methods such as RAIN, JTK CYCLE, MetaCycle, or Echo. The use of FFT and Wavelet is applauded, but these methods do not have tests of significance for rhythms and can be biased when analyzing data in which there could only be 1-3 circadian cycles. Because the conclusions appear to be based on 11-12 neurons that were recorded for 2-4 days, the reader is concerned that the methods are not yet perfected to provide strong evidence for circadian regulation of spontaneous firing of ORNs. The average data (e.g., Figure 3Bii and 3Cii) highlight the apparent lack of daily rhythms. In summary, the results would be more compelling if more than 50% of the recordings had significant circadian amplitudes and with similar periods and phases.

(8) The statement that circadian patterns of ORN firing are lost with the Orco antagonist (OLC15) is not strongly supported. The manuscript should be revised to quantify how Orco changed circadian amplitude in the 12 recorded neurons. Measures of circadian amplitude can avoid confusing/vague statements like Line 394 "low and high frequency bands appeared to merge during the activity phase around ZT 0 in the animals that showed clear circadian rhythms (N = 5 of 11 in LD)". The conclusion that Orco blocks circadian firing appears to be contradicted by Figure 6, which indicates that ~6 of these neurons had circadian periods detected by wavelet. The manuscript would be strengthened with details about the specificity and reproducibility of the Orco antagonist. The authors quantify the gradual decrease in firing with the slope of a linear fit to estimate how the "effectiveness [of OLC15] increased over time." They conclude that the drug "obliterated circadian rhythms and attenuated the spontaneous activity in several, but not all experiments (N = 8 of 12)." The report would be greatly strengthened with corroborating data from additional Orco antagonists and additional doses of OLC15 (the authors use only 50 uM OLC15).

(9) The manuscript includes several statements that are more speculation than conclusion. For example, there is no evidence for tuning or plasticity in this report. Statements like the following should be removed or addressed with experiments that show changes in odor response specificity or sensitivity: "ORN signalosomes are highly plastic endogenous PTFL clocks comprising receptors for circadian and ultradian Zeitgebers that allow to tune into internal physiological and external environmental rhythms as basis for active sensing." (Discussion Line 622). The paper concludes that (line 380) "mean frequency of spontaneous spiking and the frequency of bursting expressed daily modulation, and are both most likely controlled via a circadian clock that targets the leak channel Orco." This is too bold given the available results.

(10) Because Orco conductance is modulated by cyclic nucleotides, it remains highly plausible that circadian regulation occurs upstream at the level of signaling pathways (e.g., calcium, calcium-binding proteins, GPCRs, cyclases, phosphodiesterases). The possibility that circadian oscillations of cyclic nucleotides are generated by the canonical TTFL mechanism has not been excluded. In fact, extensive work in Drosophila has demonstrated that the TTFL-based molecular clock proteins are required for circadian rhythms in olfaction.

(11) A defining feature of circadian oscillators is the feedback mechanism that generates a time delay (e.g., PERIOD/TIMELESS repressing their own transcription). While the authors describe how cyclic nucleotides can regulate Orco conductance, they do not provide a convincing explanation of how Orco activity could, in turn, feed back into the proposed PTFL to sustain oscillations. For these reasons, the authors should consider:

(a) Providing a broader discussion of non-TTFL models of circadian rhythms (e.g., redox cycles, post-translational modifications).

(b) Reassessing Orco expression using a higher-resolution temporal sampling ({greater than or equal to}6 timepoints per 24 h).

(c) Clarifying or revising the PTFL model to explicitly address how feedback would be achieved. Alternatively, the data may be more consistent with Orco conductance rhythms being regulated by post-translational mechanisms downstream of the canonical TTFL oscillator, as suggested by the Drosophila olfactory system literature.

Minor weaknesses:

(1) The authors should compare the firing patterns of ORN neurons to the bursts, clusters, and packets of retinal efferent spikes reported in Liu JS and Passaglia CL (2011; JBR). By comparing measures in moths to measures in Limulus, the authors might be able to address the question: Is the daily firing pattern of ORN neurons likely a conserved feature of circadian control of sensory sensitivity?

(2) The methods need further details. For example, it is unclear if or how single neuron activity was discriminated and whether the results were compromised by the relatively large environmental fluctuations in temperature (21-27oC), humidity (35-60%), or other cues known to modulate spontaneous firing.

Author response:

The following is the authors’ response to the original reviews.

Joint Public Review

This manuscript puts forward the provocative idea that a posttranslational feedback loop regulates daily and ultradian rhythms in neuronal excitability. The authors used in vivo long-term tip recordings of the long trichoid sensilla of male hawkmoths to analyze spontaneous spiking activity indicative of the ORNs' endogenous membrane potential oscillations. This firing pattern was disrupted by pharmacological blockade of the Orco receptor. They then use these recordings together with computational modeling to predict that Orco receptor neuron (ORN) activity is required for circadian, not ultradian, firing patterns. Orco did not show a circadian expression pattern in a qPCR experiment, and its conductance was proposed to be regulated by cyclic nucleotide levels. This evidence led the authors to conclude that a post-translational feedback loop (PTFL) clockwork, associated with the ORN plasma membrane, allows for temporal control of pheromone detection via the generation of multi-scale endogenous membrane potential oscillations. The findings will interest researchers in neurophysiology, circadian rhythms, and sensory biology. However, the manuscript has limited experimental evidence to support its central hypothesis and is undermined by several questionable assumptions that underlie their data analysis and model builds, as well as insufficient biological data, including critical controls to validate and/or fully justify the model the authors are proposing.

We thank the reviewers for their thorough and thoughtful comments and believe that the manuscript is much stronger now after the revision which incorporates the requested changes. We added results of new experiments and additional analyses. Although these new insights did not change the previous conclusions, we significantly reworked the Discussion and added further references to clarify the conclusions we want to make.

Please note that we used ORN as acronym for “olfactory receptor neuron” throughout the manuscript. ORNs contain odorant receptors (ORs), and in insects these ORs associate with the olfactory receptor co-receptor (Orco) to be trafficked to the membrane of the cilium of the ORN, where they can be contacted by pheromones and odorants. In Manduca sexta, evidence is accumulating for G-protein coupled metabotropic pheromone transduction and not for OR-Orco dependent ionotropic transduction, as shown for Drosophila melanogaster. In both insect species, besides its chaperone function, Orco can form leaky cation channels, which can regulate the spontaneous spiking activity of ORNs. In this study, we explored this role of Orco.

Strengths:

The study is notable for its combination of long-term in vivo tip recordings with computational modeling, which is technically challenging and adds weight to the authors' claims. The link between Orco, cyclic nucleotides, and circadian regulation is potentially important for sensory neuroscience, and the modeling framework itself - a stochastic Hodgkin-Huxley formulation that explicitly incorporates channel noise - is a solid and forward-looking contribution. Together, these elements make the study conceptually bold and of clear interest to circadian and olfactory biologists.

Major weaknesses:

At the same time, several limitations temper the conclusions. The pharmacological evidence relies on a single antagonist and concentration, without key controls. The circadian analysis is based on relatively small numbers of neurons, with rhythms detected only in subsets, and the alignment procedure used in constant darkness raises concerns of bias. The molecular evidence is sparse, with only three qPCR timepoints, and the model, while creative, rests on assumptions that are not yet fully supported by in vivo data.

Please see our responses to the detailed comments.

Detailed comments are provided below:

(1) The role for Orco proposed in the authors' model largely stems from the effects seen following the administration of (a single dose) of the Orco antagonist, OLC15. However, this hypothesis is undercut by the lack of adequate pharmacological controls, including a basic multipoint OLC15 dose-response series in addition to the administration of blockers for the other channels that are embedded in their model, but which were ruled out as being involved in the modulation of biological rhythms. In addition, these studies would (ideally) also benefit from the inclusion of the same concentration (series) of an inactive OLC15 analog to better control for off-target effects.

The Orco agonist VUAA1 (Jones et al., 2011) binds directly to Orco and increases the channel open time probability. In M. sexta hawkmoths, we have already published that VUAA 1 increases the low spontaneous activity of ORNs in a dose-dependent fashion (Nolte et al., 2013). Chen and Luetje (2012) systematically varied the chemical structure of VUAA1 to identify new Orco ligands and discovered 22 Orco ligand candidates (OLCs) that either activated or inhibited Orco. In their heterologous expression system, Orco was most sensitive to inhibition by OLC15. Based on these results, we published a dose-response curve of OLC15 inhibition (1-100 µM) using in vivo tip recordings of pheromone-sensitive long trichoid sensilla of M. sexta (Nolte et al., 2016). There, we also demonstrated that OLC15 dose-dependently antagonizes the VUAA1-dependent activation of Orco.

Furthermore, we tested other published Orco antagonists, which were characterized in heterologous assays, in primary cell cultures of hawkmoth ORNs, as well as in in vivo assays in intact hawkmoths. We focused on amiloride-derived antagonists, because we previously identified an amiloride-sensitive cation channel in hawkmoth ORNs. We found that, in contrast to OLC15, the amilorides HMA and MIA were not Orco-specific antagonists but instead affected different ion channel targets depending on the time of day (Nolte et al., 2016). Based on those experiments and the dose-response curves we determined that the Orco agonist VUAA1 (Jones et al., 2011) and the Orco antagonist OLC15 (Chen and Luetje, 2012) worked best in hawkmoth ORNs to target Orco pharmacologically. Due to those results and other comparative tests with other published Orco antagonists we settled since then in all further experiments on a dose of 50 µM OLC15 as most adequate to antagonize Orco functions in Manudca. In the current study, we focus on Orco without excluding the possibility that other ion channels in the ORNs contribute to the control of membrane potential rhythms.

We have clarified the Methods section accordingly.

(2) The expression pattern of Orco was assessed using qPCR at only three timepoints. Rhythmic transcripts can easily be missed with such sparse sampling (Hughes et al., 2017). A minimum of six evenly spaced timepoints across a 24-hour cycle would be required to confidently rule out circadian transcriptional regulation. In addition, the use of the timeless mRNA control from another study is not acceptable. Furthermore, qPCR analysis measures transcript abundance, not transcription, as the authors repeatedly state. Transcriptional studies would require nuclear run-off or, more recently, can be done with snRNAseq analysis. Taken together, these concerns undermine the authors' desire to rule out TTFL-based control that directly led them to implicate a PTTF-based model.

We agree with the referees that more time points and a direct comparison between timeless and Orco mRNA levels should be included in this manuscript. We included these additional qPCR experiments and edited the manuscript to make clear that we measure transcript abundance, but we will not perform snRNAseq analysis due to time- and financial constraints.

(3) The modelling presented is based on Orco as a ZT-dependent conductance tied to the cAMP oscillations that were reported by this group in the cockroach and from the presence and functionality in Manduca of homomeric Orco complexes that are devoid of tuning ORs. While these complexes have been generated in cell culture and other heterologous expression systems, as well as presumably exist in vivo in the Drosophila empty neuron and other tuning OR mutants, there is no evidence that these complexes exist in wild-type Manduca ORNs. While this doesn't necessarily undermine every aspect of their models, the authors should note the presence of Orco/OR complexes rather than Orco homomeric complexes.

Our ELISAs found circadian oscillations in cAMP levels not only in antennae of the Madeira cockroach (Schendzielorz et al., 2014, 2012), but also in hawkmoth antennae (Schendzielorz et al., 2015). For clarification, we added the 2015 citation to the Modeling chapter in the Methods section.

We agree with the referees that we cannot distinguish between Orco homo- and heteromers in the different compartments of our hawkmoth ORNs but we know that both are expressed in the pheromone-sensitive ORNs. Thus, as the referee suggests, we added text regarding the presence and localization of OR-Orco heteromers. Consistent data collected across different experiments (heterologous expression systems, primary cell cultures of hawkmoth ORNs, in vivo/in situ studies) support that Orco homomers are present in hawkmoth ORNs. In addition to co-expression of MsexOrco and MsexSNMP-1 with either MsexOr-1 or MsexOr-4 in a heterologous expression system, MsexOrco expression alone was already sufficient to increase intracellular Ca2+ levels spontaneously as a result of its property as leaky, non-specific cation channel, and in response to VUAA1 application (Nolte et al., 2013). Both in developing hawkmoth pupae and differentiating primary cell cultures of hawkmoth ORNs, Orco expression started during a developmental time window where ORNs did not yet express pheromone receptors but where Orco affected spontaneous activity and intracellular Ca2+ levels dependent on VUAA1 (Nolte et al., 2016). In vitro patch clamp studies of differentiating cultured hawkmoth ORNs during this time window of pupal development characterized ion channels/currents with properties of Orco as a leaky, non-specific cation channel/current that depends on protein kinase C and cyclic nucleotides (Dolzer et al., 2021, 2008; Krannich and Stengl, 2008; Stengl, 1993). Thus, Orco homomers are present in developing hawkmoth ORNs during a time window where ORNs already express spontaneous activity but they do not heteromerize with pheromone receptors. However, we do not know whether and in what ratio homo- and heteromers of Orco and ORs are present in the respective sensillum compartments of adult hawkmoths because all OR-specific antibodies tested did not work in immunocytochemical studies of hawkmoth antennae (Nolte et al., 2013; Stengl, 1994; Stengl and Hildebrand, 1990). Our hypothesis of differential distribution of Orco homomers in the some and dendrite compartment, and OR-Orco heteromers in the cilia is based on differential immunocytochemical localization of Drosophila ORs mainly in the cilia compartment (Benton et al., 2006).

We clarified our manuscript accordingly.

(4) Some aspects of the authors' models, most notably the decision to phase align/optimize their DD and OLC15 recordings, are likely to bias their interpretations.

It is consensus that insects display daily and circadian rhythms in pheromone-dependent mating, odor-gated feeding, and egg-laying behavior that phase-locks to environmental rhythms, corresponding with daily/circadian rhythms of sensory neuron physiology (e.g., Merlin et al., 2007; Rymer et al., 2007; Schendzielorz et al., 2015, 2012). However, circadian rhythms can be easily masked by stress, like the disturbances during an experimentally very challenging long-term recording experiment over several days. In addition, we observed over the years in our animal raising facility that in 17:7 light-dark cycles the originally nocturnal hawkmoths M. sexta distribute their activity patterns over the course of the day, finding nocturnal as well as diurnal hawkmoths. Thus, light-dark cycles were not enough to ensure phase-synchronized behavioral rhythms, and it is very likely that the nocturnal hawkmoths, next to stress signals, rely heavily on pheromone/odor dependent synchronization as also found in other moth species (Ghosh et al., 2024). Because we focus on spontaneous activity and not on pheromone-dependent physiology in this study, we used isolated males that were never exposed to the female pheromones, taking phase dispersal into account. Therefore, it became necessary in free-running conditions to first determine the respective behavioral rhythm for each animal, and then to phase-align their activity patterns to allow for statistical analysis. Otherwise, circadian differences would average out in a phase-dispersed free-running population. As requested by the referees in point (7), we added RAIN to test for rhythmicity in each of our recordings and revised the manuscript accordingly.

Furthermore, in preliminary experiments we briefly exposed hawkmoths to pheromone the night before the start of the experiment. However, we failed to obtain phase-synchronized spiking rhythms. Most likely, a circadian pattern of pheromone exposure would have been necessary as zeitgeber, which could not be used here due to long-term pheromone-dependent effects in spiking activity. These results are added as supplementary figure to Fig 3.

(5) The tip recordings from long trichoid sensilla are critical aspects of this study. These recordings were carried out on upper sensillar tips located on the distal-most second annulus. Since there are approximately 80 annuli on the Manduca antennae, it is unclear whether the recordings are representative of the antennal response.

We think the reviewers might have misinterpreted our description of the recording site. In the Methods, we state that we clip off the 20 most distal annuli (leaving a stump of about 60 annuli) and insert the reference electrode into the flagellum up to the second annulus from the cut end, i.e., the recording sites are located at 2/3 – 3/4 of the antenna length as seen from the head of the animal. We clarified this in the Methods section.

In addition, our lab did show with antibody stainings against Orco that apparently all ORNs that innervate long and short trichoid sensilla along the whole flagellum express the same staining pattern (Nolte et al., 2016). Lee and Strausfeld (1990) mapped all types of antennal sensilla, and together with pheromone-dependent tip-recordings of Kaissling et al. (1989) it was shown that most of the male antennal sensilla are pheromone-sensitive long trichoid sensilla, with one of the two innervating ORNs always responding to bombykal, ensuring high sensitivity to pheromone detection. Furthermore, our patch clamp recordings of primary cell cultures of whole male antennae found largely overlapping ion channel populations across ORNs (review: (Stengl, 2010)). This would indicate that all ORNs, whether they express ORs sensitive to pheromone or general odorants, could potentially share the same Orco-dependent spontaneous activity rhythms. Furthermore, in our lab, different experimenters from different years that recorded from long trichoid sensilla on different annuli did not detect obvious differences in neither the spontaneous activity nor the pheromone responses (c.f., Dolzer et al., 2003; Gawalek and Stengl, 2018; Schneider et al., 2025). Thus, it is very likely that we are reporting a general encoding mechanism that is not locally restricted along the antennal flagellum and is very likely shared by all types of OR-Orco expressing ORNs.

(6.1) The authors do not provide any data in support of their cAMP/cGMP-based Orco gating…

There are publications supporting cyclic nucleotide gating of Orco in Drosophila, but only after previous phosphorylation via protein kinase C (PKC; review: (Wicher and Miazzi, 2021)). Since Orco is very conserved among insect species, it is likely that PKC- and cGMP/cAMP-dependent regulations are present for Orco in other insect species. To test this, we are currently characterizing second messenger-dependence of spontaneous spiking activity, which is the focus of a follow-up manuscript. Nevertheless, to provide more evidence for our hypothesis of the current manuscript, we added a new set of tip-recording experiments that demonstrate cAMP-dependent gating of Orco. Because of the addition of this figure, we merged figures 8-10 into Figure 8 and added the cAMP data as Figure 9.

(6.2) … and the PTTF model proposed is somewhat disappointing.

For a detailed introduction of our PTFL membrane clock hypothesis please see our opinion paper that we refer to in the manuscript (Stengl and Schneider, 2024). We added clarification of how Orco activation can influence cAMP levels. A more elaborate PTFL clock model including many more of the identified ion channels in hawkmoth ORNs is the focus of another manuscript to come.

(6.3) The model seems to be influenced by their long-held proposal that insect olfactory signaling has a critical metabotropic component involving cyclic nucleotides, PKC, etc, a view that may be influenced by the use of Orco homomeric complexes generated in HEK cells.

Indeed, we propose a metabotropic pheromone-transduction cascade, which in moths and cockroaches is based on G-protein-mediated activation of phospholipase C but not on adenylyl cyclase activation. Our hypothesis is not influenced by HEK cell heterologous expression studies of Orco but is supported by our own work comparing in vivo tip recordings of intact hawkmoths with patch clamp experiments on hawkmoth primary cell cultures of olfactory receptor neurons, which are able to respond to their species-specific pheromones in vitro (Schneider et al., 2025; Stengl, 2010; Stengl and Funk, 2013; Wicher and Miazzi, 2021). In addition, a multitude of publications by other laboratories with in vivo and in vitro studies using physiological, genetic, and immunocytochemical assays all support a metabotropic signal transduction cascade in insect olfaction (Stengl, 2010; Stengl and Funk, 2013; Takagi et al., 2025; Wicher and Miazzi, 2021). In contrast, the hypothesis suggesting a solely ionotropic pheromone- and general odor-dependent transduction cascade for all insect species is based on very sparse experimental evidence, based primarily on heterologous expression studies such as HEK cells that lack the insect’s WT molecular surroundings, and thus, cannot predict OR-Orco function in vivo. Furthermore, the ionotropic hypothesis is heavily based upon the argument that an inverse 7TM receptor cannot couple to G-proteins, which lacks careful backup via biochemical and structural studies. In addition, the ionotropic hypothesis lacks support via carefully performed physiological in vivo studies in different insect species that paid attention to analysis of the distinct kinetic components of ORN´s odor/pheromone responses and that employ physiological concentrations and durations of odor/pheromone stimuli (please see our most recent publication by Schneider et al. (2025)). We added references to the possible odor transduction mechanisms to the introduction.

(6.4) Nevertheless, structural studies on Orco do not support a cyclic nucleotide binding site, although PKC-based phosphorylation has been implicated in the fine-tuning/adaptation of olfactory signaling.

While structural studies did not find evidence for conserved known cyclic nucleotide binding sites on Orco, this does not exclude the presence of indirect cAMP effects via e.g., Orco subunits complexing with other molecules under direct cAMP control, such as other ion channel subunits. Furthermore, it does not exclude so far unknown binding sites, or via sites that fold out only after a specific sequence of previous phosphorylations of the many phosphorylation sites on Orco. Indeed, physiological studies in Drosophila presented evidence for cyclic nucleotide dependence of Orco after previous PKC-dependent phosphorylation (Getahun et al., 2013). Our ongoing in vivo experiments in hawkmoths further corroborate a zeitgeber time-dependent PKC- and cyclic nucleotide-dependent modulation of Orco. These detailed studies will be published in a follow-up publication. In the revised version of this manuscript, we added tip-recording experiments that indicate cAMP involvement in Orco gating (new Figure 9).

(7) Because only 5/11 LD and 7/10 DD animals showed daily rhythms, with averages lacking clear daily modulation, the methods are not sufficiently reliable enough to reveal novel underlying mechanisms of circadian rhythm generation. The reported results are therefore not yet reliable or quantifiable. To quantify their results, the authors should apply tests for circadian rhythmicity using methods such as RAIN, JTK CYCLE, MetaCycle, or Echo. The use of FFT and Wavelet is applauded, but these methods do not have tests of significance for rhythms and can be biased when analyzing data in which there could only be 1-3 circadian cycles. Because the conclusions appear to be based on 11-12 neurons that were recorded for 2-4 days, the reader is concerned that the methods are not yet perfected to provide strong evidence for circadian regulation of spontaneous firing of ORNs. The average data (e.g., Figure 3Bii and 3Cii) highlight the apparent lack of daily rhythms. In summary, the results would be more compelling if more than 50% of the recordings had significant circadian amplitudes and with similar periods and phases.

The long-term tip-recordings of intact hawkmoths are very challenging and take a very long time to accomplish, thus, we are very happy that we succeeded in obtaining so many of them (N=40). We are thankful to the reviewers’ suggestion to use RAIN since this analysis revealed circadian rhythms in 7 of 11 LD recordings, 8 of 12 DD recordings, and 2 of 12 OLC15 recordings. Please see also our response to (4) above, commenting the phase-dispersal of activity rhythms observed in our experiments, as well as in the behavior of hawkmoth males in the mating cage.

(8) The statement that circadian patterns of ORN firing are lost with the Orco antagonist (OLC15) is not strongly supported. The manuscript should be revised to quantify how Orco changed circadian amplitude in the 12 recorded neurons. Measures of circadian amplitude can avoid confusing/vague statements like Line 394 “low and high frequency bands appeared to merge during the activity phase around ZT 0 in the animals that showed clear circadian rhythms (N = 5 of 11 in LD)”. The conclusion that Orco blocks circadian firing appears to be contradicted by Figure 6, which indicates that ~6 of these neurons had circadian periods detected by wavelet. The manuscript would be strengthened with details about the specificity and reproducibility of the Orco antagonist. The authors quantify the gradual decrease in firing with the slope of a linear fit to estimate how the “effectiveness [of OLC15] increased over time.” They conclude that the drug “obliterated circadian rhythms and attenuated the spontaneous activity in several, but not all experiments (N = 8 of 12).” The report would be greatly strengthened with corroborating data from additional Orco antagonists and additional doses of OLC15 (the authors use only 50 uM OLC15).

According to the valuable suggestions of the referees, we used RAIN to detect circadian rhythms in the spiking attributes in each individual animal. Since only 2 of 12 animals displayed a circadian rhythm in OLC15, statistical comparison of circadian amplitudes is not possible. We revised the results section accordingly and added to the figure legend to make it clearer that the heat maps in Fig 5 are representative from one animal each and not averages across animals.

As the reviewer states correctly in (7), wavelet results of circadian rhythmicity must be interpreted carefully because of the low number of circadian cycles in ~3-4 day recordings. Since the heatmaps in Figure 5 visually revealed the presence of ultradian rhythms, the main focus of the wavelet analysis in Figure 6 is in the detection and quantification of ultradian periods up to 20 h.

We revised the Methods section to include references to previous experiments that characterized the effect of different doses of OLC15 and other Orco antagonists and agonists in M. sexta antennae (Nolte et al., 2016). Please see also our response to (1).

(9) The manuscript includes several statements that are more speculation than conclusion. For example, there is no evidence for tuning or plasticity in this report. Statements like the following should be removed or addressed with experiments that show changes in odor response specificity or sensitivity: "ORN signalosomes are highly plastic endogenous PTFL clocks comprising receptors for circadian and ultradian Zeitgebers that allow to tune into internal physiological and external environmental rhythms as basis for active sensing." (Discussion Line 622). The paper concludes that (line 380) "mean frequency of spontaneous spiking and the frequency of bursting expressed daily modulation, and are both most likely controlled via a circadian clock that targets the leak channel Orco." This is too bold given the available results.

We revised the manuscript accordingly and clarified which statements are supported via published evidence and which are predictions based upon our novel hypothesis published in our opinion paper (Stengl and Schneider, 2024).

(10.1) Because Orco conductance is modulated by cyclic nucleotides, it remains highly plausible that circadian regulation occurs upstream at the level of signaling pathways (e.g., calcium, calcium-binding proteins, GPCRs, cyclases, phosphodiesterases).

We agree with the referees that it is very likely that there are multiple layers of interconnected feedback cycles that control Orco localization and activity. Our novel hypothesis suggests interlocked TTFL and PTFL control of physiological circadian rhythms, not strictly hierarchical TTFL control, which would require a daily turnover of membrane proteins and transcriptional control via the established TTFL clock in insect ORNs. We are currently searching for TTFL control at all levels of odor/pheromone transduction using ZT-dependent transcriptomics in combination with qPCR and single-nucleus transcriptomics, involving also all the molecules suggested by the referees. These studies are ongoing, are very time- and money-consuming, and are beyond the scope of this manuscript. However, we added a set of experiments to this manuscript in which we demonstrate that the effect of increased cAMP on the spontaneous spiking activity is mediated by Orco (new Figure 9).

(10.2) The possibility that circadian oscillations of cyclic nucleotides are generated by the canonical TTFL mechanism has not been excluded. In fact, extensive work in Drosophila has demonstrated that the TTFL-based molecular clock proteins are required for circadian rhythms in olfaction.

Our experiments that test circadian TTFL control at different levels of the cAMP transduction cascade in hawkmoth antennae are on the way and are part of another publication. In section 6.2 we already stated that our experiments do not exclude that Orco is under indirect control of the TTFL. We revised our discussion accordingly.

The experiments published for TTFL dependent control of Drosophila olfaction that we are aware of (Krishnan et al., 1999; Tanoue et al., 2004) do not exclude interlinked PTFL and TTFL clocks. Krishnan et al. (1999) demonstrated that the TTFL clock in antennal olfactory receptor neurons correlates with circadian rhythms in odor responses measured in electroantennogram (EAG) recordings, not in single sensillum recordings as in our experiments. EAG recordings comprise not only voltage responses of the olfactory sensory neurons but also voltage changes generated in non-neuronal antennal cells such as trichogen and tormogen cells that built the transepithelial potential gradient via vATPases that generates the high K+ concentration in the sensillum lymph (Jain et al., 2024; Klein, 1992; Thurm and Küppers, 1980). In addition, EAG recordings most likely contain responses of afferent neurons originating from somata in the brain that maintain central control of the antennae. Thus, EAG recordings are difficult to interpret.

(11) A defining feature of circadian oscillators is the feedback mechanism that generates a time delay (e.g., PERIOD/TIMELESS repressing their own transcription). While the authors describe how cyclic nucleotides can regulate Orco conductance, they do not provide a convincing explanation of how Orco activity could, in turn, feed back into the proposed PTFL to sustain oscillations. For these reasons, the authors should consider:

(a) Providing a broader discussion of non-TTFL models of circadian rhythms (e.g., redox cycles, post-translational modifications).

We revised the discussion accordingly.

(b) Reassessing Orco expression using a higher-resolution temporal sampling ({greater than or equal to}6 timepoints per 24 h).

We added those experiments to the revised version of the manuscript (see our response to (2)).

(c) Clarifying or revising the PTFL model to explicitly address how feedback would be achieved. Alternatively, the data may be more consistent with Orco conductance rhythms being regulated by post-translational mechanisms downstream of the canonical TTFL oscillator, as suggested by the Drosophila olfactory system literature.

We added possible negative feedback elements to the Discussion to explain how our proposed PTFL could in principle work independent of TTFL clock.

Minor weaknesses:

(1) The authors should compare the firing patterns of ORN neurons to the bursts, clusters, and packets of retinal efferent spikes reported in Liu JS and Passaglia CL (2011; JBR). By comparing measures in moths to measures in Limulus, the authors might be able to address the question: Is the daily firing pattern of ORN neurons likely a conserved feature of circadian control of sensory sensitivity?

We have revised the discussion accordingly.

(2) The methods need further details. For example, it is unclear if or how single neuron activity was discriminated and whether the results were compromised by the relatively large environmental fluctuations in temperature (21-27oC), humidity (35-60%), or other cues known to modulate spontaneous firing.

These large fluctuations stem from doing experiments at different seasons (higher temperature and humidity in the summer months, lower temperature and humidity in winter). Throughout each individual experiment, conditions were stable. We clarified the Methods section accordingly.

Recommendations for the authors:

The authors should post the code for their computational model to a repository like GitHub.

The code for the computational model is now available at https://github.com/a-c-schneider/VijayanForlinoEtAl2025_Model.git

References

Benton R, Sachse S, Michnick SW, Vosshall LB. 2006. Atypical Membrane Topology and Heteromeric Function of Drosophila Odorant Receptors In Vivo. PLOS Biology 4:e20. DOI: https://doi.org/10.1371/journal.pbio.0040020

Chen S, Luetje CW. 2012. Identification of New Agonists and Antagonists of the Insect Odorant Receptor Co-Receptor Subunit. PLOS ONE 7:e36784. DOI: https://doi.org/10.1371/journal.pone.0036784

Dolzer J, Fischer K, Stengl M. 2003. Adaptation in pheromone-sensitive trichoid sensilla of the hawkmoth Manduca sexta. Journal of Experimental Biology 206:1575–1588. DOI: https://doi.org/10.1242/jeb.00302

Dolzer J, Krannich S, Stengl M. 2008. Pharmacological Investigation of Protein Kinase C- and cGMP-Dependent Ion Channels in Cultured Olfactory Receptor Neurons of the Hawkmoth Manduca sexta. Chemical Senses 33:803–813. DOI: https://doi.org/10.1093/chemse/bjn043

Dolzer J, Schröder K, Stengl M. 2021. Cyclic nucleotide-dependent ionic currents in olfactory receptor neurons of the hawkmoth Manduca sexta suggest pull–push sensitivity modulation. European Journal of Neuroscience 54:4804–4826. DOI: https://doi.org/10.1111/ejn.15346

Gawalek P, Stengl M. 2018. The Diacylglycerol Analogs OAG and DOG Differentially Affect Primary Events of Pheromone Transduction in the Hawkmoth Manduca sexta in a Zeitgebertime-Dependent Manner Apparently Targeting TRP Channels. Frontiers in Cellular Neuroscience 12:218. DOI: https://doi.org/10.3389/fncel.2018.00218

Getahun MN, Olsson SB, Lavista-Llanos S, Hansson BS, Wicher D. 2013. Insect Odorant Response Sensitivity Is Tuned by Metabotropically Autoregulated Olfactory Receptors. PLOS ONE 8:e58889. DOI: https://doi.org/10.1371/journal.pone.0058889

Ghosh S, Suray C, Bozzolan F, Palazzo A, Monsempès C, Lecouvreur F, Chatterjee A. 2024. Pheromone-mediated command from the female to male clock induces and synchronizes circadian rhythms of the moth Spodoptera littoralis. Current biology 34:1414-1425.e5. DOI: https://doi.org/10.1016/j.cub.2024.02.042, PMID: 38479388

Jain K, Prelic S, Hansson BS, Wicher D. 2024. Expression of Drosophila melanogaster V-ATPases in Olfactory Sensillum Support Cells. Insects 15:1016. DOI: https://doi.org/10.3390/insects15121016

Jones PL, Pask GM, Rinker DC, Zwiebel LJ. 2011. Functional agonism of insect odorant receptor ion channels. Proceedings of the National Academy of Sciences 108:8821–8825. DOI: https://doi.org/10.1073/pnas.1102425108

Kaissling KE, Hildebrand JG, Tumlinson JH. 1989. Pheromone receptor cells in the male moth Manduca sexta. Archives of Insect Biochemistry and Physiology 10:273–279. DOI: https://doi.org/10.1002/arch.940100403

Klein U. 1992. The insect V-ATPase, a plasma membrane proton pump energizing secondary active transport: immunological evidence for the occurrence of a V-ATPase in insect ion-transporting epithelia. Journal of Experimental Biology 172:345–354. DOI: https://doi.org/10.1242/jeb.172.1.345

Krannich S, Stengl M. 2008. Cyclic Nucleotide-Activated Currents in Cultured Olfactory Receptor Neurons of the Hawkmoth Manduca sexta. Journal of Neurophysiology 100:2866–2877. DOI: https://doi.org/10.1152/jn.01400.2007

Krishnan B, Dryer SE, Hardin PE. 1999. Circadian rhythms in olfactory responses of Drosophila melanogaster. Nature 400:375–378. DOI: https://doi.org/10.1038/22566

Lee JK, Strausfeld NJ. 1990. Structure, distribution and number of surface sensilla and their receptor cells on the olfactory appendage of the male mothManduca sexta. Journal of Neurocytology 19:519–538. DOI: https://doi.org/10.1007/BF01257241

Merlin C, Lucas P, Rochat D, François M-C, Maïbèche-Coisne M, Jacquin-Joly E. 2007. An Antennal Circadian Clock and Circadian Rhythms in Peripheral Pheromone Reception in the Moth Spodoptera littoralis. Journal of Biological Rhythms 22:502–514. DOI: https://doi.org/10.1177/0748730407307737

Nolte A, Funk NW, Mukunda L, Gawalek P, Werckenthin A, Hansson BS, Wicher D, Stengl M. 2013. In situ Tip-Recordings Found No Evidence for an Orco-Based Ionotropic Mechanism of Pheromone-Transduction in Manduca sexta. PLOS ONE 8:e62648. DOI: https://doi.org/10.1371/journal.pone.0062648

Nolte A, Gawalek P, Koerte S, Wei H, Schumann R, Werckenthin A, Krieger J, Stengl M. 2016. No Evidence for Ionotropic Pheromone Transduction in the Hawkmoth Manduca sexta. PLOS ONE 11:e0166060. DOI: https://doi.org/10.1371/journal.pone.0166060

Rymer J, Bauernfeind AL, Brown S, Page TL. 2007. Circadian rhythms in the mating behavior of the cockroach, Leucophaea maderae. Journal of Biological Rhythms 22:43–57. DOI: https://doi.org/10.1177/0748730406295462, PMID: 17229924

Schendzielorz J, Schendzielorz T, Arendt A, Stengl M. 2014. Bimodal Oscillations of Cyclic Nucleotide Concentrations in the Circadian System of the Madeira Cockroach Rhyparobia maderae. Journal of Biological Rhythms 29:318–331. DOI: https://doi.org/10.1177/0748730414546133

Schendzielorz T, Peters W, Boekhoff I, Stengl M. 2012. Time of Day Changes in Cyclic Nucleotides Are Modified via Octopamine and Pheromone in Antennae of the Madeira Cockroach. Journal of Biological Rhythms 27:388–397. DOI: https://doi.org/10.1177/0748730412456265

Schendzielorz T, Schirmer K, Stolte P, Stengl M. 2015. Octopamine Regulates Antennal Sensory Neurons via Daytime-Dependent Changes in cAMP and IP3 Levels in the Hawkmoth Manduca sexta. PLOS ONE 10:e0121230. DOI: https://doi.org/10.1371/journal.pone.0121230

Schneider AC, Schröder K, Chang Y, Nolte A, Gawalek P, Stengl M. 2025. Hawkmoth Pheromone Transduction Involves G-Protein–Dependent Phospholipase Cβ Signaling. eNeuro 12:ENEURO.0376-24.2024. DOI: https://doi.org/10.1523/ENEURO.0376-24.2024, PMID: 39880675

Stengl M. 2010. Pheromone Transduction in Moths. Frontiers in Cellular Neuroscience 4:133. DOI: https://doi.org/10.3389/fncel.2010.00133

Stengl M. 1994. Inositol-trisphosphate-dependent calcium currents precede cation currents in insect olfactory receptor neurons in vitro. Journal of Comparative Physiology A 174:187–194. DOI: https://doi.org/10.1007/BF00193785

Stengl M. 1993. Intracellular-Messenger-Mediated Cation Channels in Cultured Olfactory Receptor Neurons. Journal of Experimental Biology 178:125–147. DOI: https://doi.org/10.1242/jeb.178.1.125

Stengl M, Funk NW. 2013. The role of the coreceptor Orco in insect olfactory transduction. Journal of Comparative Physiology A 199:897–909. DOI: https://doi.org/10.1007/s00359-013-0837-3

Stengl M, Hildebrand JG. 1990. Insect olfactory neurons in vitro: morphological and immunocytochemical characterization of male-specific antennal receptor cells from developing antennae of male Manduca sexta. Journal of Neuroscience 10:837–847. DOI: https://doi.org/10.1523/JNEUROSCI.10-03-00837.1990, PMID: 2319305

Stengl M, Schneider AC. 2024. Contribution of membrane-associated oscillators to biological timing at different timescales. Frontiers in Physiology 14:1243455. DOI: https://doi.org/10.3389/fphys.2023.1243455

Takagi S, Abuin L, Mermet J, Lee D, Benton R. 2025. A GPCR signaling pathway in insect odor detection. DOI: https://doi.org/10.1101/2025.10.03.680299

Tanoue S, Krishnan P, Krishnan B, Dryer SE, Hardin PE. 2004. Circadian Clocks in Antennal Neurons Are Necessary and Sufficient for Olfaction Rhythms in Drosophila. Current Biology 14:638–649. DOI: https://doi.org/10.1016/j.cub.2004.04.009, PMID: 15084278

Thurm U, Küppers J. 1980. Epithelial physiology of insect sensilla. In: Locke M, Smith DS (Eds). Insect Biology in the Future. Academic Press. p. 735–763. DOI: https://doi.org/10.1016/B978-0-12-454340-9.50039-2

Wicher D, Miazzi F. 2021. Functional properties of insect olfactory receptors: ionotropic receptors and odorant receptors. Cell and Tissue Research 383:7–19. DOI: https://doi.org/10.1007/s00441-020-03363-x

Version 2

Abstract

The mating behavior of nocturnal Manduca sexta hawkmoths is under strict temporal control. It is orchestrated via circadian and ultradian oscillations in sex-pheromone stimuli as social zeitgeber. The extremely sensitive pheromone-detecting olfactory receptor neurons (ORNs) that innervate the long trichoid sensilla on the male’s antennae are peripheral circadian clocks. They express the transcriptional-translational feedback loop (TTFL) circadian clockwork, best characterized in Drosophila melanogaster. In hawkmoths, it is still unknown whether or how the ORN TTFL clockwork regulates the daily rhythms in pheromone sensitivity and in temporal resolution of ultradian pheromone pulses as prerequisites to the temporal regulation of hawkmoth mating behavior.

We hypothesize that, rather than the slow TTFL clock, a more rapidly adaptive post-translational feedback loop (PTFL) clockwork, associated with the ORN plasma membrane, allows for temporal control of pheromone detection via generation of multiscale endogenous membrane potential oscillations. The potential oscillations of the PTFL clock could rapidly synchronize to oscillations of pheromone stimuli at different timescales, thus enabling the prediction of stimulus patterns as a mechanism for active sensing. With in vivo long-term tip recordings of long trichoid sensilla of male hawkmoths, we analyzed the spontaneous spiking activity indicative of the ORNs’ endogenous membrane potential oscillations. Consistent with our hypothesis of a multiscale PTFL clock in hawkmoth ORNs, spontaneous spiking was modulated on ultradian and circadian timescales, with maximum activity at night. When we blocked the evolutionarily conserved olfactory receptor coreceptor (Orco), the circadian modulation was abolished but the ultradian frequency modulation of the spontaneous activity remained. Consistent with PTFL control, Orco was not under the transcriptional control of the TTFL clock, but its modulation was dependent on cAMP. We could replicate the experimental data in a conductance-based computational model of an ORN. In this model, Orco conductivity changed as a function of fluctuating 2nd messenger levels. This study demonstrates that a PTFL clock is sufficient to impose a circadian pattern on ORN sensitivity.

Introduction

Biological clocks generate endogenous physiological and behavioral oscillations on multiple timescales that persist in the absence of any external timing cues under constant conditions (Lamont and Amir, 2010). Such clocks are essential to generate internal rhythms and to permit their synchronization to external zeitgebers at timescales from milliseconds to years. These synchronized endogenous multiscale clocks thus provide both a continuous time frame for physiology and behavior - the organism’s “present time” - and enable organisms to predict and respond to regular, periodic environmental changes (Stengl and Schneider, 2024). Even though endogenous clocks are widespread in prokaryotes and eukaryotes (Géron et al., 2024; Hall and Rosbash, 1993; Millar, 2016; Thaiss et al., 2014), current knowledge about how multiscale clocks are linked over widely diverging timescales remains sparse.

Insect circadian clock neurons, which are tightly connected to visual brain circuits, are amongst the best studied biological clocks (Hall and Rosbash, 1993; Nishiitsutsuji-Uwo and Pittendrigh, 1968; Reinhard et al., 2024). They regulate physiological and behavioral oscillations that underly sleep-wake rhythms tied to the 24 h light-dark cycles on Earth (Dubowy and Sehgal, 2017). Circadian clock neurons possess a molecular clockwork of transcriptional-translational feedback loops (TTFLs). This TTFL clock generates endogenous oscillations of clock gene mRNA and protein levels with a cycle period of ∼24 h via negative feedback by proteins that inhibit their own transcription (Hardin, 2011). It is assumed for both insects and mammals that this TTFL comprises the master clockwork that regulates all physiological and behavioral rhythms in the circadian time range through genetically controlled output pathways (Hastings et al., 2020).

Molecular TTFL clockworks were not only found in central brain clock neurons but also in sensory neurons, such as olfactory receptor neurons (ORNs) in insect antennae, where they play a role in odor-dependent circadian behaviors like feeding, pollination, reproduction, and navigation (Flecke et al., 2010; Flecke and Stengl, 2009; Krishnan et al., 1999; Merlin et al., 2009, 2006; Page and Koelling, 2003; Plautz et al., 1997; Sauman and Reppert, 1998; Schendzielorz et al., 2015; Schuckel et al., 2007; Tanoue et al., 2004; Zhou et al., 2005). Little is known about how the TTFL clockwork of insect ORNs regulates circadian rhythms in chemosensory sensitivity that underlie the circadian control of these different behaviors.

At night, male nocturnal Manduca sexta hawkmoths are very sensitive to the sex pheromone blend that their conspecific females release in an intermittent pattern (Itagaki and Conner, 1988). The intermittency is a prerequisite to maintain the male’s search flight, which is suspended at pulse frequencies >30 Hz (Baker et al., 1985; Baker and Vogt, 1988; Stengl, 2010). The pheromone pulses are resolved of up to 3 Hz by pheromone-specific ORNs in the antenna’s long trichoid sensilla (Lee and Strausfeld, 1990; Marion-Poll and Tobin, 1992). Pheromone-specificity is provided by odor receptor (OR) subunits in the ciliary membranes of ORNs that heteromerize with the conserved olfactory receptor co-receptor Orco (Nakagawa and Vosshall, 2009; Stengl and Funk, 2013; Wicher and Miazzi, 2021). The insect odor transduction cascade is still under debate and seems to involve both ionotropic and metabotropic pathways (Sato et al., 2008; Schneider et al., 2025; Wicher et al., 2008). While there is no general agreement on the role of Orco, it is known that in heterologous expression systems Orco homo- and heteromers form a slow, leaky, non-specific cation channel, which depolarizes the cells (Jones et al., 2011; Sargsyan et al., 2011; Sato et al., 2008; Wicher, 2018; Wicher et al., 2008). In vivo tip recordings of long trichoid sensilla in M. sexta confirmed that Orco acts as leaky cation channel, acting as a pacemaker channel, because it promotes ultradian membrane potential oscillations that underly the spontaneous action potential activity (Nolte et al., 2013). These potential oscillations turn the ORNs into ultradian temporal filters that are sensitive to the frequency of the pheromone pulses they encounter. Pheromone pulses that arrive at a similar frequency as the endogenous sub-threshold membrane potential oscillations would synchronize these oscillations and would be more likely to elicit action potential trains. Therefore, this pacemaker channel-dependent mechanism would allow for active sensing if endogenous membrane potential oscillations are in the physiological range that is necessary for tracking intermittent pheromone pulses.

It is not known whether the Orco-dependent membrane potential oscillations also occur at circadian timescales, and whether or how they are linked to ultradian potential oscillations. One hypothesis is that the molecular TTFL clockwork of ORNs exclusively controls endogenous circadian rhythms of the membrane potential via circadian expression of pacemaker channels like Orco. Alternatively, the plasma membrane of ORNs could by itself constitute an autonomous clock that is entrained by but does not rely on the TTFL clock. Constant levels of ion channels that do not require daily degradation and daily transcription via the TTFL clock, but with modulation of open-time probability by posttranslational modifications, would constitute a more economical and faster adjustable mechanism to generate endogenous membrane potential oscillations (Stengl and Schneider, 2024). This would facilitate active sensing by rapidly tuning odorant/pheromone detection to behaviorally relevant zeitgeber signals in the insect’s environment.

Here, we performed a combination of experimental and computational modeling studies to investigate whether Orco plays a role in active sensing at the circadian and/or ultradian timescale via generation of endogenous membrane potential oscillations. We hypothesize that Orco is part of a post-transcriptional feedback loop (PTFL) clockwork in the plasma membrane, rather than being TTFL controlled. To capture naturally occurring, physiologically relevant processes in insect ORNs we performed in vivo long-term tip recordings of the spontaneous spiking activity of pheromone-sensitive trichoid sensilla of intact but restrained hawkmoth males.

We corroborated the experimental data with a Hodgkin-Huxley (HH) based computational ORN model. In general, the widely used deterministic HH models do not include circadian control of ion channels. Moreover, the highly dynamic, bursting firing patterns of ORNs cannot be described by the traditional deterministic modeling of neuronal activity. This constitutes a drawback because stochasticity plays a crucial role in shaping firing patterns which are influenced not only by external stimuli but also by the inherent noise involved in channel gating. Therefore, to address the above points realistically, we developed a Langevin formulation of the HH model, which explicitly accounts for the noise introduced by the random opening and closing of ion channels and circadian modulation of the Orco channel conductivity. By integrating channel noise, we could explore how circadian regulation of ionic currents contributes to the variability of firing patterns and how this noise interacts with external stimuli to influence olfactory signal processing.

Our model aims to highlight the advantages of using a stochastic framework for modeling the circadian activity of ORNs. The combined experimental and modeling efforts generate a more complete picture and predictions concerning these autonomous processes in sensory neurons, which are required for future exploration of physiologically relevant anticipatory behaviors that are driven by biological clocks.

Methods

Animals

Nocturnal M. sexta hawkmoths were bred and raised from eggs in the rearing facility at the University of Kassel. Males and females were separated during pupal stages. Males were housed isolated from the females to avoid exposure to female pheromones. All experiments were performed with adult males reared in a long-day photoperiod (L:D 17:7 h to prevent diapause) at 25 °C with relative humidity of about 55%. Larvae were fed with an artificial diet modified after Bell and Joachim (1976); adult moths could feed on sugar solution with added vitamins (Roth) ad libitum (Riffell et al., 2008).

While all animals were exposed to the same lighting regime as zeitgeber to entrain them up until the start of the experiments, we also attempted pheromone exposure as additional circadian zeitgeber to phase-align spontaneous ORN activity in a subset of males (Ghosh et al., 2024). Without allowing direct contact, virgin males were exposed at zeitgeber time (ZT) 16 to females and naturally occurring pheromone blends in closed mating cages for 30 mins. Males were then moved back into isolation before tip recordings (see below) were started at ZT 17 on the same day. Since we focused on spontaneous spiking activity rather than on pheromone-dependent responses and because repetitive or long pheromone exposure changes the composition of 2nd messenger modulated ion channels in ORNs (Gawalek and Stengl, 2018), we did not employ any further pheromone synchronization protocols.

Solutions

Ringer compositions were taken from (Pézier et al., 2007) and contained (in mM): 6.4 KCl, 12 MgCl2, 1 CaCl2, 12 NaCl, 10 HEPES, 340 glucose for hemolymph ringer (HLR) at 450 mosmol/kg, and 172 KCl, 3 MgCl2, 1 CaCl2, 25 NaCl, 10 HEPES, 22.5 glucose for sensillum lymph ringer (SLR) at 475 mosmol/kg, both at pH 6.5.

The Orco antagonist OLC15 (N-(4-butylphenyl)-2-((4-ethyl-5-(2-pyridinyl)-4H-1,2,4-triazol-3-yl)thio)acetamide (Chen and Luetje, 2012); kindly provided by Dr. D. Wicher, Max Planck Institute for Chemical Ecology, Jena, Germany) was dissolved in DMSO to a concentration of 100 mM (stock solution stored at 4-8 °C) before being diluted in SLR to a final concentration of 50 µM to be used within one week (stored at 4 °C). We determined in previous studies that, compared to other Orco antagonists, OLC15 is the most specific (Chen and Luetje, 2012; Nolte et al., 2016, 2013; Pask et al., 2013). It is effective at 50 µM, according to dose-response curves, and dose-dependently antagonizes the action of the Orco agonist VUAA1 in intact hawkmoths (Nolte et al., 2016, 2013). DMSO percentage was 0.05% in the final solution.

The cAMP analog 8-bromo-cAMP acetoxymethyl ester (8-Br-cAMP-AM; Biolog Life Science Institute GmbH & Co. KG) was dissolved in DMSO to a concentration of 10 mM (stock solution stored at -20°C) and diluted to a final concentration of 100 mM in either SLR or SLR + 50 µM OLC15.

Electrophysiology

The spontaneous electrical activity of the unstimulated pheromone-sensitive ORNs in a single long trichoid sensillum was recorded extracellularly as in vivo tip recording (Kaissling et al., 1987). 1-2-d-old male M. sexta that were raised and entrained to zeitgebers as described above were fixed in a custom-made Teflon holder with adhesive tape. The right antenna used for recordings was oriented with its dorsal surface facing up and immobilized near the base with dental wax (Boxing wax strips, KerrHawe SA). Electrodes (Ag/AgCl wire in a glass pipette filled with SLR for the recording electrode and HLR for the reference electrode) were pulled (DMZ-Universal Puller, Zeitz Instruments) from borosilicate capillaries (OD:1.56mm, ID: 1.17mm, without filament; Science Products) to a tip diameter of about 2 μm. All recordings were made at a distance of 2/3 to 3/4 of the antenna length towards its distal end because pheromone-sensitive long-trichoid sensilla (sensilla trichodea type 1) occur in a characteristic, homogenous pattern along the hawkmoth antenna (Kaissling et al., 1989; Kalinová et al., 2001; Lee and Strausfeld, 1990). Furthermore, all ORNs in long trichoid sensilla express the same Orco staining pattern (Nolte et al., 2016), share largely overlapping ion channel populations, and obvious differences in the activity of ORNs from different annuli have not been reported (c.f., Dolzer et al., 2003; Gawalek and Stengl, 2018; Schneider et al., 2025). Thus, it is very likely that pheromone encoding mechanisms are not locally restricted along the flagellum. Recording and reference electrodes were located on the same annulus: After cutting off the distal 20 annuli of the antenna with micro scissors, the reference electrode was inserted to a depth of 2 annuli into the antennal lumen from the cut end. The cut was covered with electrode gel (GE Medical Systems Information Technology) to prevent desiccation. Under microscopic control, some tips of the long trichoid sensilla of the annulus were cut off with sharpened Dumont forceps, and the recording electrode was slipped over a single sensillum of an upper sensillar row. Electrodes were moved with manual micromanipulators (MMJ, Märzhäuser Wetzlar; and Leitz, Leica Microsystems) or with motorized micromanipulators (Microstar, Scientifica) with a joystick controller (Scientifica).

ORN activity was amplified 200-fold (custom built amplifier with 1012 Ω input impedance, or ELC-01 MX in combination with DPA-2FX, npi, or EXT 10-2F, npi), low pass filtered at 1.3 kHz, digitized at 20 kHz with a Digidata (1550A or 1550B, Molecular Devices), and saved for offline analysis with Clampex 10 (pCLAMP suite, Molecular Devices).

All tip recording experiments were performed at room temperature (21-27 °C) and a relative humidity of 35-60%. These environmental fluctuations across experiments were based on seasonal influences. Throughout each experiment, conditions were stable. To ensure that neither electrode reattachment to the same trichoid sensillum nor application of DMSO influenced spiking activity, we performed two sets of paired experiments at ZT 1-3 in which we recorded the spontaneous activity before and after removing and reattaching the SLR-filled recording electrode to the same sensillum, or recorded the spontaneous activity of the same sensillum in SLR and SLR + 0.15% DMSO. Both treatments did not affect spontaneous spiking activity (Figure 1B).

Attributes of pheromone-sensitive ORN spiking patterns.

(A) Spikes were identified by their prominence in the raw ORN recording. The inter-spike interval (ISI) was defined as the time from one spike to the next. A train with ≥2 spikes with ISI ≤ 50 ms was considered a burst. The intra-burst interval was the ISIs within a burst, the inter-burst interval (IBI) the ISIs between two bursts, omitting any individual spikes in between. The inter-event interval (IEI) included the ISIs between bursts and individual spikes. Detaching and reattaching the recording electrode to a long trichoid sensillum (Bi; paired t-test t(9) = -2.184; p = 0.057; N = 10) or adding DMSO to the SLR (Bii; Wilcoxon signed-rank test Z = 0.153; p = 0.922; N = 10) did not affect the mean spiking frequency.

We recorded ORN activity either in long-day conditions (17:7 LD) or constant darkness (DD) to examine the free-running peripheral ORN clock. For LD experiments, the light regime continued as in the rearing facility during the experiment. For DD and OLC15 experiments, the lights were turned off at the beginning of the recording, i.e., all animals were exposed to the same light zeitgeber until the beginning of the experiment. To examine the effect of Orco on the spontaneous action potential (AP) activity pattern, we used the recording electrode to infuse OLC15 in DD conditions. To examine the effect of cAMP on Orco gating, we recorded the spontaneous ORN activity at the end of the moth’s activity phase (ZT 1-3) in SLR as control, and either 8-Br-cAMP-AM, or 8-Br-cAMP-AM + OLC15, infused through the recording electrode.

ORNs generate spontaneous APs. However, high spiking activity during control recordings is indicative of stress, whereas low/no activity could be indicative of damage. Therefore, we only considered experiments in which the average spiking frequency was between 2 Hz and 0.008 Hz in control conditions.

Data analysis

To characterize the spontaneous ORN spiking activity, we first detected spikes with a self-written program in Python. Using the scipy.signal package and find_peaks function, we used a moving mean with 100 samples (5 ms) window size for baseline correction, followed by applying a 8th order low-pass Bessel filter (2000 Hz cutoff frequency) to the raw voltage trace to facilitate spike detection within the following conditions: spike full width (3-15 ms), threshold search for amplitude of spike (0.5-2.5 mV) and threshold prominence (0.1). Two or more consecutive spikes with intervals ≤ 50 ms were considered a burst (Dolzer et al., 2001). We then calculated the following attributes in bins of either 10 min or 1 h for further analysis: instantaneous spike frequency (ISI, 1 / inter-spike interval), inter-burst interval (IBI, time between two consecutive bursts with no single spikes in between), inter-event interval (IEI, time between two events, i.e. two consecutive single spikes, two consecutive bursts, a burst and a consecutive spike, and vice versa), burst duration (time from first to last spike in a burst), percentage of all spikes in the experiment that are members of burst, and mean number of spikes per burst (Figure 1A). All statistics were done with those binned values.

Hawkmoth behavioral and physiological rhythms desynchronize across the population in the absence of zeitgebers like circadian rhythms of pheromone stimulation and daily light-dark cycles. Therefore, we first used RAIN analysis (Thaben and Westermark, 2014) to identify circadian rhythms (24 ± 4 h) in the attributes above for each animal and condition (LD, DD, OLC15) in the 1 h-binned data. Data was downsampled to 2 h resolution and thus included every other 1 h bin to speed up computational time. To further quantify the circadian changes in attributes across animals, it was then necessary to phase align the recordings obtained in DD and OLC15. To achieve this, we used an optimization procedure to calculate the optimal temporal lags for each dataset, aiming to maximize the overall cross-correlation between the calculated binned spike frequency series. The optimization was performed using simulated annealing, which provided a set of time shifts (lags) for each series. These lags were applied to shift the frequency data in a manner that maximized the synchronization between the different subjects’ data, resulting in a “subjective ZT” with subjective ZT 0 at the first maximum of the binned ISI for each individual. The objective function used for optimization was the negative sum of pairwise correlations between all series after applying the shifts. The resulting alignment allowed us to compare the frequency patterns across the different conditions.

Once the data were aligned, we computed the mean frequency across all subjects for each bin and each condition. Then, the first high activity period was identified by finding the maximum in the mean frequency series. Subsequently, we identified the second high activity period as the time 24 hours after the first one and the first low activity period was found by locating the minimum in the mean frequency series between the two high activity periods. Besides the mean frequency, we also searched for significant oscillations in IBI, IEI, burst duration, percentage of spikes that are part of bursts, and mean number of spikes per burst. For this we found the maximum value of these statistics within a 3 h range centered in the first low activity period and compared it with the values at the second high activity period.

Continuous wavelet and Fourier analysis were performed using PyBoat (version 0.9.12) (Mönke et al., 2020) to examine the time series of neuronal firing frequency. The continuous wavelet method allows us to detect and characterize rhythmic patterns within the data by identifying frequency components that change over time. Specifically, we used wavelet analysis to determine whether a neuron exhibited circadian and ultradian rhythms. Unlike traditional Fourier analysis, which provides an overall frequency composition but assumes stationarity, wavelet analysis is particularly useful for biological signals because it can reveal how these frequencies evolve dynamically. By applying this approach, we could assess not only the presence of ultradian rhythms but also whether their strength or occurrence fluctuated in a circadian manner, indicating modulation by the 24-hour cycle. For this analysis, we used the 10 min-binned data to have sufficiently high resolution to detect ultradian rhythms. Periods were determined as local maxima in the Fourier power spectra.

Real-time quantitative polymerase chain reaction (qPCR)

We followed the qPCR methodology detailed in (Schneider et al., 2025). Briefly, male moths were isolated and cultured until the second day after eclosion, and the antennae from four individuals were collected every 4 hours for one day, starting at ZT 0, using liquid nitrogen, with four biological replicates per time point. Total RNA was extracted using TRIzol® Reagent (Invitrogen) and quantified for concentration and purity with the NanoDrop ND-1000 spectrophotometer (Thermo Fisher Scientific). First-strand complementary DNA (cDNA) was synthesized using the PrimeScript™ RT reagent Kit with gDNA Erase (Takara Bio Inc.). Primers for Orco (F: GCTCGCTACCACCAAATTGC, R:TCGTGACCCAACTGACAACA; Gene ID: LOC115452348) and for timeless (tim) (F: TTAAGCCGACCGTAGTGCTG, R: CGTCTTCCGTCCATGTGTCT; Gene ID: LOC115446009) were designed using the Primer-BLAST online program of NCBI. The qPCR was conducted following the protocol of TB Green® Premix Ex Taq™ II (Tli RNase H Plus) (Takara Bio Inc.), with the program: 95 °C for 30 s, followed by 40 cycles of 95 °C for 5 s, and 60 °C for 30 s. Relative expression levels of Orco were calculated using the 2-ΔΔCt method (Livak and Schmittgen, 2001).

Statistics

All statistics were done with self-written python codes, SigmaPlot 12 (Systat Software), or JASP (versions 0.19.03 and 0.95.4, University of Amsterdam) with a significance level α = 0.05. Unless noted otherwise, data were checked for normal distribution with Anderson-Darling or Shapiro-Wilk tests and for equal variance with Levene’s test. If one of these assumptions failed, we used non-parametric tests, otherwise we used parametric tests.

To assess the statistical significance of observed differences in neuronal activity of the aligned data between ZTs, we applied the Wilcoxon signed-rank test to pairwise comparisons between the first low activity period and the subsequent high activity period and discarded the transient high activity at the beginning of most recordings.

For paired recordings (electrode reattachment, addition of DMSO, or cAMP/cAMP+OLC15 in the same animal, we used either paired student-t test or Wilcoxon signed-rank test, depending on normal distribution of data.

For other data, we used an ANOVA on ranks with either Tukey’s or Dunn’s post-hoc test for multiple comparisons as indicated.

ORN Model

In this study, we introduce a single-compartment Hodgkin-Huxley (HH) type model of the spontaneous activity of M. sexta ORNs (available on GitHub (Forlino et al., 2025)). The model contains 5 sets of ion channels that were previously identified in electrophysiological recordings (Dolzer et al., 2021): the fast sodium (Na+) and potassium (K+) channels responsible for the generation of spikes, low voltage-activated calcium channels (LVA) responsible for the initiation of bursts, calcium-gated potassium channels (BK) responsible for the termination of bursts, and the Orco channel, which is a voltage-independent non-specific, cAMP-gated channel. A schematic of the model is provided in Figure 2A. We modeled Orco as a ZT-dependent conductance to represent the effect of the oscillations in cAMP concentration that were identified in both hawkmoth and cockroach antennae (Schendzielorz et al., 2014, 2015, 2012) (Figure 2B, C). For simplicity we chose to use a sine function to directly represent the oscillations in Orco conductivity, but similar results can be achieved by modeling it as an output of a Goodwin model of the circadian clock via cAMP concentration.

The conductance-based model of a simplified hawkmoth ORN.

(A) Schematic of the model containing a voltage gated sodium channel (Na+), a voltage gated potassium channel (K+), a low voltage activated calcium channel (LVA), a calcium and voltage activated potassium channel (BK), a leak channel, and a cAMP-gated leaky non-specific cation channel (Orco). The Orco channel is represented as a linear conductance whose value oscillates with a sinusoidal shape as a function of the ZT. The circadian rhythm in the Orco conductance is hypothesized to result from a cAMP-dependent increase in channel open-time probability (as found in Drosophila Orco: (Getahun et al., 2013)) via a circadian oscillation of antennal cAMP concentrations (Schendzielorz et al., 2015). This oscillation is driven by the circadian clock and serves as an input; the internal mechanism of the clock is not modeled explicitly. Arrows indicate the flow of ions through the channels. (B) Parametrized circadian oscillation in Orco conductivity, introduced as an external input to our model, representing the gating of Orco by cAMP whose concentration oscillates on a circadian timescale. The times of minimum and maximum cAMP concentration are indicated by colored arrows. (C) Parametrized IV curve of the Orco channel showing a linear behavior with the slope depending on the cAMP concentration. Line colors correspond to the arrows in B. (D) Markov-chain representing all possible states and transitions for each type of ion channel included in the model. The per capita transition rates (α and β) depend on membrane potential and in the case of the BK channel also on the intracellular Ca2+-concentration. The “N” state vectors contain the population of channels in each state “n”. Directed edges are numbered in red. The only conducting state of each ion channel, representing the condition where all its gates are open, is shown in green.

Each type of ion channel possesses one or more gates that can open and close individually with transition rates α and β for opening and closing, respectively, of activation gates (n, m) and inactivation gates (h). As a result, each possible combination of open and closed gates defines a possible state in which each ion channel can be found. A schematic of the Markov chain depicting all possible states and transitions for each ion channel is shown in Figure 2D. Each ion channel can conduct current only in the state where all its gates are open.

First, we constructed a deterministic 22-dimensional model to serve as the mean field of the Langevin model following the methodology proposed by Pu and Thomas (2020). Here, one dimension tracks the membrane potential, another the intracellular Ca2+ concentration, and the remaining 20 dimensions track the number of channels that are in a specific state of the Markov chains. Considering the conservation of the total number of ion channels, each type of ion channel required one less dimension than the total possible states.

The complete Langevin model adds to the mean field an independent noise source for each transition in the Markov chain. These noise sources are biophysically motivated since the transition between ion channel states is a stochastic process, with the probability of transition determined by the kinetics of each kind of ion channel. The variability in the spiking behavior of a neuron emerges from this non-deterministic gating of individual ion channels. The detailed mathematical description of the model can be found in the Supplementary Material.

Results

We tested the hypothesis that sensory neurons, such as insect ORNs, perform active, anticipatory sensing based upon an endogenous plasma membrane clock with the pacemaker channel Orco as the core element. We focused on pheromone-sensitive long trichoid sensilla of male M. sexta hawkmoth antennae which express daily rhythms in pheromone sensitivity and temporal resolution that are controlled by a circadian clock. We performed minimally invasive in vivo tip recordings over the course of several days to search for daily rhythms in spontaneous spiking activity as a measure of endogenous membrane potential oscillations.

Pheromone-sensitive hawkmoth ORNs exhibit Orco-dependent circadian rhythms in spontaneous spiking activity

The ORNs that innervate the pheromone-sensitive long trichoid sensilla on male M. sexta antennae display spontaneous spiking patterns of irregular bursts with interspersed single action potentials in short-term recordings (Dolzer et al., 2001). In the present study we documented the same pattern in long-term recordings over the course of 2-7 days. This enabled us to search for daily and circadian rhythmicity in spontaneous ORN spiking activity in the absence of pheromone stimulation.

RAIN analysis of the spontaneous spiking patterns revealed significant daily rhythms in 7 of 11 animals under long-day conditions (17:7 h LD) (Figure 3A). The maxima in spontaneous activity occurred mostly during the dark phase when nocturnal hawkmoths are active (Figure 3A (top), Bi light blue arrow), and minima occurred during the light phase when hawkmoths rest/sleep (Figure 3A (bottom), Bi dark blue arrow). High spiking activity in the dark phase had a mean maximum frequency of ∼3 Hz (delta range: 0.5 – 4 Hz) (Figure 3Bii). In contrast, the minimum spiking frequencies during the hawkmoth’s resting phase were below 0.5 Hz (Figure 3A (bottom)).

Spontaneous activity of pheromone-sensitive olfactory receptor neurons (ORNs) of male M. sexta shows Orco-dependent circadian modulation.

(A) Example tip recordings of the spontaneous activity of one long trichoid sensillum during a period of high (top, average frequency ∼ 3 Hz) and low (bottom, average frequency ∼ 0.2 Hz) activity. (Bi) Spontaneous spike frequency of one long-term ORN recording during 17:7 light-dark cycles (same experiment as in A, colored arrows indicate respective zeitgeber times (ZT)), indicated by the black-white bar at the top, increased during each activity phase and was low during the resting phases. Spike frequencies were averaged for each 1 h bin. (Bii) Mean spontaneous ORN spike frequency across all LD animals in 1 h bins. Data from the individual in Bi is highlighted in red. Recordings had different lengths, therefore the number of recordings used for the element-wise mean for each 1 h bin decreased with the time since the start of the recording, indicated in the top panel. (Biii) Mean spontaneous ORN spike frequency across all LD animals in 1 h bins during the first 48 hours revealed circadian activity. (Ci) The spontaneous spike frequency of one long-term ORN recording in constant darkness (DD, the expected times of lights on are represented by grey bars above) exhibited a circadian pattern. The peak activity shifted in DD due to the animal’s endogenous, free-running circadian period τ = 21.27 h. (Cii) Mean spontaneous ORN spike frequency across all DD animals in 1 h bins. Data from the individual in Ci is highlighted in red. (Ciii) Mean spontaneous ORN spike frequency across all DD animals in 1 h bins during the first 48 hours after aligning the time to the first maximum in spontaneous activity (see Methods) revealed circadian activity. (Di) The spontaneous spike frequency of one long-term ORN recording in constant darkness with infusion of the Orco antagonist OLC15 (orange frame) dissipated the circadian rhythm of spontaneous activity. (Dii) Mean spontaneous ORN spike frequency across all OLC15 animals in 1 h bins. Data from the individual in Di is highlighted in red. In contrast to LD and DD conditions, the spike frequency decreases over time with prolonged exposure to OLC15. (Diii) Mean spontaneous ORN spike frequency across all OLC15 animals in 1 h bins (mean ± SD) during the first 48 hours after aligning the time to the first maximum in spontaneous activity (see Methods). In contrast to LD and DD conditions, the circadian change in spontaneous spiking frequency disappeared. (E) The slopes of the linear fits to the binned spiking activity of each individual animal in the three different conditions (see Methods). Each dot indicates the slope for one experiment. The line indicates the mean. The slope for OLC15 is significantly different from control LD.

Since ORN firing patterns had a clear daily rhythm in LD both in individual animals (Figure 3Bi) and averaged across animals (Figure 3Bii, Biii), we examined whether these rhythms depended on cycling light cues or whether they persisted in constant darkness (DD), as expected for an endogenous, circadian clock-driven rhythm. In DD, the spontaneous ORN firing pattern remained rhythmic (RAIN: 8 of 12 animals), albeit with less robustness: The single peak of high spiking activity that was observed at the same ZT across several days in LD divided into multiple bouts of high spiking activity in the activity phase in 8 of 12 animals (Figure 3Ci). Also, in DD conditions, the high spiking activity phase shifted into the subjective day due to the individual endogenous circadian periods of about 23.5 +/-2.8 h (21.27 h in Figure 3Ci). Using pheromone exposure as additional zeitgeber also did not synchronize the animals (Figure 3 – Supplement). Thus, averaging the instantaneous spiking activity across all DD animals did not depict clear circadian rhythms (Figure 3Cii). Therefore, we phase-aligned the mean frequency of spontaneous activity (see Methods). After alignment, circadian rhythms were visible across DD animals (Figure 3Ciii), comparable to the daily rhythms in LD (cf. Figure 3 Biii and Ciii).

Next, we examined whether Orco, a leaky, non-specific cation channel, is the dominant depolarizing pacemaker current of ORNs that drives circadian rhythms in spontaneous spiking activity. Infusion of the Orco antagonist OLC15 into the sensillum lymph obliterated circadian rhythms and attenuated the spontaneous activity in several, but not all experiments (RAIN: 2 of 12 animals expressed circadian rhythms) (Figure 3Di). This attenuation resulted in a linear decrease in spiking activity over several days (Figure 3Dii, Diii). Even after phase-aligning the individual experiments, no circadian rhythm was present in the average across all OLC15-treated animals (Figure 3Diii).

Because OLC15 is not membrane-permeable on its own it was infused with low concentrations (0.05%) of DMSO through the recording electrode into the sensillum lymph and, therefore, effectiveness increased over time. Hence, we fit the spiking activity of each animal and in each condition with linear regression lines over the whole recording time to compare the slopes of the fits (Figure 3E). In LD, the mean ± SD slope was 0.004 ± 0.006 Hz/h, indicating that the average spiking activity remained similar throughout the duration of the recording. In DD, the slope was -0.003 ± 0.018 Hz/h and was not different from LD. In contrast, the mean slope during OLC15 treatment was -0.007 ± 0.022 Hz/h, significantly smaller than in LD (ANOVA on ranks with Dunn’s post-hoc test; H(2) = 7.681, p = 0.021). The negative slope corroborated the finding that spiking activity decreased over time in the presence of the Orco blocker.

Activity rhythms between hawkmoths desynchronize and have only weak phase coupling in the absence of pheromone stimulation, even in LD cycles (Ghosh et al., 2024). Therefore, we first determined the percentage of animals that expressed circadian rhythmicity (period of 24 ± 4 h) for each attribute with RAIN (Figure 4A). In LD, all attributes except #AP per burst and intra-burst frequency showed daily modulation in most animals. Circadian rhythms were still present in more than half of the animals in DD for mean spiking frequency, mean burst frequency, and mean event frequency. In contrast, the attributes of only a few animals remained rhythmic with the Orco channel blocker OLC15.

Blocking Orco removed circadian modulation of spontaneous ORN spiking patterns.

RAIN analyses demonstrated significant circadian rhythmicity for most animals in most attributes of the spontaneous spiking activity in LD and DD conditions but not in OLC15 (A). Values and background colors indicate the percentage of animals that expressed significant circadian rhythmicity. Circadian differences of attributes were further quantified between the time windows of low vs. high activity (Wilcoxon signed-rank test, α = 0.05; B-H). Significant differences occurred mostly in LD (white boxes) and DD (gray boxes) conditions, but never when Orco was blocked with OLC15 (orange boxes). In both LD and DD, the mean spiking frequency was increased during high activity (Bi, Bii). Only in LD but not in DD, the mean burst frequency (Ci, Cii) increased, whereas the relative number of spikes belonging to a burst (Di, Dii) decreased significantly during the high activity period. Both in LD and DD the mean event frequency (Fi,Fii) increased significantly. Mean burst duration (Ei, Eii), mean number of spikes per bust (Gi,Gii), and mean intra-burst intervals (Hi, Hii) did not differ between low vs. high spiking activity in LD and DD. When blocking Orco, the attributes for low vs. high activity did not differ for any of the attributes tested (Biii-Hiii). In DD and OLC15conditions, activity phases were aligned as described in the Methods; thus, low activity in DD was at subjective ZT 19 and in OLC15 at subjective ZT 13, high activity was in both cases at subjective ZT 24. ZTs for low and high activity in LD were 10 and 0, respectively.

To further quantify the circadian differences, especially of the desynchronized animals, we compared different attributes of spontaneous activity between periods of high and low activity, for DD conditions after phase alignment (see Methods) instead of comparing fixed ZT or circadian time (CT) intervals. Despite the high variability in ORN spiking between individuals, the mean frequency of spontaneous ORN activity was significantly higher during the high activity phase than during low activity in both LD and DD recordings (Figure 4Bi, Bii). In LD, but not in DD, the frequency of bursting increased significantly during the high activity phase (Figure 4Ci, Cii). The percentage of spikes that are part of bursts was less variable and lower during the high activity phase in LD but not significantly different in DD (Figure 4Di, Dii). The mean burst duration was less variable during the high than the low activity phase in LD, but not significantly shorter in any condition (Figure 4Ei, Eii). Furthermore, both in LD and DD, the mean inter-event frequencies were significantly higher during the high activity phase (Figure 4Fi,Fii), indicating a significantly higher overall spontaneous spiking activity during the hawkmoth’s activity phase. The mean number of spikes per burst, and intra-burst frequencies did not differ significantly between high or low activity periods in both LD and DD conditions (Figure 4Gi, Gii, Hi, Hii). The addition of Orco antagonist OLC15 deleted any significant differences of spontaneous activity found in LD and DD (Figure 4Biii-Hiii). In conclusion, the mean frequency of spontaneous spiking and the frequency of bursting expressed circadian modulation, and are both most likely controlled via a circadian clock that involves Orco.

Orco imposes circadian modulation on the ultradian rhythms in spontaneous ORN activity

After establishing that Orco influences the circadian changes in ORN spiking activity between the animal’s resting and activity phase we examined if Orco also influences ORN spiking rhythms on the largely different ultradian timescales in the frequency range between 0.01-1000 Hz. In all conditions, two broad bands of ultradian instantaneous frequencies were evident: one upper band of >10 to ∼100 Hz (beta/gamma range) and a lower band between 0.1 to ≤10 Hz (delta/theta/alpha range) (Figure 5). The upper band primarily represents the frequencies of spikes within a burst, while the lower band represents single spikes and frequencies between bursts.

Blocking Orco removed circadian regulation of ultradian rhythms in the spontaneous ORN spiking pattern.

Heat map of the instantaneous frequencies (1/ISI) over two consecutive days of long-term tip recordings of one pheromone-sensitive long trichoid sensillum under different conditions (LD (A), DD (B), OLC15 (C)) in each panel. Representative recordings from one animal in each panel. Pixel color indicates the counts of instantaneous frequencies in that respective bin. The points fall mostly in a band of high frequencies (>10 – ∼100 Hz) and low frequencies (0.01 – ∼10 Hz), with the high-frequency band representing the instantaneous frequencies of spikes within a burst and the low-frequency band the instantaneous frequency between bursts. (A, B) The high frequency band indicates daily and circadian modulation of frequency prevalence. In addition, the low frequency band (<0.01 Hz to ∼ 10 Hz) also displayed circadian modulation of the frequency composition. (C) Infusion of OLC15 deleted the circadian modulation in both frequency bands but not the ultradian rhythm of frequency prevalence.

In LD, two types of daily/circadian modulation of the ultradian frequencies occurred (Figure 5A). The upper frequency band was mostly modulated in prevalence (its frequency of occurrence) but not in its frequency range. The lower band was modulated both in prevalence and frequency range, which resulted in a sinusoidal change in the mean of the frequency band over 24 h. During the hawkmoth’s activity phase, higher frequencies of up to ∼10 Hz occurred, while frequencies dropped below 1 Hz during the resting phase. Because of this circadian increase in the lower frequency band, it merged with the high frequency band when both bands had maximum prevalence during the activity phase around ZT 0.

The circadian modulation of both frequency bands was maintained in constant darkness (Figure 5B) but phase-shifted with respect to the projected ZT due to the endogenous circadian period of each individual. Again, this indicates that a clock is modulating the spontaneous ORN spiking activity over the course of a day. The addition of the Orco antagonist OLC15 deleted any circadian modulation of either frequency band. Interestingly, the instantaneous frequencies in both bands still showed correlated changes in prevalence and merging, but now at ultradian periods only.

Additional Fourier analysis (see Methods) of spontaneous spiking activity over the course of several days revealed faster ultradian and slower infradian rhythms, ranging from 0.6 h to 38 h, in addition to the circadian (24 ± 4 h) rhythm (Figure 6Ai, Bi, Ci). Some rhythms (factorials of 24 h, like 12 h, 6 h, etc.) appeared to be harmonics of the circadian rhythm, which were not excluded in our analysis. Most animals expressed ultradian rhythms of 2-4 h in LD as well as DD conditions. Fourier confirmed the decrease of animals with circadian spiking activity in OLC15 (58.3% in DD, 33.3% in OLC15). Please note that discrepancies in the number of animals with circadian rhythms between RAIN and Fourier/wavelet analyses are due to the fundamentally different sensitivities in methodology. In brief, Fourier analysis assumes a stationary sinusoidal component, whereas RAIN as a rank-based umbrella test detects consistent raise-fall patterns, including asymmetric rhythms. Nonetheless, both methods confirmed a dramatic decrease in animals with circadian spiking patterns in OLC15. In addition, 33.3% of animals expressed infradian spiking patterns (up from 0% in DD) with OLC15. Ultradian periods could be detected in all animals in all conditions but their distribution did not differ between conditions (1-way ANOVA;F(2, 117) = 0.87, p = 0.421). These results were corroborated by continuous wavelet analysis, which provided time-resolved spectral power across different rhythmic periods (Figure 6Aii, Bii, Cii).In control LD and DD conditions, wavelet power for ultradian periods increased during times of elevated firing rate and decreased during periods of low activity (Figure 6Aii and Bii, middle and bottom panels). This dynamic pattern indicates that ultradian rhythms were not constant over time but were instead modulated with a circadian rhythm, becoming more prominent during circadian peaks in neuronal activity. Such circadian gating of ultradian power was clearly visible in the wavelet spectrograms as alternating bands of higher and lower ultradian power aligned with daily activity cycles. Following Orco blockade with OLC15, this modulation was disrupted (Figure 6Cii middle panel): although ultradian components persisted (Figure 6Cii bottom panel), they no longer showed systematic variation with the circadian cycle, suggesting a decoupling of ultradian rhythms from circadian control.

Blocking Orco affected ultradian and infradian frequencies in the spontaneous ORN firing patterns.

The Fourier analysis of spontaneous spiking activity revealed rhythms with ultradian (<20 h; dark grey), circadian (20-28 h; white), and infradian periods (>28 h; light grey) in LD (Ai), DD (Bi), and OLC15 (Ci). Each line represents one animal where each triangular marker is at the local maximum in the frequency spectra obtained for that animal. The histograms above illustrate how often specific periods of spiking activity rhythms occurred averaged over all animals. Circadian rhythms were detected in LD (N = 5 of 11), DD (N = 7 of 10), and OLC15 (N = 4 of 12). Wavelet analysis of LD (Aii), DD (Bii), and OLC15 (Cii) confirmed the occurrence of multiscale periods. Example plots for one animal each, the same animals as in Figure 5. For each panel, the top plot depicts the mean firing frequency, the middle plot the wavelet power in the same period range as panels i, and the bottom plot the wavelet power for infradian periods up to 7 h on a log scale to highlight infradian time scales. Gray dashed lines indicate cone of uncertainty.

Orco expression is not under the control of the molecular TTFL clockwork

So far, our results indicated a prominent role of Orco in the circadian modulation of ORN spiking activity. Since it is the general opinion that the master circadian clock is the TTFL clock that dominates all physiological and behavioral circadian rhythms via transcriptional control, we examined whether Orco transcript abundance shows a daily rhythm, as previously found for clock genes in hawkmoth antennae (Schneider et al., 2025; Schuckel et al., 2007). However, in contrast to the circadian clock gene tim, qPCR of whole male hawkmoth antennae across different ZTs revealed no significant daily rhythm of Orco transcript abundance (Figure 7; ANOVA on ranks, H(5) = 6.91, p = 0.227). Thus, we suggest that Orco function is controlled via a posttranslational circadian clockwork mechanism that is associated directly with the plasma membrane (Stengl and Schneider, 2024; Stengl and Schröder, 2021).

Orco is not under control of the TTFL-based molecular circadian clock.

Relative expression levels of Orco and tim mRNA from male hawkmoth antennae raised under LD conditions. Orco expression levels did not differ significantly between ZTs (ANOVA on ranks, H(5) = 6.91, p = 0.227: N = 4 per ZT). tim expression levels changed significantly throughout the day (1-way ANOVA, F(5, 18) = 6.215, p = 0.002) Red lines indicate mean, tim served as positive control.

Circadian changes in Orco conductance are sufficient to modulate spiking activity in a computational model of a hawkmoth ORN

The maximum conductivity of Drosophila Orco as a leaky ion channel is modulated by cAMP (Getahun et al., 2013; Martín et al., 2001; Sargsyan et al., 2011; Wicher et al., 2008). In hawkmoth antennae, cAMP levels oscillate throughout the day and are involved in sensitizing the ORNs to pheromone (Dolzer et al., 2021; Flecke et al., 2010; Flecke and Stengl, 2009; Schendzielorz et al., 2014, 2015). Hence, cAMP modulation of Orco conductance could be a putative mechanism for the Orco-dependent circadian modulation of ORN spiking activity, integrating Orco in a circadian PTFL based membrane clockwork (Stengl and Schneider, 2024).

To test the hypothesis that Orco is a cAMP-dependent, leaky ion channel, we constructed an ORN model (Methods and Supplementary Material) and tuned the parameters to provide a quantitative match with the biological data. The developed model provides a unified description of experimental observations. For instance, the ORN model yields irregular spiking activity with a bimodal distribution, consisting of both bursts and isolated spikes, successfully reproducing the experimental recordings (Figure 8A). This suggests a unified mechanism underlying the initiation of both phenomena: If the pacemaker-driven depolarization reaches the Na+ threshold a firing process begins, and at this point randomness determines if other slower ion channels responsible for bursting are sufficiently activated to sustain a burst of several spikes.

The model adequately reproduces the firing pattern of the biological ORN.

(A) The model activity (bottom trace) displays both isolated spikes and bursts of variable length as in the original recording (top trace). The model represents an intracellular recording whereas the biological recordings were done with extracellular electrodes. (B) Semi-logarithmic plot of burst length distribution of an experimental LD recording (top) and simulated model (bottom) of spontaneous action potential activity of pheromone-sensitive neurons. Both the model and the recordings display exponential decrease of the count’s density as a function of the burst length, shown as a linear decrease in the logarithmic scale. (C) Comparison of the circadian modulationof model and ORN spiking attributes. Top panels (Ci, Cii) show mean spike frequency, whereas bottom panels (Ciii, Civ) show the percentage of spikes belonging to bursts. The simulated model reproduced the circadian dynamics observed in the experimental recordings, demonstrating good agreement between model and biological data. (D) Heat map of the simulated model, pixel color indicates the density of instantaneous frequencies in that respective bin. The points can be roughly divided in a band of high frequencies (top, ∼50 – 80 Hz) and low frequencies (bottom, ∼0.1 – 5 Hz), with the high-frequency band representing the instantaneous frequencies of spikes within a burst and the low-frequency band the instantaneous frequency between bursts. (Di) Simulated spikes with circadian regulation by the Orco channel; the inter-burst frequencies varied widely with circadian rhythmicity while the frequencies within burst remained approximately constant. The simulated results also reproduced the “merging” effect between the two bands when the bottom one approaches the top one. (Dii) Simulated spikes without circadian regulation of Orco had no circadian oscillations.

Additionally, the ORN model replicates the distribution of the number of spikes per burst, which follows a decreasing exponential trend in both experimental and simulated results (Figure 8B).The model also successfully reproduces the circadian modulation of spiking activity: the maximum firing rate occurs when the Orco conductance is maximal, whereas the maximum percentage of spikes that are part of bursts correlates with the lowest Orco conductance, resulting in antiphase oscillations, as were observed in the experimental recordings (Figure 8C).

The frequency prevalences in the model over several days clustered in two distinct frequency bands: an upper band that corresponds to the intra-burst frequencies and a lower band, corresponding to inter-burst frequencies (Figure 8D), in agreement with the biological experiments (Figure 5A, B). When circadian oscillations in Orco conductance were included, inter-burst frequencies showed significant circadian variations, whereas intra-burst frequencies remained relatively constant, reminiscent of the biological data. Notably, the simulation captured the experimentally observed “merging” effect, where the low frequency band approaches the top band during high-activity periods (Figure 8D).

This merging of spiking and bursting frequencies suggests that the system approached a bifurcation, transitioning from a bursting regime to tonic spiking activity. Further evidence for the near bifurcation condition of the neurons is the distribution of number of spikes per burst that followed a decreasing exponential trend in contrast to a constant value, or a bounded range of spikes per burst, which would be expected from a standard bursting neuron model (Figure 8B). The observation that the circadian modulation of spiking activity requires only one parameter, namely changes in the Orco conductivity, to describe the circadian effects on the ORN spiking pattern indicates the crucial role of the Orco channel in circadian control.

cAMP regulates the circadian function of Orco

Our computational model predicted that the daily/circadian oscillations in cAMP levels (Schendzielorz et al., 2015) control the circadian modulation of spontaneous spiking activity of hawkmoth ORNs. To test this prediction in hawkmoth ORNs, we increased cAMP levels by infusing 8-Br-cAMP-AM through the recording electrode into the sensillar lymph (Figure 9) at ZT 1-3. This significantly increased the mean frequency of the spontaneous spiking activity (Figure 9A; Wilcoxon signed-rank test Z = -2.701; p = 0.004). To check whether this increase in mean frequency was caused by direct or indirect interaction of cAMP with the Orco channel, we infused the Orco antagonist OLC15 together with 8-Br-cAMP-AM through the same recording electrode. Blocking Orco prevented a significant change in the spontaneous spiking frequency via the cAMP analogue (Figure 9B; paired t-test t(9) = -1.446; p = 0.182). This further corroborates cAMP-dependent modulation of the circadian pacemaker channel Orco.

cAMP modulates open-time probability of Orco in M. sexta.

Perfusion of 8-Br-cAMP-AM significantly increased the spontaneous spiking activity at the end of the activity phase (ZT 1-3; A; Wilcoxon signed-rank test Z = -2.701; p = 0.004). However, activity did not increase when 8-Br-cAMP-AM was added together with the Orco blocker OLC15 (B; paired t-test t(9) = -1.446; p = 0.182). Electrodes were filled with SLR containing 0.1% and 0.15% DMSO as control in A and B, respectively. Data from individual animals are connected by lines.

Discussion

With in vivo long-term tip recordings of pheromone-sensitive long trichoid sensilla of male M. sexta hawkmoth antennae, we searched for a membrane-associated PTFL clockwork component that orchestrates membrane potential oscillations to facilitate active sensing of pheromone patterns during mating flights. Indeed, we detected interlinked circadian and ultradian frequencies in the spontaneous spiking activity, indicative of underlying membrane potential oscillations in ORNs. We found that the olfactory receptor co-receptor (Orco) acts as a pacemaker channel that controls the circadian modulation of the spiking activity without affecting the ultradian frequency ranges that are relevant for the primary fast, phasic pheromone response (Dolzer et al., 2003; Nolte et al., 2016; Schneider et al., 2025). Our computational model confirmed that the circadian modulation of open-time probability of Orco as a leaky pacemaker channel was sufficient to account for our experimental results. Since Orco was not under circadian transcriptional control of the TTFL present in hawkmoth ORNs, our findings are consistent with our novel hypothesis that the ORN’s plasma membrane constitutes an endogenous, plastic post-translational feedback loop (PTFL) clockwork. Consistent with experimental data (Schendzielorz et al., 2015), we propose that PTFL-generated endogenous multiscale membrane potential oscillations comprising interlinked oscillations in Ca2+ and cAMP levels provide circadian modulation of Orco conductance rather than TTFL-commanded circadian Orco transcription and turnover.

The expression of a 2nd messenger-dependent pacemaker/leak channel in ORNs that does not require daily transcription would constitute an economic and rapidly adjustable mechanism of active sensing. It could generate endogenous membrane potential oscillations that are tuned to detect and resonate with behaviorally relevant temporal patterns of zeitgeber signals, such as the circadian and ultradian pheromone fluctuations in the insect’s environment (Stengl and Schneider, 2024; Stengl and Schröder, 2021). The present study provides the first evidence for a systems view of biological timing that involves a PTFL membrane clock in insect sensory neurons, instead of the prevailing hierarchical view that assumes all endogenous circadian oscillations are outputs of the genetic TTFL clock (Stengl and Schneider, 2024; Stengl and Schröder, 2021).

Orco acts as a pacemaker channel that controls membrane potential oscillations underlying spontaneous ORN spiking activity

The resting membrane potential of most silent neurons lies close to the negative equilibrium potential for potassium ions, mainly due to potassium leak channels (Goldstein et al., 2001; Hille, 2001; Patel and Honoré, 2001; Talley et al., 2001). Spontaneously active neurons require the expression of leak or pacemaker channels that open at the hyperpolarized resting membrane potential, thus depolarizing the neuron(Bose et al., 2014; Cochet-Bissuel et al., 2014; Das et al., 2016; Golowasch et al., 2017; Lüthi and McCormick, 1998; Ratliff et al., 2021; Robinson and Siegelbaum, 2003; Sharma et al., 2023). When the pacemaker-driven depolarization reaches spike thresholds, voltage-gated Na+, K+ and Ca2+ channels promote Ca2+ influx and spiking. Voltage- and Ca2+-gated hyperpolarizing K+ channels then repolarize the neuron to negative membrane potentials, re-starting the cycle. Thus, membrane potential oscillations with interlinked oscillations in intracellular Ca2+ levels drive spontaneous spiking activity which depends on the type of pacemaker channel and antagonistic ion channels present.

In insect ORNs, at least two types of pacemaker channels exist: the highly conserved, slow, leaky Orco, which we focus on in this study, and an inversely voltage-dependent HCN type channel (Butterwick et al., 2018; Dolzer et al., 2021; Kodirov, 2022; Lee and MacKinnon, 2017; Mandala et al., 2025; Nolte et al., 2016, 2013; Sato et al., 2008; Stengl and Funk, 2013; Stengl and Schneider, 2024; Stengl and Schröder, 2021; Wicher et al., 2008; Wicher and Miazzi, 2021). When expressed in vitro in heterologous expression systems, Drosophila as well as Manduca Orco homo- and heteromers assemble into leaky, non-specific cation channels with a reversal potential around 0 mV, constitutively leaking Na+, K+, and Ca2+ into ORNs (Nolte et al., 2016, 2013; Sato et al., 2008; Stengl and Funk, 2013; Wicher et al., 2008; Wicher and Miazzi, 2021). Thus, while this study did not aim to unravel the overall generation of spontaneous activity in ORNs, here, we examined whether the leak channel Orco in hawkmoth ORNs is the core pacemaker channel responsible for membrane potential oscillations that drive spontaneous circadian activity (Stengl, 2010; Stengl and Schneider, 2024).

Differential location and function of Orco homo-or heteromers in the distinct membrane compartments of ORNs that face different ionic environments appear likely (Stengl and Funk, 2013; Stengl and Schröder, 2021). We propose that homomeric Manduca Orco pacemaker channels, as core element of a PTFL membrane clock, are restricted to the soma/inner dendrite compartment of ORNs (Stengl and Schneider, 2024; Stengl and Schröder, 2021). In contrast, we expect Manduca Orco heteromers to serve as obligatory chaperons to maintain ORs in the ciliary membranes, as proven in Drosophila (Benton et al., 2006). Since the ciliary compartment faces very high concentrations of K+ in the sensillum lymph, matching intracellular K+ levels (Kaissling and Thorson, 1980; Steinbrecht, 1992), we propose that Orco in ciliary membranes remains closed via unknown mechanisms, while Ca2+ channels of the TRP ion channel family serve the primary events of pheromone transduction (Gawalek and Stengl, 2018; Stengl, 1994, 1993). Otherwise, Orco channel opening would clamp the ciliary membranes at 0 mV, strongly reducing the driving force of pheromone-dependent receptor currents, not allowing for the ORN’s extreme pheromone sensitivity (Stengl and Funk, 2013).

In support of this hypothesis, in both Drosophila (Benton et al., 2006) and Manduca (Nolte et al., 2016), Orco antisera labeled Orco in two of the three sealed off ORN membrane compartments: in the membranes of soma/ inner dendrite compartment, as well as in the membrane of outer cilia, but not in the axons of ORNs. In contrast, antisera against an OR in Drosophila detected ORs only in ciliary membranes but not in the other compartments of ORNs. Thus, the colocalization of OR-Orco heteromers only in cilia confirm that Orco is an obligatory coreceptor and chaperone of ORs localizing and maintaining the OR-Orco complexes in the ciliary membrane (Benton et al., 2006). Since antisera against Manduca ORs are not available, we do not known whether the pheromone-sensitive OR-Orco heteromers are solely located to ciliary membranes also in the hawkmoth while Orco homomers are found in the soma/inner dendrite compartment (Fandino et al., 2019). However, we know that Orco exists as homomeric complexes in wild-type hawkmoth ORNs because Orco expression starts during development well before ORs are present and because Orco agonist VUAA1 activates Orco in primary cell cultures of not yet pheromone-sensitive maturing ORNs (Nolte et al., 2016; Schweitzer et al., 1976).

In summary, we propose Orco to be a 2nd messenger cascade-dependent core element of a PTFL membrane clock that is located in the soma/inner dendrite compartment of ORNs, which controls the ORN’s membrane potential and is coupled to - but not dominated by - the TTFL clock in the hawkmoth ORNs (Schuckel et al., 2007; Stengl and Schneider, 2024).

Orco is suggested to be a second messenger-dependent control hub of a PTFL membrane clock that regulates circadian membrane potential oscillations in hawkmoth ORNs

Different 2nd messenger cascades regulate Drosophila Orco via Ca2+/calmodulin and via phosphorylation/de-phosphorylation at multiple sites (Cao et al., 2016; Getahun et al., 2016; Guo and Smith, 2017; Mukunda et al., 2016, 2014; Sargsyan et al., 2011). The Ca2+/calmodulin binding site controls the localization of the OR-Orco complex in the ciliary membrane of ORNs, provides for negative feedback control of Orco conductance, and depends on previous odor stimulation (Bahk and Jones, 2016; Cao et al., 2016; Guo and Smith, 2022; Jain et al., 2021). After phosphorylation of Orco by protein kinase C (PKC), Orco directly or indirectly becomes cyclic nucleotide-dependent, with cGMP and cAMP increasing its open-time probability (Getahun et al., 2013; Wicher et al., 2008). While we do not know the second messenger-dependent modulation of hawkmoth Orco, due to Orco’s highly conserved structure we assume similar Ca2+/calmodulin-, PKC-, and cyclic nucleotide-dependent control of Orco’s localization and open-time probability.

Since blocking Orco in M. sexta decreased the spontaneous activity and deleted its circadian but not ultradian modulation, Orco is apparently the main target for modulation of circadian activity in hawkmoth ORNs. The ultradian rhythms in spontaneous activity are presumably controlled by additional pacemaker channels. Since Orco transcript abundance was constant throughout the day, Orco is not controlled directly by the circadian TTFL clock. Rather, a self-assembling PTFL membrane clock controls open-time probability and/or membrane localization of Orco via posttranslational modifications in a circadian manner. Because circadian rhythms in cAMP levels and phase-leading circadian rhythms in IP3 were found in hawkmoth antennae (Schendzielorz et al., 2015), it is likely that Orco’s phosphorylation state regulates its cAMP dependence in a daily rhythm. Indeed, our results showed that the increase of spontaneous activity caused by 8-Br-cAMP application is canceled with co-application of the Orco blocker OLC15. Thus, we assume that a PTFL membrane clock determines the circadian oscillations in the pheromone sensitivity and temporal resolution of ORNs via Ca2+, PKC, and cAMP-dependent membrane potential control (Dolzer et al., 2021, 2008; Flecke et al., 2010, 2006; Redkozubov, 2000; Schendzielorz et al., 2015, 2012; Ziegelberger et al., 1990).

Our hypothesis that a circadian PTFL membrane clock in hawkmoth ORNs controls circadian rhythms in spontaneous spiking activity predicts that this PTFL clock comprises a membrane signalosome with the pacemaker channel Orco as central control hub (Stengl and Schneider, 2024). In the soma/inner dendrite compartment of the ORNs the protein complexes of this PTFL signalosome would generate the specific ultradian and circadian frequencies of endogenous potential oscillations, linked to oscillations of Ca2+ and cAMP that determine physiological setpoints of the ORN during circadian sleep-wake cycles. Thus, the PTFL membrane clock that generates endogenous membrane potential and Ca2+ level oscillations minimally consists of depolarizing, Ca2+ permeable Orco, voltage-gated Ca2+, and hyperpolarizing K+ channels. In the same PTFL soma clock, membrane-anchored adenylyl cyclases (ACs) and phosphodiesterases (PDE), that are antagonistically controlled by intracellular Ca2+ levels, are proposed to generate circadian cAMP oscillations (Halls and Cooper, 2011). Kinases, such as PKC and PKA, could couple potential oscillations with cAMP oscillations that target Orco, phosphatases, and PDEs via negative feedback. Very likely, this membrane PTFL clock is interlinked with metabolic clocks such as PTFL clocks that generate endogenous rhythms of redox elements or ATP availability (Edgar et al., 2012), as evident in, e.g., human red blood cells, which lack a nucleus (O’Neill and Reddy, 2011).

Furthermore, this PTFL-clock could contain octopamine receptors that couple to endogenous oscillations of intracellular Ca2+ and cAMP levels through activation of Gαo and Gαs, as found in hawkmoths (Dacks et al., 2006; Flecke et al., 2010). Since octopamine levels in the hemolymph exhibit circadian oscillations and octopamine sets a physiological state of heightened alertness in insects, the circadian PTFL clock in ORNs would synchronize with endogenous physiological rhythms via octopamine receptor activation (Flecke and Stengl, 2009; Schendzielorz et al., 2015). This also constitutes a possible pathway to synchronize the ORN PTFL clock with circadian TTFL clocks in both ORNs and the brain. In horseshoe crab, the master clock sends multiscale patterned spike trains through retinal efferents at night, which carries information about the circadian phase and enhances visual sensitivity based on octopamine release and Camp signaling (Battelle, 2013; Liu and Passaglia, 2011). In M. sexta, two biogenic amine-immunoreactive neurons project into the sensory cell layer of the antenna, being suited to phase-shift endogenous cAMP oscillations via stress signals such as octopamine (Flecke and Stengl, 2009; Schendzielorz et al., 2015). Comparably, the circadian pacemaker center controls endogenous rhythms in the electrical network activity in rodent brains via endogenous cAMP oscillations that are phase-controlled through G-protein coupled receptor signaling (Ono et al., 2023; Prosser et al., 1989; Prosser and Gillette, 1991). Future studies will explore the links between PTFL and TTFL clocks in ORNs.

Circadian and ultradian rhythms in spontaneous ORN activity resemble the physiologically relevant timescales for active sensing

Moths are more sensitive to pheromone stimulation during their activity phase than at rest, apparently due to antennal circadian clocks (Flecke et al., 2010; Flecke and Stengl, 2009; Merlin et al., 2007, 2006; Schendzielorz et al., 2015; Schuckel et al., 2007; Silvegren et al., 2005). The greater the distance from the female, the lower the frequency of pheromone filaments in the air. The increased pheromone sensitivity during the activity phase enhances the male’s chance to detect and locate a calling female. Pheromone concentration is encoded in the frequency of the first, phasic burst of spikes of the characteristic triphasic spiking sequence of ORNs in response to brief pheromone pulses, which is independent of Orco (Dolzer et al., 2003; Nolte et al., 2013; Schneider et al., 2025). Consistent with these results, we found here that blocking Orco does not affect the upper frequency band of spontaneous activity that correlates with the frequency ranges of the pheromone-encoding primary fast pheromone response(Dolzer et al., 2003; Nolte et al., 2016, 2013). However, through the cAMP-dependent increase in Orco conductance during the activity phase, the ORN membrane potential is closer to spike threshold and fewer pheromone molecules are necessary to produce a receptor current that brings the membrane voltage over that threshold. Thus, larger Orco conductance increases the likelihood of an ORN response to pheromone stimulation without changing the typical triphasic spiking pattern of the response.

In the odor transduction of Drosophila, OR-Orco activation drives the initial receptor current response (Sato et al., 2008; Wicher et al., 2008). In contrast, Orco is not part of an OR-Orco odor-gated receptor ion channel complex in hawkmoths but controls the spontaneous ORN activity and the late, long-lasting pheromone response (LLPR). The LLPR begins several hundred milliseconds after the initiation of the pheromone-dependent transduction current and correlates with the time course of pheromone-induced rises in cGMP levels, possibly playing a role in odor memory or long-term gain control (Boekhoff et al., 1993; Nolte et al., 2016, 2013; Stengl, 2010; Stengl et al., 2001; Wicher and Miazzi, 2021; Ziegelberger et al., 1990). Thus, in hawkmoths, Orco is not part of the primary transduction channel but rather controls membrane potential oscillations and thereby the temporal resolution and general pheromone sensitivity of ORNs. In addition, the increased spontaneous burst frequency during the hawkmoth’s activity phase improves the resolution of the anticipated pheromone pulses during its mating flight in search for females.

In the mammalian suprachiasmatic nucleus (SCN), the clock that controls the sleep-wake cycle, cAMP levels oscillate at an circadian frequency, and cAMP and cGMP induce ZT-dependent opposing phase shifts in SCN spiking activity (Prosser et al., 1989; Prosser and Gillette, 1991). In antennae of both M. sexta (Schendzielorz et al., 2015) and the cockroach Rhyparobia maderae (Schendzielorz et al., 2012), cAMP levels express daily and circadian oscillations and peak during the activity phase. In contrast, cGMP levels do not oscillate under constant conditions. They are in antiphase to the daily oscillations in cAMP levels in LD and are elevated by strong or long pheromone exposure, which adapts the sensitivity and temporal resolution of ORNs (Boekhoff et al., 1993; Dolzer et al., 2021; Flecke et al., 2006; Schendzielorz et al., 2015; Ziegelberger et al., 1990). This implies that cAMP and cGMP may have ZT-dependent effects on Orco conductivity, thereby affecting pheromone sensitivity via control of the membrane potential and intracellular Ca2+ concentrations.

The upregulation of neural activity during wakefulness/alertness seems to be ubiquitous among different species. For example, the many synchronously recorded neurons in human EEG show fast oscillations in the alpha, beta, and gamma frequency ranges during alertness and slow oscillations in the delta and theta ranges during rest and sleep. Similarly, in the lower frequency band, the spontaneous spiking activity of hawkmoth ORNs changes from below delta during rest to theta/alpha ranges during their activity phase. The upper frequency band in the gamma range, which did not depend on Orco regulation, comprises the spike frequency within a burst, reminiscent of the first phasic pheromone response that encodes pheromone concentration changes (Dolzer et al., 2003). The increased spontaneous spiking activity in the theta/alpha range during the activity phase of the hawkmoth matches the maximal temporal resolution of pheromone pulses of up to 10 Hz (Marion-Poll and Tobin, 1992), which supports a role for Orco as part of a membrane-bound PTFL clockwork that tunes the ORNs for active sensing. Even through oscillations in the field potential of electroantennogram responses to odor stimulation can reach 100 Hz in different insects (Szyszka et al., 2014), these frequencies most likely do not encode the temporal resolution of physiological odor responses. Apparently, they reflect the astounding adaptive ability of antennal ORNs to synchronize their endogenous membrane potential oscillations across the antenna to a broad range of stimulation frequencies. Pheromone pulse resolution of about 100 Hz was never reported in single olfactory sensilla recordings, and in behavioral experiments male moths stopped their upwind mate search if pheromone pulses exceeded 30 Hz, apparently due to adaptation (Baker et al., 1985; Bau et al., 2005, 2002; Tripathy et al., 2010).

Circadian oscillations in the ORN membrane potential would be ideally suited to tune sensitivity and temporal resolution of the pheromone-sensitive ORNs to the hawkmoth’s rest-activity cycle, synchronizing male and female mating behavior. Here, we have demonstrated how modulation of a leaky pacemaker channel can achieve these circadian oscillations. Whether cAMP levels are under control of the TTFL clock or oscillate based on Ca2+ concentration oscillations, and would therefore be part of a stand-alone PTFL clock, remains to be studied.

Stochastic ion channel dynamics and circadian modulation of Orco conductance explain modeled ORN firing patterns

Our modeling results provide mechanistic insight into the irregular spontaneous activity of pheromone-sensitive ORNs in M. sexta, highlighting how stochastic ion channel dynamics and circadian modulation of ion conductance together shape spiking patterns. By employing a Langevin approximation of the stochastic HH formalism, we demonstrate that the intrinsic noise arising from the random gating of ion channels is sufficient to generate both isolated spikes and bursts of spikes, as observed experimentally. Our model effectively recapitulates the spike distributions found in vivo and accounts for the role of channel noise as a critical determinant of ORN firing patterns.

A key contribution of our model is the incorporation of the Orco ion channel as a pacemaker component, with a conductance that varies linearly with cAMP concentration. We show that circadian modulation of Orco conductance alone is sufficient to produce the observed time-of-day dependent changes in spike distribution. This finding suggests that the rhythmic expression of cAMP levels across the circadian cycle may act as a primary driver of diurnal changes in ORN excitability, without requiring additional external inputs or modulation of other ion channels.

Importantly, our results reveal that the same set of ion channels can support the emergence of both single action potentials and complex bursting behaviors purely through the stochastic variability of their gating kinetics. This unifying explanation reduces the need for invoking separate biophysical mechanisms or circuit-level inputs to explain these distinct firing modes.

For simplicity, the circadian variation in cAMP concentration was modeled using a sinusoidal function. While this approximation is sufficient for capturing the qualitative features of the rhythmic modulation, future work could enhance the biological realism of the model by coupling Orco conductance dynamics to a molecular circadian oscillator, such as a Goodwin-type TTFL.

Incorporating this layer would allow for a more detailed exploration of how molecular clock components regulate cAMP synthesis and degradation. Additionally, further extensions could include downstream regulatory mechanisms such as phosphorylation and dephosphorylation events mediated by kinases and phosphatases under circadian control, potentially contributing to fine-tuned modulation of ion channel function over the daily cycle.

Together, these findings demonstrate the value of a biophysically grounded, noise-inclusive modeling framework for understanding complex temporal patterns in sensory neuron activity and open the door for future integrative models linking molecular circadian clocks to cellular excitability.

eLife Assessment

This valuable study uses technically compelling long-term in vivo recordings and computational modeling to investigate whether hawkmoth olfactory receptor neurons show circadian modulation of spontaneous firing. The authors further propose the provocative model that post-translational mechanisms, rather than the transcriptional-translational processes, may contribute to circadian regulation of neuronal excitability. However, the evidence for circadian firing in these neurons, and for post-translational modification of Orco as the underlying mechanism, remains incomplete. In contrast, the study does provide strong evidence that the application of cyclic nucleotides can modulate Orco-dependent activity at a single time point, and reports that the temporal pattern of Orco transcript abundance is not circadian.

Joint Public Review:

This manuscript puts forward the provocative idea that a posttranslational feedback loop regulates daily and ultradian rhythms in neuronal excitability. The authors used in vivo long-term tip recordings of the long trichoid sensilla of male hawkmoths to analyze spontaneous spiking activity indicative of the ORNs' endogenous membrane potential oscillations. This firing pattern was disrupted by pharmacological blockade of the Orco receptor. They then use these recordings together with computational modeling to predict that Orco receptor neuron (ORN) activity is required for circadian, not ultradian, firing patterns. Orco did not show a circadian expression pattern in a qPCR experiment, and its conductance was proposed to be regulated by cyclic nucleotide levels. This evidence led the authors to conclude that a post-translational feedback loop (PTFL) clockwork, associated with the ORN plasma membrane, allows for temporal control of pheromone detection via the generation of multi-scale endogenous membrane potential oscillations. The findings will interest researchers in neurophysiology, circadian rhythms, and sensory biology. However, the manuscript has limited experimental evidence to support its central hypothesis and is undermined by several assumptions that underlie their data analysis and model builds, as well as insufficient biological data including critical controls to validate and/or fully justify the model the authors are proposing.

Strengths:

The authors raise several intriguing model-based hypotheses regarding the mechanisms that underlie the generation of olfactory rhythms. The electrophysiological approach and the long-term recording paradigm are elegant and technically impressive. In the revised version, the authors have added additional qPCR data supporting the lack of rhythmic Orco transcript expression and included a new figure suggesting that cAMP can modulate Orco conductance.

Major weaknesses:

(1) The cAMP experiment was only conducted at one time-point, which is insufficient to support the central claim that "AMP and cGMP may have ZT-dependent effects on Orco conductivity".

(2) The revised manuscript continues to rely heavily on prior publications or defers key mechanistic questions (or important manipulations) to future studies. In its current form, the evidence presented remains insufficient to support the central claim that a PTFL constitutes the primary underlying circadian clock mechanism. The proposed model is intriguing, but the data provided do not yet directly demonstrate the novel mechanism.

Author response:

The following is the authors’ response to the previous reviews

eLife Assessment

This valuable study uses technically compelling long-term in vivo recordings and computational modeling to investigate whether hawkmoth olfactory receptor neurons show circadian modulation of spontaneous firing. The authors further propose the provocative model that post-translational mechanisms, rather than the transcriptional-translational processes, may contribute to circadian regulation of neuronal excitability.

We are pleased that our study is recognized as valuable and that our technically very challenging long-term in vivo recordings and computational modeling are appreciated. We agree that we are proposing a provocative model that opposes the current hypothesis in chronobiology, which suggests that all observed biological circadian rhythms are outputs of a transcriptional-translational feedback loop (TTFL) clock. Instead, we suggest that a cell comprises, in addition to the TTFL clock, other posttranslational feedback loop (PTFL) clocks without the need of daily transcription and daily degradation of its core elements. While the circadian TTFL clock is entrained to the daily light-dark cycle, the circadian PTFL clocks are suggested to be entrained to other daily cues such as to the availability of pheromone, or the daily changing levels of hormones and second messenger levels. Our novel hypothesis proposes that the TTFL and PTFL clocks are coupled and linked, constituting an adaptive, plastic network that can tune and phase-lock to different external and internal Zeitgeber signals. However, we certainly do not claim that the TTFL circadian clock is not at all involved in the circadian control of the ORN’s circadian membrane potential rhythms. We clarified our manuscript accordingly.

However, the evidence for circadian firing in these neurons […] remains incomplete.

As requested by the reviewers during the previous round of review, we had provided the results of RAIN analysis (Thaben and Westermark, 2014) of individual animals in our first revision (Fig. 4A), which clearly shows that two-thirds of the population express circadian rhythms in key attributes that we used to characterize the spontaneous spiking activity. In the initial review it was assumed by the reviewers that phase alignment of dispersed rhythms would bias interpretations of rhythmicity. After having established with RAIN that individuals show circadian rhythms, albeit dispersed across the population due to the lack of a zeitgeber in DD conditions, phase-alignment of recordings from DD animals is a valid next step to prepare the data for statistical analysis across the population. Phase-alignment of desynchronized rhythms is a generally accepted and necessary method employed in chronobiology (e.g., for insect ORNs: Gosh et al., 2024). It is proven as prerequisite to find and statistically analyze rhythmicity in complex, desynchronized data.

As we explained in the previous rebuttal and in our first revision, rhythms in electrical activity of insect ORNs cannot be easily synchronized by the light-dark cycle alone, but appear to require daily cycles of pheromone, as shown in other moth species (Gosh et al., 2024). Please be aware that the animals that we used here have never been exposed to pheromone, as we state in the Methods. Therefore, this lack of pheromone exposure can explain why about one third of our experimental population is not expressing any daily or circadian rhythmicity in spiking attributes. This is an important result of our manuscript, reported for the first time for Manduca sexta, providing evidence for our hypothesis that it is not the LD-entrained circadian TTFL clock that governs electrical activity rhythms in ORNs. As we explained in the first revision, and clarified further here in the second revision, we cannot phase-synchronize our animals with cycles of pheromone application in our experimental paradigm because we are researching circadian rhythms in spontaneous spiking activity and not pheromone responses. We failed to obtain phase-alignment with a single pheromone pulse the night before the experiments started. These data were added as supplementary Figure to Fig. 3 in the first revision. Here, we further revised our manuscript to clarify this important finding.

Thus, as requested by the reviewers in the initial review, we could successfully confirm our previous results of circadian firing in ORNs and the disruption of these circadian rhythms with Orco antagonist OLC15 with RAIN. In the current review, the reviewers raise no further specific critical points or comments that would doubt our careful rhythm analysis of our long-term recordings. Thus, we conclude that we provided clear evidence for our central, exciting new finding. For the first time we demonstrated an unexpected new task for Orco: Orco controls the circadian firing pattern in the spontaneous activity, and thus, of the ORN’s membrane potential, via its property as leak/pacemaker channel.

However, the evidence […] for post-translational modification of Orco as the underlying mechanism remains incomplete.

We agree with the reviewers that there are many more experiments and combined efforts of biochemists, structural biologists, and electrophysiologists required to provide complete evidence for post-translational modification of Orco and to reveal the underlying mechanism of its circadian control. It is beyond the scope of the current manuscript to provide all details of post-translational control of Orco.

The reviewers asked previously for additional evidence that Orco transcription is not controlled via the TTFL clock. As requested, we provided extended qPCR–based evidence that Orco, in contrast to timeless, is not controlled by the TTFL clock on the transcriptional level (Fig. 6 in Revision 1). Furthermore, we added a new result in Revision 1 to demonstrate cAMP-dependent post-translational modulation of Orco open-time probability (Fig 9 in Revision 1) at a ZT at which antennal cAMP levels are low (Schendzielorz et al., 2015). We already showed in Flecke et al., 2010, that the addition of cAMP at different ZTs increases the spontaneous spiking activity only at specific ZTs. Here, we show that the effect of cAMP depends on Orco. Since Orco’s circadian role is not mediated via TTFL control, it can be concluded that post-translational mechanisms provide daily/circadian temporal control. In this additional Figure we provide statistically significant proof that, in agreement with our model-prediction, Orco’s circadian control of the ORN spontaneous activity could be mediated via the second messenger cAMP. ZT-dependent input for Orco would be provided via daily changes in cAMP levels (Schendzielorz et al., 2015).

In contrast, the study does provide strong evidence that the application of cyclic nucleotides can modulate Orco-dependent activity at a single time point, and reports that the temporal pattern of Orco transcript abundance is not circadian.

We appreciate that the reviewer confirms that our revised manuscript with additional experiments now provides strong evidence that cAMP modulates Orco-dependent spontaneous activity of M. sexta ORNs. Since we already published that cAMP levels expresses daily rhythms in M. sexta antennae (Schendzielorz et al., 2015), and in vivo cAMP infusion increases spontaneous activity and sensitizes pheromone detection (Flecke et al., 2010), and our computational model here proves that circadian modulation of open time probability of Orco is sufficient to explain our experimental data, it is sufficient for our conclusions to test just the one specific zeitgeber time when endogenous cAMP levels are low and pharmacological cAMP increase has the strongest impact. To further reveal complete ZT-dependence of cAMP modulation of Orco´s control of spontaneous activity is beyond the scope of the current manuscript and not part of the current research question. The structure of Orco is extraordinarily conserved during evolution, thus, the cited experimental results from other laboratories and other species showing that Orco is a hub for posttranslational modification are very likely generalizable to different insect species. We clarified the manuscript accordingly. In Drosophila, Orco has at least 5 phosphorylation sites for protein kinase C (PKC), is cAMP-dependently sensitized, and has a Ca2+/calmodulin binding site that orchestrates the localization of the OR-Orco heteromer to the cilia. However, in fruit fly and other insects, so far, it can only be speculated how circadian control is provided for Orco, since there are no other publications that examine the circadian regulation of Orco in detail. We clarified our manuscript accordingly.

To summarize, the logical conclusion based on our newly provided data is that the current hierarchical hypothesis in chronobiology based solely on a circadian TTFL clock that controls Orco transcription does not explain our findings in hawkmoth ORNs. Therefore, we suggest a new systemic hypothesis based upon coupled TTFL and PTFL circadian clocks that can also reconcile otherwise inconsistent data published for insect and mammalian circadian clocks (please see reviewed data in: Stengl and Schneider, 2024). We clarified our manuscript in the second revision and added a new Figure 10 to illustrate our novel hypothesis.

However, the findings are incomplete to exclude a role for transcriptional-translational mechanisms and their associated multi-layered controls in circadian regulation.

We certainly do not imply excluding a role for the TTFL clock in (indirectly) affecting circadian control of the membrane potential of ORNs. The new qPCR experiments added in the first revision clearly show that the circadian control mediated via Orco is not an output of the TTFL clock via transcriptional control of Orco. Instead, we predict links between a PTFL membrane clock comprising Orco as hub to integrate posttranslational control and the TTFL nuclear clock. We clarified our manuscript accordingly, adding a new Figure 10 to further illustrate and visualize our hypothesis. The predictions of this systemic hypothesis will be challenged in further experiments that, however, are beyond the scope of the current manuscript.

Joint Public Review:

This manuscript puts forward the provocative idea that a posttranslational feedback loop regulates daily and ultradian rhythms in neuronal excitability. The authors used in vivo long-term tip recordings of the long trichoid sensilla of male hawkmoths to analyze spontaneous spiking activity indicative of the ORNs' endogenous membrane potential oscillations. This firing pattern was disrupted by pharmacological blockade of the Orco receptor. They then use these recordings together with computational modeling to predict that Orco receptor neuron (ORN) activity is required for circadian, not ultradian, firing patterns. Orco did not show a circadian expression pattern in a qPCR experiment, and its conductance was proposed to be regulated by cyclic nucleotide levels. This evidence led the authors to conclude that a post-translational feedback loop (PTFL) clockwork, associated with the ORN plasma membrane, allows for temporal control of pheromone detection via the generation of multi-scale endogenous membrane potential oscillations. The findings will interest researchers in neurophysiology, circadian rhythms, and sensory biology. However, the manuscript has limited experimental evidence to support its central hypothesis and is undermined by several assumptions that underlie their data analysis and model builds, as well as insufficient biological data including critical controls to validate and/or fully justify the model the authors are proposing.

We want to remind our reviewers that we used “ORN” as abbreviation for olfactory receptor neuron (= sensory receptor neurons, a.k.a. olfactory sensory neuron (OSN)) and not for Orco receptor neuron, although we focus on the function of Orco. Accordingly, our central finding is that Orco as ion channel is required for the daily/circadian modulation of spontaneous action potential activity generated by the olfactory receptor neurons in the absence of pheromone stimulation.

We do not understand the specific basis for the conclusions of the reviewers. Therefore, we ask to please specify what experimental evidence is missing to support our central hypothesis that Orco is not directly TTFL- but PTFL clock-controlled, and to name specifically what the “several assumptions” are that undermine our careful data analysis and model builds. Which specific argument in our previous rebuttal was wrong, was not conclusive? Furthermore, please specify your claim that “critical controls are missing”. Which controls are missing for which experiments? We did add a new figure panel in the first revision to demonstrate that neither the addition of DMSO (the OLC15 solvent) nor the repeated attachment of the recording electrode, which was necessary to obtain paired datasets, altered the spontaneous spiking activity (Fig 1B in Revision 1). Furthermore, we expanded the time series of qPCR data and added tim as positive control to Orco (Fig 7), strengthening our argument that Orco expression is not under TTFL control. Dose-response curves of various Orco agonists and antagonists have been published before (see our references in Revision 1) and are therefore not repeated here.

Our newly added data confirm what our modeling predicted: cAMP increases spontaneous ORN activity dependent on Orco. Previous publications provide evidence for daily rhythms in cAMP concentrations in hawkmoth antennae (Schendzielorz et al., 2012).

As is true for any other hypothesis, a hypothesis can only be falsified but not validated and needs to be tested by many experiments from many laboratories over a long time until it will be replaced by the next hypothesis that better explains accumulating contradicting evidence. We are very much looking forward to experimental challenges of our provocative new hypothesis by colleagues in the field of olfaction and of chronobiology. We are convinced that our manuscript will greatly stimulate the field, possibly provoking a paradigm switch in chronobiology and in olfactory research.

Strengths:

The authors raise several intriguing model-based hypotheses regarding the mechanisms that underlie the generation of olfactory rhythms. The electrophysiological approach and the long-term recording paradigm are elegant and technically impressive. In the revised version, the authors have added additional qPCR data supporting the lack of rhythmic Orco transcript expression and included a new figure suggesting that cAMP can modulate Orco conductance.

We thank the reviewers for their acknowledgement of our careful work and hope that our further revisions and clarifications help to argue our case.

Major weaknesses:

(1) The cAMP experiment was only conducted at one time-point, which is insufficient to support the central claim that "AMP and cGMP may have ZT-dependent effects on Orco conductivity".

We agree with the reviewers and revised our discussion accordingly to clarify that in this manuscript it is not our central claim that cAMP and cGMP may have ZT-dependent effects on Orco conductivity. Instead, our data show for the first time that Orco controls circadian rhythms of spontaneous activity of ORNs and that the circadian rhythmicity of spontaneous activity is lost when Orco is blocked. Therefore, we provide novel experimental evidence that Orco is a prerequisite to the circadian rhythmicity of spontaneous activity and thus, to circadian rhythms in the membrane potential of ORNs. Furthermore, as requested by the reviewers we provided clear evidence in the first revision that Orco is not controlled at the transcriptional level by the TTFL clock, in contrast to the TTFL clock protein TIMELESS. Thus, it follows logically that Orco is under post-transcriptional control. Since cAMP levels show circadian oscillations and Orco is gated by cAMP (Fig 9 in Revision 1), we used our computational model to show that a cAMP-dependent increase in Orco conductance alone, via daily oscillating concentrations of cAMP, is sufficient to explain our findings. Therefore, we propose here that daily/circadian oscillations of cAMP modify spontaneous spiking activity via Orco on a posttranscriptional level. But it certainly does not provide all evidence for respective mechanisms of how this cAMP modulation of Orco is obtained, since this is beyond the scope of the current manuscript.

Since we realized that it is difficult for our readers to visualize a circadian PTFL membrane clock we added a new hypothesis-Figure (Figure 10) and considerably focused and clarified especially the discussion of our manuscript. We pointed out that a membrane-associated signalosome that comprises delayed negative feedback mechanisms, and, thus, constitutes an oscillator, a “membrane clock” that generates oscillations. Based upon our data we propose a membrane-associated signalosome constituting a PTFL circadian clock with Orco as central element. This PTFL membrane clock generates superimposed ultradian and circadian rhythms in its outputs: rhythms in the membrane potential, Ca2+, and cAMP levels. The PTFL clock comprises positive feedforward elements that upregulate its outputs, resulting in more depolarization, higher Ca2+- and higher cAMP levels. Via the clock’s delayed negative feedback mechanisms these outputs are downregulated, again, resulting in hyperpolarization, decreasing Ca2+- and cAMP levels. This signalosome comprises the pacemaker channel Orco as a central hub that is controlled via changes in voltage, Ca2+, and cAMP levels. Nevertheless, we predict coupling between the multiscale PTFL membrane clock and the TTFL circadian clock in the nucleus to obtain stable circadian rhythms. As likely mechanism of coupling we predict that Ca2+- and cAMP-dependent kinases interlink both types of clocks, thereby obtaining robust and at the same time flexible interlinked cellular rhythms.

We hope to now successfully clarify and to visualize our central hypothesis of our manuscript that Orco is not directly controlled by a TTFL circadian clock but is a central element of a membrane-associated posttranslational feedback loop clock (PTFL) clock that is linked to but not forced by the TTFL clock which is predicted to control intracellular Ca2+ homeostasis in a circadian rhythm.

(2) The revised manuscript continues to rely heavily on prior publications or defers key mechanistic questions (or important manipulations) to future studies. In its current form, the evidence presented remains insufficient to support the central claim that a PTFL constitutes the primary underlying circadian clock mechanism. The proposed model is intriguing, but the data provided do not yet directly demonstrate the novel mechanism.

We do not understand why the reviewers considers it to be problematic that we “continue to rely heavily on prior publications”. Certainly, we built upon previous publications of our lab as well as on manifold experimental data published by other laboratories in the field of insect olfaction. Our ample citations demonstrate that we have an overview both of the current state of literature and relevant previous literature, dating back to the very first experiments that pioneered pheromone transduction in insects. Based upon our extensive knowledge and experimental data collected in insect olfaction and based on very careful, critical, rigorous analysis of our data and data published by others, we were able to come up with a novel interpretation of the current literature about insect olfaction that differs considerably from the current main views. We consider this to be our strength and judge it as good scientific practice and not a flaw of our work. However, since we do not focus on OR-Orco heteromers and their function in pheromone/odor transduction in the cilia in the current manuscript, we considerably shortened this part of the discussion, avoiding pointing out that highly sensitive moth pheromone transduction greatly differs from less sensitive general odor transduction in Drosophila. Furthermore, since here we focus on cAMP, but not on cGMP-dependent modulation of Orco, we also deleted/considerably shortened this part of our discussion.

We certainly agree with the reviewers that, while the data provided in the current manuscript are a logical basis for developing our novel hypothesis, they are not a direct and sufficient demonstration of proof and we are not able yet to directly demonstrate and explain the novel mechanism predicted. We would like to point out that if we provided this final proof, it would not be any more a novel hypothesis, but only a novel finding.

We agree with the reviewers that our provocative hypothesis requires rigorous testing by many further experiments, hopefully not only by our laboratory, but hopefully stimulating new experimental challenges by other laboratories employing different species. But certainly, these experiments with proof-of-principle will take many years and are beyond the scope of our current research paper.

As per eLife’s assessment system we would like to ask the reviewers to provide detailed feedback as to which experiments/results within the scope of this manuscript would complete this work, or how they think this study should be framed in the light of the results that we obtained. Nevertheless, we hope that with our current careful review the reviewers will be more convinced by our arguments and experiments as valid basis for our provocative new hypothesis.

Version 3

Abstract

The mating behavior of nocturnal Manduca sexta hawkmoths is under strict temporal control. It is orchestrated via circadian and ultradian oscillations in sex-pheromone stimuli as social Zeitgeber. The extremely sensitive pheromone-detecting olfactory receptor neurons (ORNs) that innervate the long trichoid sensilla on the male’s antennae are peripheral circadian clocks. They express the transcriptional-translational feedback loop (TTFL) circadian clockwork, best characterized in Drosophila melanogaster. In hawkmoths, it is still unknown whether or how the ORN TTFL clockwork regulates the daily rhythms in pheromone sensitivity and in temporal resolution of ultradian pheromone pulses as prerequisites to the temporal regulation of hawkmoth mating behavior. We hypothesize that, rather than the slow TTFL clock, a more rapidly adaptive post-translational feedback loop (PTFL) clockwork, assembled in a signalosome in the ORN plasma membrane, allows for temporal control of pheromone detection via generation of multiscale endogenous membrane potential oscillations. The potential oscillations of the PTFL clock could rapidly synchronize to oscillations of pheromone stimuli at different time scales, thus enabling the prediction of stimulus patterns as a mechanism for active sensing. With in vivo long-term tip recordings of long trichoid sensilla of male hawkmoths, we analyzed the spontaneous spiking activity indicative of the ORNs’ endogenous membrane potential oscillations. Consistent with our hypothesis of a multiscale PTFL clock in hawkmoth ORNs, spontaneous spiking was modulated on ultradian and circadian time scales, with maximum activity at night. When we blocked the evolutionarily conserved olfactory receptor coreceptor (Orco), the circadian modulation was abolished but the ultradian frequency modulation of the spontaneous activity remained. Consistent with PTFL control, Orco was not under the transcriptional control of the TTFL clock, but its modulation of spontaneous spiking activity was dependent on cAMP. We could replicate the experimental data in a conductance-based computational model of an ORN. In this model, Orco conductance changed as a function of fluctuating 2nd messenger levels. This study demonstrates that a PTFL clock is sufficient to impose a circadian pattern on ORN sensitivity.

Introduction

Biological clocks generate endogenous physiological and behavioral oscillations on multiple timescales that persist in the absence of any external timing cues under constant conditions (Lamont and Amir, 2010). Such clocks are essential to generate internal rhythms and to permit their synchronization to external Zeitgebers at time scales from milliseconds to years. These synchronized endogenous multiscale clocks thus provide both a continuous time frame for physiology and behavior - the organism’s “present time” - and enable organisms to predict and respond to regular, periodic environmental changes (Stengl and Schneider, 2024). Even though endogenous clocks are widespread in prokaryotes and eukaryotes (Géron et al., 2024; Hall and Rosbash, 1993; Millar, 2016; Thaiss et al., 2014), current knowledge about how multiscale clocks are linked over widely diverging time scales remains sparse.

Insect circadian clock neurons, which are tightly connected to visual brain circuits, are amongst the best studied biological clocks (Hall and Rosbash, 1993; Nishiitsutsuji-Uwo and Pittendrigh, 1968; Reinhard et al., 2024). They regulate physiological and behavioral oscillations that underlie sleep-wake rhythms tied to the 24 h light-dark (LD) cycles on Earth (Dubowy and Sehgal, 2017). Circadian clock neurons possess a molecular clockwork of transcriptional-translational feedback loops (TTFLs). Entrained to the LD cycle, this TTFL clock generates endogenous oscillations of clock gene mRNA and protein levels with a cycle period of ∼24 h via negative feedback by proteins that inhibit their own transcription (Hardin, 2011). It is assumed for both insects and mammals that this TTFL comprises the master clockwork that regulates all physiological and behavioral rhythms in the circadian time range through genetically controlled output pathways (Hastings et al., 2020).

Molecular TTFL clockworks were not only found in central brain clock neurons but also in sensory neurons, such as olfactory receptor neurons (ORNs) in insect antennae, where they play a role in odor-dependent circadian behaviors like feeding, pollination, reproduction, and navigation (Flecke et al., 2010; Flecke and Stengl, 2009; Krishnan et al., 1999; Merlin et al., 2009, 2006; Page and Koelling, 2003; Plautz et al., 1997; Sauman and Reppert, 1998; Schendzielorz et al., 2015; Schuckel et al., 2007; Tanoue et al., 2004; Zhou et al., 2005). Little is known about how the TTFL clockwork of insect ORNs regulates circadian rhythms in chemosensory sensitivity that underlie the circadian control of these different behaviors.

At night, male nocturnal Manduca sexta hawkmoths are very sensitive to the sex pheromone blend that their conspecific females release in an intermittent pattern (Itagaki and Conner, 1988). The intermittency is a prerequisite to maintain the male’s search flight, which is suspended at pulse frequencies >30 Hz (Baker et al., 1985; Baker and Vogt, 1988; Stengl, 2010). The pheromone pulses are resolved of up to 3 Hz by pheromone-specific ORNs in the antenna’s long trichoid sensilla (Lee and Strausfeld, 1990; Marion-Poll and Tobin, 1992). Pheromone-specificity is provided by odor receptor (OR) subunits in the ciliary membranes of ORNs that heteromerize with the conserved olfactory receptor co-receptor Orco (Nakagawa and Vosshall, 2009; Stengl and Funk, 2013; Wicher and Miazzi, 2021). The insect odor transduction cascade is still under debate and seems to involve both ionotropic and metabotropic pathways (Sato et al., 2008; Schneider et al., 2025; Wicher et al., 2008). While there is no general agreement on all roles of Orco, it is known that in heterologous expression systems Orco homo- and heteromers form a slow, leaky, non-specific cation channel, which continuously depolarizes the cells (Jones et al., 2011; Sargsyan et al., 2011; Sato et al., 2008; Wicher, 2018; Wicher et al., 2008). In vivo tip recordings of long trichoid sensilla in M. sexta confirmed that Orco acts as leaky cation channel, acting as a pacemaker channel, because it promotes ultradian membrane potential oscillations that underlie the spontaneous action potential activity (Nolte et al., 2013). These potential oscillations turn the ORNs into ultradian temporal filters that are sensitive to the frequency of the pheromone pulses they encounter. Pheromone pulses that arrive at a similar frequency as the endogenous sub-threshold membrane potential oscillations would synchronize these oscillations and would be more likely to elicit action potential trains. Therefore, this pacemaker channel-dependent mechanism would allow for active sensing if endogenous membrane potential oscillations are in the physiological range that is necessary for tracking intermittent pheromone pulses.

It is not known whether the Orco-dependent membrane potential oscillations also occur at circadian time scales, and whether or how they are linked to ultradian potential oscillations. One hypothesis is that the molecular TTFL clockwork of ORNs exclusively controls endogenous circadian rhythms of the membrane potential via circadian expression of pacemaker channels like Orco. Alternatively, the plasma membrane of ORNs could by itself comprise a signalosome that constitutes an autonomous clock that is entrained by, but does not rely on, the TTFL clock. Constant levels of colocalized ion channels with antagonistic effects that do not require daily degradation and daily transcription via the TTFL clock, but with modulation of conductance by post-translational modifications, would constitute a more economical and faster adjustable mechanism to generate endogenous membrane potential oscillations (Stengl and Schneider, 2024). This would facilitate active sensing by signalosome-anchored receptors, rapidly tuning odorant/pheromone detection to behaviorally relevant Zeitgeber signals in the insect’s environment, other than the LD cycle that is linked to the TTFL clock.

Here, we performed a combination of experimental and computational modeling studies to investigate whether Orco plays a role in active sensing at the circadian and/or ultradian time scale via generation of endogenous membrane potential oscillations. We hypothesize that Orco is part of a post-transcriptional feedback loop (PTFL) clockwork, assembled in a signalosome in the ORN somatic plasma membrane, rather than being TTFL controlled. To capture naturally occurring, physiologically relevant processes in insect ORNs we performed in vivo long-term tip recordings of the spontaneous spiking activity of pheromone-sensitive trichoid sensilla of intact but restrained hawkmoth males.

We corroborated the experimental data with a Hodgkin-Huxley (HH) based computational ORN model. In general, the widely used deterministic HH models do not include circadian control of ion channels. Moreover, the highly dynamic, bursting firing patterns of ORNs cannot be described by the traditional deterministic modeling of neuronal activity. This constitutes a drawback because stochasticity plays a crucial role in shaping firing patterns which are influenced not only by external stimuli but also by the inherent noise involved in channel gating. Therefore, to address the above points realistically, we developed a Langevin formulation of the HH model, which explicitly accounts for the noise introduced by the random opening and closing of ion channels and circadian modulation of the Orco channel conductance. By integrating channel noise, we could explore how circadian regulation of ionic currents contributes to the variability of firing patterns and how this noise interacts with external stimuli to influence olfactory signal processing.

Our model aims to highlight the advantages of using a stochastic framework for modeling the circadian activity of ORNs. The combined experimental and modeling efforts generate a more complete picture and predictions concerning these autonomous processes in sensory neurons, which are required for future exploration of physiologically relevant anticipatory behaviors that are driven by biological clocks.

Methods

Animals

Nocturnal M. sexta hawkmoths were bred and raised from eggs in the rearing facility at the University of Kassel. Males and females were separated during pupal stages. Males were housed isolated from the females to avoid exposure to female pheromones. All experiments were performed with adult males reared in a long-day photoperiod (L:D 17:7 h to prevent diapause) at 25 °C with relative humidity of about 55%. Larvae were fed with an artificial diet modified after Bell and Joachim (1976); adult moths could feed on sugar solution with added vitamins (Roth) ad libitum (Riffell et al., 2008).

While all animals were exposed to the same lighting regime as Zeitgeber to entrain them up until the start of the experiments, we also attempted pheromone exposure as additional circadian Zeitgeber to phase-align spontaneous ORN activity in a subset of males (Ghosh et al., 2024). Without allowing direct contact, virgin males were exposed at Zeitgeber time (ZT) 16 to females and naturally occurring pheromone blends in closed mating cages for 30 mins. Males were then moved back into isolation before tip recordings (see below) were started at ZT 17 on the same day. Since we focused on spontaneous spiking activity rather than on pheromone-dependent responses and because repetitive or long pheromone exposure changes the composition of 2nd messenger modulated ion channels in ORNs (Gawalek and Stengl, 2018), we did not employ any further pheromone synchronization protocols.

Solutions

Ringer compositions were taken from (Pézier et al., 2007) and contained (in mM): 6.4 KCl, 12 MgCl2, 1 CaCl2, 12 NaCl, 10 HEPES, 340 glucose for hemolymph ringer (HLR) at 450 mosmol/kg, and 172 KCl, 3 MgCl2, 1 CaCl2, 25 NaCl, 10 HEPES, 22.5 glucose for sensillum lymph ringer (SLR) at 475 mosmol/kg, both at pH 6.5.

The Orco antagonist OLC15 (N-(4-butylphenyl)-2-((4-ethyl-5-(2-pyridinyl)-4H-1,2,4-triazol-3-yl)thio)acetamide (Chen and Luetje, 2012); kindly provided by Dr. D. Wicher, Max Planck Institute for Chemical Ecology, Jena, Germany) was dissolved in DMSO to a concentration of 100 mM (stock solution stored at 4-8 °C) before being diluted in SLR to a final concentration of 50 µM to be used within one week (stored at 4 °C). We determined in previous studies that, compared to other Orco antagonists, OLC15 is the most specific (Chen and Luetje, 2012; Nolte et al., 2016, 2013; Pask et al., 2013). It is effective at 50 µM, according to dose-response curves, and dose-dependently antagonizes the action of the Orco agonist VUAA1 in intact hawkmoths (Nolte et al., 2016, 2013). DMSO percentage was 0.05% in the final solution.

The cAMP analog 8-bromo-cAMP acetoxymethyl ester (8-Br-cAMP-AM; Biolog Life Science Institute GmbH & Co. KG) was dissolved in DMSO to a concentration of 10 mM (stock solution stored at −20°C) and diluted to a final concentration of 100 µM in either SLR or SLR + 50 µM OLC15.

Electrophysiology

The spontaneous electrical activity of the unstimulated pheromone-sensitive ORNs in a single long trichoid sensillum was recorded extracellularly as in vivo tip recording (Kaissling et al., 1987). 1-2-d-old male M. sexta that were raised and entrained to Zeitgebers as described above were fixed in a custom-made Teflon holder with adhesive tape. The right antenna used for recordings was oriented with its dorsal surface facing up and immobilized near the base with dental wax (Boxing wax strips, KerrHawe SA). Electrodes (Ag/AgCl wire in a glass pipette filled with SLR for the recording electrode and HLR for the reference electrode) were pulled (DMZ-Universal Puller, Zeitz Instruments) from borosilicate capillaries (OD:1.56 mm, ID: 1.17 mm, without filament; Science Products) to a tip diameter of about 2 μm. All recordings were made at a distance of 2/3 to 3/4 of the antenna length towards its distal end because pheromone-sensitive long-trichoid sensilla (sensilla trichodea type 1) occur in a characteristic, homogenous pattern along the hawkmoth antenna (Kaissling et al., 1989; Kalinová et al., 2001; Lee and Strausfeld, 1990). Furthermore, all ORNs in long trichoid sensilla express the same Orco staining pattern (Nolte et al., 2016), share largely overlapping ion channel populations, and obvious differences in the activity of ORNs from different annuli have not been reported (c.f., Dolzer et al., 2003; Gawalek and Stengl, 2018; Schneider et al., 2025). Thus, it is very likely that pheromone encoding mechanisms are not locally restricted along the flagellum. Recording and reference electrodes were located on the same annulus: After cutting off the distal 20 annuli of the antenna with micro scissors, the reference electrode was inserted to a depth of 2 annuli into the antennal lumen from the cut end. The cut was covered with electrode gel (GE Medical Systems Information Technology) to prevent desiccation. Under microscopic control, some tips of the long trichoid sensilla of the annulus were cut off with sharpened Dumont forceps, and the recording electrode was slipped over a single sensillum of an upper sensillar row. Electrodes were moved with manual micromanipulators (MMJ, Märzhäuser Wetzlar; and Leitz, Leica Microsystems) or with motorized micromanipulators (Microstar, Scientifica) with a joystick controller (Scientifica).

ORN activity was amplified 200-fold (custom built amplifier with 1012 Ω input impedance, or ELC-01 MX in combination with DPA-2FX, npi, or EXT 10-2F, npi), low pass filtered at 1.3 kHz, digitized at 20 kHz with a Digidata (1550A or 1550B, Molecular Devices), and saved for offline analysis with Clampex 10 (pCLAMP suite, Molecular Devices).

All tip recording experiments were performed at room temperature (21-27 °C) and a relative humidity of 35-60%. These environmental fluctuations across experiments were based on seasonal influences. Throughout each experiment, conditions were stable. To ensure that neither electrode reattachment to the same trichoid sensillum nor application of DMSO influenced spiking activity, we performed two sets of paired experiments at ZT 1-3 in which we recorded the spontaneous activity before and after removing and reattaching the SLR-filled recording electrode to the same sensillum, or recorded the spontaneous activity of the same sensillum in SLR and SLR + 0.15% DMSO. Both treatments did not affect spontaneous spiking activity (Figure 1B).

Attributes of pheromone-sensitive ORN spiking patterns.

(A) Spikes were identified by their prominence in the raw ORN recording. The inter-spike interval (ISI) was defined as the time from one spike to the next. A train with ≥2 spikes with ISI ≤ 50 ms was considered a burst. The inter-burst interval (IBI) was the ISI between the last spike of the first burst and the first spike of the subsequent burst, omitting any individual spikes in between. Burst duration was the time between the first and last spike of a burst. The inter-event interval (IEI) included the ISI between bursts and individual spikes. The number of action potentials (spikes) per burst, and percentage of total action potentials as part of bursts were computed. The intra-burst interval was the ISI within a burst. Frequency of spikes, bursts, events and intra-bursts were determined as inverse of ISI, IBI, IEI and intra-burst interval. Detaching and reattaching the recording electrode to a long trichoid sensillum (Bi; paired t-test t(9) = −2.184; p = 0.057; N = 10) or adding DMSO to the sensillum lymph ringer (SLR) (Bii; Wilcoxon signed-rank test Z = 0.153; p = 0.922; N = 10) did not affect the mean spiking frequency.

We recorded ORN activity either in long-day conditions (17:7 LD) or constant darkness (DD) to examine the free-running peripheral ORN clock. For LD experiments, the light regime continued as in the rearing facility during the experiment. For DD and OLC15 experiments, the lights were turned off at the beginning of the recording, i.e., all animals were exposed to the same light Zeitgeber until the beginning of the experiment. To examine the effect of Orco on the spontaneous action potential (AP) activity pattern, we used the recording electrode to infuse OLC15 in DD conditions. To examine the effect of cAMP on Orco gating, we recorded the spontaneous ORN activity at the end of the moth’s activity phase (ZT 1-3) in SLR as control, and either 8-Br-cAMP-AM, or 8-Br-cAMP-AM + OLC15, infused through the recording electrode.

ORNs generate spontaneous action potentials. However, high spiking activity during control recordings is indicative of stress, whereas low/no activity could be indicative of damage. Therefore, we only considered experiments in which the average spiking frequency was between 2 Hz and 0.008 Hz in control conditions.

Data analysis

To characterize the spontaneous ORN spiking activity, we first detected spikes with a self-written program in Python. Using the scipy.signal package and find_peaks function, we used a moving mean with 100 samples (5 ms) window size for baseline correction, followed by applying a 8th order low-pass Bessel filter (2000 Hz cutoff frequency) to the raw voltage trace to facilitate spike detection within the following conditions: spike full width (3-15 ms), threshold search for amplitude of spike (0.5-2.5 mV) and threshold prominence (0.1). Two or more consecutive spikes with intervals ≤ 50 ms were considered a burst (Dolzer et al., 2001). We then calculated the following attributes in bins of either 10 min or 1 h for further analysis: instantaneous spike frequency (ISI, 1 / inter-spike interval), inter-burst interval (IBI, time between two consecutive bursts with no single spikes in between), inter-event interval (IEI, time between two events, i.e. two consecutive single spikes, two consecutive bursts, a burst and a consecutive spike, and vice versa), burst duration (time from first to last spike in a burst), percentage of all spikes in the experiment that are members of burst, and mean number of spikes per burst (Figure 1A). All statistics were done with those binned values.

Hawkmoth behavioral and physiological rhythms desynchronize across the population in the absence of Zeitgebers like circadian rhythms of pheromone stimulation and daily light-dark cycles. Therefore, we first used RAIN analysis (Thaben and Westermark, 2014) to identify circadian rhythms (24 ± 4 h) in the attributes above for each animal and condition (LD, DD, OLC15) in the 1 h-binned data. Data was downsampled to 2 h resolution and thus included every other 1 h bin to speed up computational time. To further quantify the circadian changes in attributes across animals, it was then necessary to phase-align the recordings obtained in DD and OLC15. To achieve this, we used an optimization procedure to calculate the optimal temporal lags for each dataset, aiming to maximize the overall cross-correlation between the calculated binned spike frequency series. The optimization was performed using simulated annealing, which provided a set of time shifts (lags) for each series. These lags were applied to shift the frequency data in a manner that maximized the synchronization between the different subjects’ data, resulting in a “subjective ZT” with subjective ZT 0 at the first maximum of the binned ISI for each individual. The objective function used for optimization was the negative sum of pairwise correlations between all series after applying the shifts. The resulting alignment allowed us to compare the frequency patterns across the different conditions.

Once the data were aligned, we computed the mean frequency across all subjects for each bin and each condition. Then, the first high activity period was identified by finding the maximum in the mean frequency series. Subsequently, we identified the second high activity period as the time 24 hours after the first one and the first low activity period was found by locating the minimum in the mean frequency series between the two high activity periods. Besides the mean frequency, we also searched for significant oscillations in IBI, IEI, burst duration, percentage of spikes that are part of bursts, and mean number of spikes per burst. For this we found the maximum value of these statistics within a 3 h range centered in the first low activity period and compared it with the values at the second high activity period.

Continuous wavelet and Fourier analysis were performed using PyBoat (version 0.9.12) (Mönke et al., 2020) to examine the time series of neuronal firing frequency. The continuous wavelet method allows us to detect and characterize rhythmic patterns within the data by identifying frequency components that change over time. Specifically, we used wavelet analysis to determine whether a neuron exhibited circadian and ultradian rhythms. Unlike traditional Fourier analysis, which provides an overall frequency composition but assumes stationarity, wavelet analysis is particularly useful for biological signals because it can reveal how these frequencies evolve dynamically. By applying this approach, we could assess not only the presence of ultradian rhythms but also whether their strength or occurrence fluctuated in a circadian manner, indicating modulation by the 24-hour cycle. For this analysis, we used the 10 min-binned data to have sufficiently high resolution to detect ultradian rhythms. Periods were determined as local maxima in the Fourier power spectra.

Real-time quantitative polymerase chain reaction (qPCR)

We followed the qPCR methodology detailed in (Schneider et al., 2025). Briefly, male moths were isolated and cultured until the second day after eclosion, and the antennae from four individuals were collected every 4 hours for one day, starting at ZT 0, using liquid nitrogen, with four biological replicates per time point. Total RNA was extracted using TRIzol® Reagent (Invitrogen) and quantified for concentration and purity with the NanoDrop ND-1000 spectrophotometer (Thermo Fisher Scientific). First-strand complementary DNA (cDNA) was synthesized using the PrimeScript™ RT reagent Kit with gDNA Erase (Takara Bio Inc.). Primers for Orco (F: GCTCGCTACCACCAAATTGC, R:TCGTGACCCAACTGACAACA; Gene ID: LOC115452348) and for timeless (tim) (F: TTAAGCCGACCGTAGTGCTG, R: CGTCTTCCGTCCATGTGTCT; Gene ID: LOC115446009) were designed using the Primer-BLAST online program of NCBI. The qPCR was conducted following the protocol of TB Green® Premix Ex Taq™ II (Tli RNase H Plus) (Takara Bio Inc.), with the program: 95 °C for 30 s, followed by 40 cycles of 95 °C for 5 s, and 60 °C for 30 s. Relative expression levels of Orco were calculated using the 2-ΔΔCt method (Livak and Schmittgen, 2001).

Statistics

All statistics were done with self-written python codes, SigmaPlot 12 (Systat Software), or JASP (versions 0.19.03 and 0.95.4, University of Amsterdam) with a significance level α = 0.05. Unless noted otherwise, data were checked for normal distribution with Anderson-Darling or Shapiro-Wilk tests and for equal variance with Levene’s test. If one of these assumptions failed, we used non-parametric tests, otherwise we used parametric tests.

To assess the statistical significance of observed differences in neuronal activity of the aligned data between ZTs, we applied the Wilcoxon signed-rank test to pairwise comparisons between the first low activity period and the subsequent high activity period and discarded the transient high activity at the beginning of most recordings.

For paired recordings (electrode reattachment, addition of DMSO, or cAMP/cAMP+OLC15 in the same animal, we used either paired student-t test or Wilcoxon signed-rank test, depending on normal distribution of data.

For other data, we used an ANOVA on ranks with either Tukey’s or Dunn’s post-hoc test for multiple comparisons as indicated.

ORN Model

In this study, we introduce a single-compartment Hodgkin-Huxley (HH) type model of the spontaneous activity of M. sexta ORNs (available on GitHub (Forlino et al., 2025)). The model contains 5 sets of ion channels that were previously identified in electrophysiological recordings (Dolzer et al., 2021): the fast sodium (Na+) and potassium (K+) channels responsible for the generation of spikes, low voltage-activated calcium channels (LVA) responsible for the initiation of bursts, calcium-gated potassium channels (BK) responsible for the termination of bursts, and the Orco channel, which is a voltage-independent non-specific, cAMP-gated channel. A schematic of the model is provided in Figure 2A. We modeled Orco as a ZT-dependent conductance to represent the effect of the oscillations in cAMP concentration that were identified in both hawkmoth and cockroach antennae (Schendzielorz et al., 2014, 2015, 2012) (Figure 2B, C). For simplicity we chose to use a sine function to directly represent the oscillations in Orco conductance, but similar results can be achieved by modeling it as an output of a Goodwin model of the circadian clock via cAMP concentration.

The conductance-based model of a simplified hawkmoth ORN.

(A) Schematic of the model containing a voltage gated sodium channel (Na+), a voltage gated potassium channel (K+), a low voltage activated calcium channel (LVA), a calcium and voltage activated potassium channel (BK), a leak channel, and a cAMP-gated leaky non-specific cation channel (Orco). The Orco channel is represented as a linear conductance whose value oscillates with a sinusoidal shape as a function of the ZT. The circadian rhythm in the Orco conductance is hypothesized to result from a cAMP-dependent increase in channel open-time probability (as found in Drosophila Orco: (Getahun et al., 2013)) via a circadian oscillation of antennal cAMP concentrations (Schendzielorz et al., 2015). This oscillation is driven by the circadian clock and serves as an input; the internal mechanism of the clock is not modeled explicitly. Arrows indicate the flow of ions through the channels. (B) Parametrized circadian oscillation in Orco conductance introduced as an external input to our model, representing the gating of Orco by cAMP whose concentration oscillates on a circadian time scale. The times of minimum and maximum cAMP concentration are indicated by colored arrows. (C) Parametrized IV curve of the Orco channel showing a linear behavior with the slope depending on the cAMP concentration. Line colors correspond to the arrows in B. (D) Markov-chain representing all possible states and transitions for each type of ion channel included in the model. The per capita transition rates (α and β) depend on membrane potential and in the case of the BK channel also on the intracellular Ca2+-concentration. The “N” state vectors contain the population of channels in each state “n”. Directed edges are numbered in red. The only conducting state of each ion channel, representing the condition where all its gates are open, is shown in green.

Each type of ion channel possesses one or more gates that can open and close individually with transition rates α and β for opening and closing, respectively, of activation gates (n, m) and inactivation gates (h). As a result, each possible combination of open and closed gates defines a possible state in which each ion channel can be found. A schematic of the Markov chain depicting all possible states and transitions for each ion channel is shown in Figure 2D. Each ion channel can conduct current only in the state where all its gates are open.

First, we constructed a deterministic 22-dimensional model to serve as the mean field of the Langevin model following the methodology proposed by Pu and Thomas (2020). Here, one dimension tracks the membrane potential, another the intracellular Ca2+ concentration, and the remaining 20 dimensions track the number of channels that are in a specific state of the Markov chains. Considering the conservation of the total number of ion channels, each type of ion channel required one less dimension than the total possible states.

The complete Langevin model adds to the mean field an independent noise source for each transition in the Markov chain. These noise sources are biophysically motivated since the transition between ion channel states is a stochastic process, with the probability of transition determined by the kinetics of each kind of ion channel. The variability in the spiking behavior of a neuron emerges from this non-deterministic gating of individual ion channels. The detailed mathematical description of the model can be found in the Supplementary Material.

Results

We tested the hypothesis that sensory neurons, such as insect ORNs, perform active, anticipatory sensing based upon an endogenous plasma membrane clock with the pacemaker channel Orco as the core element. We focused on pheromone-sensitive long trichoid sensilla of male M. sexta hawkmoth antennae which express daily rhythms in pheromone sensitivity and temporal resolution that are controlled by a circadian clock. We performed minimally invasive in vivo tip recordings over the course of several days to search for daily rhythms in spontaneous spiking activity as a measure of endogenous membrane potential oscillations.

Pheromone-sensitive hawkmoth ORNs exhibit Orco-dependent circadian rhythms in spontaneous spiking activity

The ORNs that innervate the pheromone-sensitive long trichoid sensilla on male M. sexta antennae display spontaneous spiking patterns of irregular bursts with interspersed single action potentials in short-term recordings (Dolzer et al., 2001). In the present study we documented the same pattern in long-term recordings over the course of 2-7 days. This enabled us to search for daily and circadian rhythmicity in spontaneous ORN spiking activity in the absence of pheromone stimulation.

RAIN analysis of the spontaneous spiking patterns revealed significant daily rhythms in 7 of 11 animals under long-day conditions (17:7 h LD) (Figure 3A). The maxima in spontaneous activity occurred mostly during the dark phase when nocturnal hawkmoths are active (Figure 3A (top), Bi light blue arrow), and minima occurred during the light phase when hawkmoths rest/sleep (Figure 3A (bottom), Bi dark blue arrow). High spiking activity in the dark phase had a mean maximum frequency of ∼3 Hz (delta range: 0.5 – 4 Hz) (Figure 3Bii). In contrast, the minimum spiking frequencies during the hawkmoth’s resting phase were below 0.5 Hz (Figure 3A (bottom)).

Spontaneous activity of pheromone-sensitive olfactory receptor neurons (ORNs) of male M. sexta shows Orco-dependent circadian modulation.

(A) Example tip recordings of the spontaneous activity of one long trichoid sensillum during a period of high (top, average frequency ∼ 3 Hz) and low (bottom, average frequency ∼ 0.2 Hz) activity. (Bi) Spontaneous spike frequency of one long-term ORN recording during 17:7 light-dark cycles (same experiment as in A, colored arrows indicate respective Zeitgeber times (ZT)), indicated by the black-white bar at the top, increased during each activity phase and was low during the resting phases. Spike frequencies were averaged for each 1 h bin. (Bii) Mean spontaneous ORN spike frequency across all LD animals in 1 h bins. Data from the individual in Bi is highlighted in red. Recordings had different lengths, therefore the number of recordings used for the element-wise mean for each 1 h bin decreased with the time since the start of the recording, indicated in the top panel. (Biii) Mean spontaneous ORN spike frequency across all LD animals in 1 h bins during the first 48 hours revealed circadian activity. (Ci) The spontaneous spike frequency of one long-term ORN recording in constant darkness (DD, the expected times of lights on are represented by grey bars above) exhibited a circadian pattern. The peak activity shifted in DD due to the animal’s endogenous, free-running circadian period τ = 21.27 h. (Cii) Mean spontaneous ORN spike frequency across all DD animals in 1 h bins. Data from the individual in Ci is highlighted in red. (Ciii) Mean spontaneous ORN spike frequency across all DD animals in 1 h bins during the first 48 hours after aligning the time to the first maximum in spontaneous activity (see Methods) revealed circadian activity. (Di) The spontaneous spike frequency of one long-term ORN recording in constant darkness with infusion of the Orco antagonist OLC15 (orange frame) dissipated the circadian rhythm of spontaneous activity. (Dii) Mean spontaneous ORN spike frequency across all OLC15 animals in 1 h bins. Data from the individual in Di is highlighted in red. In contrast to LD and DD conditions, the spike frequency decreases over time with prolonged exposure to OLC15. (Diii) Mean spontaneous ORN spike frequency across all OLC15 animals in 1 h bins (mean ± SD) during the first 48 hours after aligning the time to the first maximum in spontaneous activity (see Methods). In contrast to LD and DD conditions, the circadian change in spontaneous spiking frequency disappeared. (E) The slopes of the linear fits to the binned spiking activity of each individual animal in the three different conditions (see Methods). Each dot indicates the slope for one experiment. The line indicates the mean. The slope for OLC15 is significantly different from control LD.

Since ORN firing patterns had a clear daily rhythm in LD both in individual animals (Figure 3Bi) and averaged across animals (Figure 3Bii, Biii), we examined whether these rhythms depended on cycling light cues or whether they persisted in constant darkness (DD), as expected for an endogenous, circadian clock-driven rhythm. In DD, the spontaneous ORN firing pattern remained rhythmic (RAIN: 8 of 12 animals), albeit with less robustness: The single peak of high spiking activity that was observed at the same ZT across several days in LD divided into multiple bouts of high spiking activity in the activity phase in 8 of 12 animals (Figure 3Ci). Also, in DD conditions, the high spiking activity phase shifted into the subjective day due to the individual endogenous circadian periods of about 23.5 ± 2.8 h (21.27 h in Figure 3Ci). Using pheromone exposure as additional Zeitgeber also did not synchronize the animals (Figure 3 – Supplement). Thus, averaging the instantaneous spiking activity across all DD animals did not depict clear circadian rhythms (Figure 3Cii). Therefore, we phase-aligned the mean frequency of spontaneous activity (see Methods). After alignment, circadian rhythms were visible across DD animals (Figure 3Ciii), comparable to the daily rhythms in LD (cf. Figure 3 Biii and Ciii).

Next, we examined whether Orco, a leaky, non-specific cation channel, is the dominant depolarizing pacemaker current of ORNs that drives circadian rhythms in spontaneous spiking activity. Infusion of the Orco antagonist OLC15 into the sensillum lymph obliterated circadian rhythms and attenuated the spontaneous activity in several, but not all experiments (RAIN: 2 of 12 animals expressed circadian rhythms) (Figure 3Di). This attenuation resulted in a linear decrease in spiking activity over several days (Figure 3Dii, Diii). Even after phase-aligning the individual experiments, no circadian rhythm was present in the average across all OLC15-treated animals (Figure 3Diii).

Because OLC15 is not membrane-permeable on its own it was infused with low concentrations (0.05%) of DMSO through the recording electrode into the sensillum lymph and, therefore, effectiveness increased over time. Hence, we fit the spiking activity of each animal and in each condition with linear regression lines over the whole recording time to compare the slopes of the fits (Figure 3E). In LD, the mean ± SD slope was 0.004 ± 0.006 Hz/h, indicating that the average spiking activity remained similar throughout the duration of the recording. In DD, the slope was −0.003 ± 0.018 Hz/h and was not different from LD. In contrast, the mean slope during OLC15 treatment was −0.007 ± 0.022 Hz/h, significantly smaller than in LD (ANOVA on ranks with Dunn’s post-hoc test; H(2) = 7.681, p = 0.021). The negative slope corroborated the finding that spiking activity decreased over time in the presence of the Orco blocker.

Activity rhythms between hawkmoths desynchronize and have only weak phase coupling in the absence of pheromone stimulation, even in LD cycles (Ghosh et al., 2024). Therefore, we first determined the percentage of animals that expressed circadian rhythmicity (period of 24 ± 4 h) for each attribute with RAIN (Figure 4A). In LD, all attributes except #AP per burst and intra-burst frequency showed daily modulation in most animals. Circadian rhythms were still present in more than half of the animals in DD for mean spiking frequency, mean burst frequency, and mean event frequency. In contrast, the attributes of only a few animals remained rhythmic with the Orco channel blocker OLC15.

Blocking Orco removed circadian modulation of spontaneous ORN spiking patterns.

RAIN analyses demonstrated significant circadian rhythmicity for most animals in most attributes of the spontaneous spiking activity in LD and DD conditions but not in OLC15 (A). Values and background colors indicate the percentage of animals that expressed significant circadian rhythmicity. Circadian differences of attributes were further quantified between the time windows of low vs. high activity (Wilcoxon signed-rank test, α = 0.05; B-H). Significant differences occurred mostly in LD (white boxes) and DD (gray boxes) conditions, but never when Orco was blocked with OLC15 (orange boxes). In both LD and DD, the mean spiking frequency was increased during high activity (Bi, Bii). Only in LD but not in DD, the mean burst frequency (Ci, Cii) increased, whereas the relative number of spikes belonging to a burst (Di, Dii) decreased significantly during the high activity period. Both in LD and DD the mean event frequency (Fi,Fii) increased significantly. Mean burst duration (Ei, Eii), mean number of spikes per bust (Gi,Gii), and mean intra-burst intervals (Hi, Hii) did not differ between low vs. high spiking activity in LD and DD. When blocking Orco, the attributes for low vs. high activity did not differ for any of the attributes tested (Biii-Hiii). In DD and OLC15conditions, activity phases were aligned as described in the Methods; thus, low activity in DD was at subjective ZT 19 and in OLC15 at subjective ZT 13, high activity was in both cases at subjective ZT 24. ZTs for low and high activity in LD were 10 and 0, respectively.

To further quantify the circadian differences, especially of the desynchronized animals, we compared different attributes of spontaneous activity between periods of high and low activity for DD conditions after phase alignment (see Methods) instead of comparing fixed ZT or circadian time (CT) intervals. Despite the high variability in ORN spiking between individuals, the mean frequency of spontaneous ORN activity was significantly higher during the high activity phase than during low activity in both LD and DD recordings (Figure 4Bi, Bii). In LD, but not in DD, the frequency of bursting increased significantly during the high activity phase (Figure 4Ci, Cii). The percentage of spikes that are part of bursts was less variable and lower during the high activity phase in LD but not significantly different in DD (Figure 4Di, Dii). The mean burst duration was less variable during the high than the low activity phase in LD, but not significantly shorter in any condition (Figure 4Ei, Eii). Furthermore, both in LD and DD, the mean inter-event frequencies were significantly higher during the high activity phase (Figure 4Fi,Fii), indicating a significantly higher overall spontaneous spiking activity during the hawkmoth’s activity phase. The mean number of spikes per burst, and intra-burst frequencies did not differ significantly between high or low activity periods in both LD and DD conditions (Figure 4Gi, Gii, Hi, Hii). The addition of Orco antagonist OLC15 deleted any significant differences of spontaneous activity found in LD and DD (Figure 4Biii-Hiii). In conclusion, the mean frequency of spontaneous spiking and the frequency of bursting expressed circadian modulation, and are both most likely controlled via a circadian clock that involves Orco.

Orco imposes circadian modulation on the ultradian rhythms in spontaneous ORN activity

After establishing that Orco influences the circadian changes in ORN spiking activity between the animal’s resting and activity phase, we examined if Orco also influences ORN spiking rhythms on the largely different ultradian time scales in the frequency range between 0.01-1000 Hz. In all conditions, two broad bands of ultradian instantaneous frequencies were evident: one upper band of >10 to ∼100 Hz (beta/gamma range) and a lower band between 0.1 to ≤10 Hz (delta/theta/alpha range) (Figure 5). The upper band primarily represents the frequencies of spikes within a burst, while the lower band represents single spikes and frequencies between bursts.

Blocking Orco removed circadian regulation of ultradian rhythms in the spontaneous ORN spiking pattern.

Heat map of the instantaneous frequencies (1/ISI) over two consecutive days of long-term tip recordings of one pheromone-sensitive long trichoid sensillum under different conditions (LD (A), DD (B), OLC15 (C)) in each panel. Representative recordings from one animal in each panel. Pixel color indicates the counts of instantaneous frequencies in that respective bin. The points fall mostly in a band of high frequencies (>10 – ∼100 Hz) and low frequencies (0.01 – ∼10 Hz), with the high-frequency band representing the instantaneous frequencies of spikes within a burst and the low-frequency band the instantaneous frequency between bursts. (A, B) The high frequency band indicates daily and circadian modulation of frequency prevalence. In addition, the low frequency band (<0.01 Hz to ∼ 10 Hz) also displayed circadian modulation of the frequency composition. (C) Infusion of OLC15 deleted the circadian modulation in both frequency bands but not the ultradian rhythm of frequency prevalence.

In LD, two types of daily/circadian modulation of the ultradian frequencies occurred (Figure 5A). The upper frequency band was mostly modulated in prevalence (its frequency of occurrence) but not in its frequency range. The lower band was modulated both in prevalence and frequency range, which resulted in a sinusoidal change in the mean of the frequency band over 24 h. During the hawkmoth’s activity phase, higher frequencies of up to ∼10 Hz occurred, while frequencies dropped below 1 Hz during the resting phase. Because of this circadian increase in the lower frequency band, it merged with the high frequency band when both bands had maximum prevalence during the activity phase around ZT 0.

The circadian modulation of both frequency bands was maintained in constant darkness (Figure 5B) but phase-shifted with respect to the projected ZT due to the endogenous circadian period of each individual. Again, this indicates that a clock is modulating the spontaneous ORN spiking activity over the course of a day. The addition of the Orco antagonist OLC15 deleted any circadian modulation of either frequency band. Interestingly, the instantaneous frequencies in both bands still showed correlated changes in prevalence and merging, but now at ultradian periods only.

Additional Fourier analysis (see Methods) of spontaneous spiking activity over the course of several days revealed faster ultradian and slower infradian rhythms, ranging from 0.6 h to 38 h, in addition to the circadian (24 ± 4 h) rhythm (Figure 6Ai, Bi, Ci). Some rhythms (factorials of 24 h, like 12 h, 6 h, etc.) appeared to be harmonics of the circadian rhythm, which were not excluded in our analysis. Most animals expressed ultradian rhythms of 2-4 h in LD as well as DD conditions. Fourier confirmed the decrease of animals with circadian spiking activity in OLC15 (58.3% in DD, 33.3% in OLC15). Please note that discrepancies in the number of animals with circadian rhythms between RAIN and Fourier/wavelet analyses are due to the fundamentally different sensitivities in methodology. In brief, Fourier analysis assumes a stationary sinusoidal component, whereas RAIN as a rank-based umbrella test detects consistent raise-fall patterns, including asymmetric rhythms. Nonetheless, both methods confirmed a dramatic decrease in animals with circadian spiking patterns in OLC15. In addition, 33.3% of animals expressed infradian spiking patterns (up from 0% in DD) with OLC15. Ultradian periods could be detected in all animals in all conditions but their distribution did not differ between conditions (1-way ANOVA; F(2, 117) = 0.87, p = 0.421). These results were corroborated by continuous wavelet analysis, which provided time-resolved spectral power across different rhythmic periods (Figure 6Aii, Bii, Cii). In control LD and DD conditions, wavelet power for ultradian periods increased during times of elevated firing rate and decreased during periods of low activity (Figure 6Aii and Bii, middle and bottom panels). This dynamic pattern indicates that ultradian rhythms were not constant over time but were instead modulated with a circadian rhythm, becoming more prominent during circadian peaks in neuronal activity. Such circadian gating of ultradian power was clearly visible in the wavelet spectrograms as alternating bands of higher and lower ultradian power aligned with daily activity cycles. Following Orco blockade with OLC15, this modulation was disrupted (Figure 6Cii middle panel): although ultradian components persisted (Figure 6Cii bottom panel), they no longer showed systematic variation with the circadian cycle, suggesting a decoupling of ultradian rhythms from circadian control.

Blocking Orco affected ultradian and infradian frequencies in the spontaneous ORN firing patterns.

The Fourier analysis of spontaneous spiking activity revealed rhythms with ultradian (<20 h; dark grey), circadian (20-28 h; white), and infradian periods (>28 h; light grey) in LD (Ai), DD (Bi), and OLC15 (Ci). Each line represents one animal where each triangular marker is at the local maximum in the frequency spectra obtained for that animal. The histograms above illustrate how often specific periods of spiking activity rhythms occurred averaged over all animals. Circadian rhythms were detected in LD (N = 5 of 11), DD (N = 7 of 10), and OLC15 (N = 4 of 12). Wavelet analysis of LD (Aii), DD (Bii), and OLC15 (Cii) confirmed the occurrence of multiscale periods. Example plots for one animal each, the same animals as in Figure 5. For each panel, the top plot depicts the mean firing frequency, the middle plot the wavelet power in the same period range as panels i, and the bottom plot the wavelet power for infradian periods up to 7 h on a log scale to highlight infradian time scales. Gray dashed lines indicate cone of uncertainty.

Orco expression is not under the control of the molecular TTFL clockwork

So far, our results indicated a prominent role of Orco in the circadian modulation of ORN spiking activity. Since it is the general opinion that the master circadian clock is the TTFL clock that dominates all physiological and behavioral circadian rhythms via transcriptional control, we examined whether Orco transcript abundance shows a daily rhythm, as previously found for clock genes in hawkmoth antennae (Schneider et al., 2025; Schuckel et al., 2007). However, in contrast to the circadian clock gene tim, qPCR of whole male hawkmoth antennae across different ZTs revealed no significant daily rhythm of Orco transcript abundance (Figure 7; ANOVA on ranks, H(5) = 6.91, p = 0.227). Thus, we suggest that Orco function is controlled via a post-translational circadian clockwork mechanism that is associated directly with the plasma membrane (Stengl and Schneider, 2024; Stengl and Schröder, 2021).

Orco is not under control of the TTFL-based molecular circadian clock.

Relative expression levels of Orco and tim mRNA from male hawkmoth antennae raised under LD conditions. Orco expression levels did not differ significantly between ZTs (ANOVA on ranks, H(5) = 6.91, p = 0.227: N = 4 per ZT). tim expression levels changed significantly throughout the day (1-way ANOVA, F(5, 18) = 6.215, p = 0.002) Red lines indicate mean, tim served as positive control.

Circadian changes in Orco conductance are sufficient to modulate spiking activity in a computational model of a hawkmoth ORN

The maximum conductance of Drosophila Orco as a leaky ion channel is modulated by cAMP (Getahun et al., 2013; Martín et al., 2001; Sargsyan et al., 2011; Wicher et al., 2008). In hawkmoth antennae, cAMP levels oscillate throughout the day and are involved in sensitizing the ORNs to pheromone (Dolzer et al., 2021; Flecke et al., 2010; Flecke and Stengl, 2009; Schendzielorz et al., 2014, 2015). Hence, cAMP modulation of Orco conductance is a putative mechanism for the Orco-dependent circadian modulation of ORN spiking activity, integrating Orco in a circadian PTFL-based membrane clockwork (Stengl and Schneider, 2024).

To test the hypothesis that Orco is a cAMP-dependent, leaky ion channel, we constructed an ORN model (Methods and Supplementary Material) and tuned the parameters to provide a quantitative match with the biological data. The developed model provides a unified description of experimental observations. For instance, the ORN model yields irregular spiking activity with a bimodal distribution, consisting of both bursts and isolated spikes, successfully reproducing the experimental recordings (Figure 8A). This suggests a unified mechanism underlying the initiation of both phenomena: If the pacemaker-driven depolarization reaches the Na+ threshold, a firing process begins, and at this point randomness determines if other slower ion channels responsible for bursting are sufficiently activated to sustain a burst of several spikes.

The model adequately reproduces the firing pattern of the biological ORN.

(A) The model activity (bottom trace) displays both isolated spikes and bursts of variable length as in the original recording (top trace). The simulated trace shows the membrane potential (an intracellular variable), while the biological data were obtained with extracellular electrodes. (B) Semi-logarithmic plot of the distribution of spikes per burst of an experimental LD recording (top) and simulated model (bottom) of spontaneous action potential activity of pheromone-sensitive neurons. The experimental distribution corresponds to the 1-hour period of highest firing frequency in the recording; the simulated distribution was obtained from a separate 1-hour simulation at fixed high Orco conductance. Both the model and the recordings display exponential decrease of the count’s density as a function of the burst length, shown as a linear decrease in the logarithmic scale. (C) Comparison of experimentally recorded and model-predicted circadian modulation of spiking attributes. Top panels show mean spike frequency (Ci, experimental; Cii, model), whereas bottom panels show the percentage of spikes belonging to bursts (Ciii, experimental; Civ, model). The simulated model reproduced the circadian dynamics observed in the experimental recordings, demonstrating good agreement between model and biological data. (D) Heat map of the simulated model, x-axis shows time (days), y-axis shows instantaneous frequency (Hz), and pixel color indicates count density on a logarithmic scale. The instantaneous frequencies can be roughly divided in a band of high frequencies (top, ∼50 – 80 Hz) and low frequencies (bottom, ∼0.1 – 5 Hz), with the high-frequency band representing the instantaneous frequencies of spikes within a burst and the low-frequency band the instantaneous frequency between bursts. (Di) Simulated spikes with circadian regulation by the Orco channel; the inter-burst frequencies varied widely with circadian rhythmicity while the frequencies within burst remained approximately constant. The simulated results also reproduced the “merging” effect between the two bands when the bottom one approaches the top one. (Dii) Simulated spikes without circadian regulation of Orco had no circadian oscillations.

The model successfully reproduces the circadian modulation of spiking activity: the maximum firing rate occurs when the Orco conductance is maximal, whereas the maximum percentage of spikes that are part of bursts correlates with the lowest Orco conductance, resulting in antiphase oscillations, as were observed in the experimental recordings (Figure 8C).

The frequency prevalences in the model over several days clustered in two distinct frequency bands: an upper band that corresponds to the intra-burst frequencies and a lower band, corresponding to inter-burst frequencies (Figure 8D), in agreement with the biological experiments (Figure 5A, B). When circadian oscillations in Orco conductance were included, inter-burst frequencies showed pronounced circadian variations, whereas intra-burst frequencies remained relatively constant, reminiscent of the biological data. Notably, the simulation captured the experimentally observed "merging” effect, where the lower frequency band approaches the upper band during high-activity periods (Figure 8D).

This merging of spiking and bursting frequencies suggests that the system approached a bifurcation condition coming from the distribution of spikes per burst (Figure 8B), computed from the 1 h period of highest firing frequency in an experimental recording and from a separate 1 h simulation at fixed high Orco conductance. In both cases, the distribution followed a decreasing exponential trend – in contrast to the constant value or bounded range expected from a standard bursting neuron model – consistent with a system operating near a bifurcation point. The observation that the circadian modulation of spiking activity requires only one parameter, namely changes in the Orco conductance, to describe the circadian effects on the ORN spiking pattern indicates the crucial role of the Orco channel in circadian control.

cAMP regulates the circadian function of Orco

Our computational model predicted that the daily/circadian oscillations in cAMP levels (Schendzielorz et al., 2015) control the circadian modulation of spontaneous spiking activity of hawkmoth ORNs. To test this prediction in hawkmoth ORNs, we increased cAMP levels by infusing 8-Br-cAMP-AM through the recording electrode into the sensillar lymph (Figure 9) at ZT 1-3, when naturally occurring cAMP levels are lowest (Schendzielorz et al., 2015) to maximize the effect. This significantly increased the mean frequency of the spontaneous spiking activity (Figure 9A; Wilcoxon signed-rank test Z = −2.701; p = 0.004). To check whether this increase in mean frequency was caused by direct or indirect interaction of cAMP with the Orco channel, we infused the Orco antagonist OLC15 together with 8-Br-cAMP-AM through the same recording electrode. Blocking Orco prevented a significant change in the spontaneous spiking frequency via the cAMP analogue (Figure 9B; paired t-test t(9) = −1.446; p = 0.182). This further corroborates cAMP-dependent modulation of the circadian pacemaker channel Orco.

cAMP modulates open-time probability of Orco in M. sexta.

Perfusion of 8-Br-cAMP-AM significantly increased the spontaneous spiking activity at the end of the activity phase (ZT 1-3; A; Wilcoxon signed-rank test Z = −2.701; p = 0.004). However, activity did not increase when 8-Br-cAMP-AM was added together with the Orco blocker OLC15 (B; paired t-test t(9) = −1.446; p = 0.182). Electrodes were filled with sensillum lymph ringer containing 0.1% and 0.15% DMSO as control in A and B, respectively. Data from individual animals are connected by lines.

Discussion

Searching for daily and circadian rhythms of Orco-dependent membrane potential control in olfactory receptor neurons (ORNs) of M. sexta, in vivo long-term tip recordings of pheromone-sensitive long trichoid sensilla of male hawkmoth antennae were performed. We detected interlinked circadian and ultradian frequencies in the spontaneous spiking activity, indicative of underlying membrane potential oscillations in ORNs. We found that the olfactory receptor co-receptor (Orco) acts as a pacemaker channel that controls the daily/circadian modulation of the spiking activity without affecting the ultradian gamma frequency range that is relevant for the primary fast, phasic pheromone response (Dolzer et al., 2003; Nolte et al., 2016; Schneider et al., 2025). Furthermore, since OLC15-dependent block of Orco did not delete ultradian action potential activity, other additional pacemaker channels must be present in ORNs that possibly control different frequency ranges of ultradian membrane potential oscillations. Since qPCR experiments at different ZTs showed that Orco transcription was not directly controlled by the circadian TTFL clock present in hawkmoth ORNs, we concluded that the daily/circadian control of Orco function is obtained at the post-translational level. Our computational model predicted that cAMP-dependent post-translational modulation of Orco conductance is sufficient to account for our electrophysiological findings if cAMP levels vary in a daily/circadian rhythm. Daily rhythms in cAMP levels are confirmed in hawkmoth antennae (Schendzielorz et al., 2015). Furthermore, here we provided evidence that cAMP analogs Orco-dependently control spontaneous activity of hawkmoth ORNs. Thus, we suggest that, colocalized in a membrane signalosome in the ORN soma compartment, constitutively depolarizing Orco homomeric ion channels generate ultradian membrane potential oscillations via delayed activation of hyperpolarizing K+ channels (Figure 10).

A signalosome assembling an endogenous multiscale PTFL membrane clock in the soma of hawkmoth olfactory receptor neurons (ORNs) with 2nd messenger-dependent Orco as central control hub is proposed to mediate membrane potential rhythms.

Schematic of an antennal pheromone-sensitive ORN that is strictly compartmentalized by tight junctions (thick black borders) into three functionally separated membrane compartments: cilium, soma/dendrite, and axon. Soma/dendrite and axon are wrapped by supporting cells (grey background), ensuring common extracellular ionic concentrations for both membrane compartments. Thus, only the cilium is exposed to the sensillum lymph that contains high (∼200 mM) K+ concentration. The ciliary membrane comprises a specialized membrane signalosome devoted to highly sensitive pheromone detection. In contrast, a signalosome in the soma/dendrite controls the membrane potential and intracellular Ca2+ concentration. Sustained depolarization via opening of cation-permeable, homomeric Orco pacemaker channels in the soma are counteracted by delayed voltage-gated hyperpolarizing K+-channels, together generating ultradian oscillations in the membrane potential with interlinked oscillations in intracellular Ca2+. Circadian/daily rhythms in cAMP concentrations mediating Orco conductance, next to Ca2+, superimpose a circadian rhythm on prevalence of ultradian oscillations. Orco is not under direct circadian transcriptional control by the circadian TTFL clock in the nucleus that is entrained to the LD cycle, thus being part of a PTFL membrane clock entrained to other Zeitgebers. PTFL- and TTFL-clocks are coupled by not-yet identified mechanisms. LD: light dark cycle; OR: odor receptor, Orco: olfactory receptor coreceptor, PTFL: posttranslational feedback loop clock; TTFL: transcriptional-translational feedback loop clock; Vm: membrane potential; sine waves: oscillations.

The ultradian membrane potential oscillations express a daily/circadian rhythm with higher frequencies during the hawkmoth’s activity phase at night compared to the rest phase during the day. Thus, we suggest that the soma signalosome constitutes an endogenous, plastic, post-translational feedback loop (PTFL) clockwork, controlling membrane potential oscillations with Orco as 2nd messenger-dependent control hub that is coupled to, but not commanded by, the circadian TTFL clock in the nucleus.

The expression of a 2nd messenger-dependent leaky pacemaker channel in ORNs, which does not require daily transcription, would constitute an economic and rapidly adjustable mechanism of active sensing. As part of a PTFL membrane clock, Orco could orchestrate endogenous membrane potential oscillations that are tuned to detect and resonate with behaviorally relevant temporal patterns of Zeitgeber signals, such as the circadian and ultradian pheromone fluctuations in the insect’s environment that are detected in the cilia of ORNs (Stengl and Schneider, 2024; Stengl and Schröder, 2021). The present study provides first evidence for a systems view of biological timing that involves multiscale PTFL membrane clocks linked to the TTFL clock in insect sensory neurons, instead of the prevailing hierarchical view that assumes all endogenous circadian oscillations are outputs of the genetic circadian TTFL clock (Stengl and Schneider, 2024; Stengl and Schröder, 2021).

Orco in the ORN soma compartment acts as a pacemaker channel that controls the membrane potential oscillations underlying spontaneous spiking activity

The resting membrane potential of most silent neurons lies close to the negative equilibrium potential for K+ ions, mainly due to potassium leak channels and the driving force by the 10x higher intracellular K+ concentration (Goldstein et al., 2001; Hille, 2001; Patel and Honoré, 2001; Talley et al., 2001). Thus, spontaneously active neurons require the expression of Na+ and/or Ca2+ permeable leak or pacemaker channels that open at the hyperpolarized resting membrane potential, thus depolarizing the neuron via influx of cations along their concentration gradients (Bose et al., 2014; Cochet-Bissuel et al., 2014; Das et al., 2016; Golowasch et al., 2017; Lüthi and McCormick, 1998; Ratliff et al., 2021; Robinson and Siegelbaum, 2003; Sharma et al., 2023). When the pacemaker-driven depolarization reaches spike threshold, voltage-gated Na+ and K+ channels in the axon promote spiking. Furthermore, voltage-dependent Ca2+ channels allow for depolarization-dependent Ca2+ influx. Voltage- and Ca2+-gated hyperpolarizing K+ channels then repolarize the neuron to negative membrane potentials, re-starting the cycle. Furthermore, different fine-tuned mechanisms of homeostasis re-establish intracellular low (pico- to nanomolar range) Ca2+ concentrations, preventing necrosis. Thus, membrane potential oscillations with interlinked oscillations in intracellular Ca2+ levels drive spontaneous spiking activity, which depends on the type of pacemaker channel and antagonistic ion channels present.

In insect ORNs, at least two types of pacemaker channels exist: the highly conserved, slow, leaky Orco, which we focus on in this study, and an inversely voltage-dependent HCN type channel (Butterwick et al., 2018; Dolzer et al., 2021; Kodirov, 2022; Lee and MacKinnon, 2017; Mandala et al., 2025; Nolte et al., 2016, 2013; Sato et al., 2008; Stengl and Funk, 2013; Stengl and Schneider, 2024; Stengl and Schröder, 2021; Wicher et al., 2008; Wicher and Miazzi, 2021). When expressed in vitro in heterologous expression systems, Drosophila as well as Manduca Orco homo- and heteromers assemble into leaky, non-specific cation channels with a reversal potential around 0 mV, being constitutively permeable for Na+, K+, and Ca2+ (Nolte et al., 2016, 2013; Sato et al., 2008; Stengl and Funk, 2013; Wicher et al., 2008; Wicher and Miazzi, 2021). Thus, while this study did not aim to unravel the overall generation of spontaneous activity in ORNs, here, we examined whether the leak channel Orco in hawkmoth ORNs is the core pacemaker channel responsible for membrane potential oscillations that drive spontaneous circadian activity (Stengl, 2010; Stengl and Schneider, 2024).

Orco homo- or heteromers are likely to have different localizations and functions in the distinct, sealed-off membrane compartments of ORNs that face different ionic environments: the cilium (also termed “outer dendrite”), and the soma / (“inner”) dendrite (Stengl and Funk, 2013; Stengl and Schröder, 2021). We propose for Manduca, that homomeric Orco pacemaker channels, as core element of a PTFL membrane clock, are restricted to the soma/dendrite compartment of ORNs, where they face the same extracellular cation concentrations as other brain neurons, with low K+, high Na+, and high Ca2+ concentrations (Stengl and Schneider, 2024; Stengl and Schröder, 2021). In contrast, we expect Manduca Orco heteromers to serve as obligatory chaperons to localize ORs to cilia and maintain them in the ciliary membranes, as proven in Drosophila (Benton et al., 2006). Since the ciliary compartment faces very high concentrations of K+ in the sensillum lymph, matching intracellular K+ levels (Kaissling and Thorson, 1980; Steinbrecht, 1992), we propose that Orco channels in the ciliary membrane remain closed. If Orco homo- or heteromers in cilia would constitute leaky pacemaker channels, their K+ permeability would clamp the ciliary membranes at 0 mV, due to the high extracellular K+ concentration, strongly reducing the driving force of pheromone-dependent receptor currents, severely compromising pheromone sensitivity (Stengl and Funk, 2013). Accordingly, it was shown previously in hawkmoth ORNs that Ca2+ channels, most likely of the TRP ion channel family, serve as transduction channels for the primary events of the highly sensitive pheromone transduction, and not a non-specific cation channel like Orco (Gawalek and Stengl, 2018; Nolte et al., 2016, 2013; Stengl, 1994, 1993).

In support of the specific localization of Orco homo- or heteromers, Orco antisera labeled Orco in two of the three sealed off ORN membrane compartments in both Drosophila (Benton et al., 2006) and Manduca (Nolte et al., 2016): in the membrane of the soma/dendrite compartment and in the membrane of the cilium, but not in the axons. In contrast, antisera against an OR in Drosophila detected ORs accumulated only in ciliary membranes but not in the other compartments of ORNs (Benton et al., 2006). Thus, the localization of OR-Orco heteromers in cilia confirm that Orco is an obligatory coreceptor and chaperone of ORs localizing and maintaining the OR-Orco complexes in the ciliary membrane (Benton et al., 2006). Since antisera against Manduca ORs are not available, we do not know for hawkmoths whether the pheromone-sensitive OR-Orco heteromers are located to ciliary membranes, while Orco homomers accumulate in the soma/dendrite compartment (Fandino et al., 2019). Nevertheless, Orco is likely to assemble as homomeric complexes in wild-type hawkmoth ORNs because Orco expression starts during development, coinciding with the beginning of spontaneous activity, well before ORs are present. Furthermore, the Orco agonist VUAA1 activates Orco in primary cell cultures of maturing ORNs before they become pheromone sensitive (Nolte et al., 2016; Schweitzer et al., 1976).

In summary, we propose that Orco homomers are specifically located to a signalosome in the soma/dendrite compartment of ORNs (Figure 10), controlling the ORN’s membrane potential. Because the Orco-dependent ultradian and circadian membrane potential oscillations are maintained under constant conditions, they are generated by an endogenous multiscale clock. Thus, we predict that a distinct signalosome with Orco homomers in the soma, but not in the cilia, serves as endogenous self-assembling multiscale PTFL membrane clock for the circadian control of the membrane potential. We hypothesize that this distinctly located PTFL membrane clock is coupled to, but not dominated by, the TTFL clock in the hawkmoth ORNs to allow for fast, plastic regulation of the membrane potential (Schuckel et al., 2007; Stengl and Schneider, 2024).

Orco is suggested to be a second messenger-dependent control hub of a PTFL membrane clock that regulates circadian membrane potential oscillations in hawkmoth ORNs

Focusing on the signalosome in the ORN soma compartment, we predict that Orco is a 2nd messenger-dependent control hub of a somatic multiscale PTFL clock. In accordance with this hypothesis, different 2nd messenger cascades regulate Drosophila Orco via a Ca2+/calmodulin binding site and via phosphorylation/de-phosphorylation at multiple sites that are conserved between species (Cao et al., 2016; Getahun et al., 2016; Guo and Smith, 2017; Mukunda et al., 2016, 2014; Sargsyan et al., 2011). In contrast to our current focus of Orco homomer function in Manduca, focus and data interpretation in Drosophila lies on the function of Orco heteromers. In Drosophila, Orco’s Ca2+/calmodulin binding site controls the localization of the OR-Orco complex in the ciliary membrane of ORNs, provides for Ca2+-dependent negative feedback control of Orco conductance, and depends on previous odor stimulation (Bahk and Jones, 2016; Cao et al., 2016; Guo and Smith, 2022; Jain et al., 2021). After phosphorylation of Orco by protein kinase C (PKC), Drosophila Orco directly or indirectly becomes cyclic nucleotide-dependent, with cGMP and cAMP increasing Orco-dependent depolarization and Ca2+ influx (Getahun et al., 2013; Wicher et al., 2008). However, it is not yet resolved in Drosophila in which of the different sensilla compartments and at what time of day Orco homo- or heteromers are controlled via 2nd messengers and how this control is regulated.

Also, in Manduca the 2nd messenger-dependent modulation of Orco is not yet understood. However, due to Orco’s highly conserved structure we assume Ca2+/calmodulin-, PKC-, and cyclic nucleotide-dependent control of hawkmoth Orco localization and/or open-time probability, likely similar as described for Drosophila Orco. It is very likely that the conserved structure of Orco with its manifold 2nd messenger-modulation sites designate it as a hub for intersecting 2nd messenger control in different insect species. Since blocking Orco in M. sexta decreased the spontaneous spiking activity in ORNs and deleted its circadian modulation, Orco is apparently the main target for circadian clock control modulating circadian membrane potential rhythms. Furthermore, the ultradian rhythms in spontaneous activity appear to be controlled by additional pacemaker channels that are not focus of the current study. Since Orco transcript abundance was constant throughout the day, Orco is not controlled directly by the circadian TTFL clock. Rather, open-time probability and/or membrane localization of Orco appear to be controlled in a circadian manner via post-translational mechanisms. Because daily rhythms in cAMP levels were found in hawkmoth antennae (Schendzielorz et al., 2015), our modeling predicted that the most parsimonious explanation for Orco’s circadian control of the membrane potential is Orco’s regulation via daily rhythms of cAMP levels. In support of this hypothesis, we demonstrated that the significant increase of spontaneous spiking activity caused by 8-Br-cAMP application is canceled with co-application of the Orco blocker OLC15. Furthermore, our modeling studies showed that this post-translational circadian control of Orco via daily/circadian levels of cAMP is sufficient to explain our physiological findings. Thus, we hypothesize that a PTFL membrane clock (Figure 10), linked to the TTFL nuclear clock, determines the circadian oscillations in pheromone sensitivity and temporal resolution of ORNs via cAMP- and Orco-dependent membrane potential control (Dolzer et al., 2021, 2008; Flecke et al., 2010, 2006; Redkozubov, 2000; Schendzielorz et al., 2015, 2012; Ziegelberger et al., 1990).

Our novel hypothesis predicts that a signalosome in the soma membrane of hawkmoth ORNs assembles a circadian PTFL membrane clock controlling daily/circadian rhythms in the membrane potential and Ca2+ concentration, comprising Orco as central control hub. Via unknown mechanisms that still need to be resolved, circadian oscillations in cAMP levels are generated in ORNs that are linked to Orco. As is characteristic for a clock (Stengl and Schneider, 2024), the PTFL membrane clock comprises positive feedforward elements that upregulate the ORN’s depolarization and intracellular Ca2+ levels, which are then sequentially downregulated via delayed negative feedback elements/mechanisms. Thus, the PTFL clock generates oscillations in the membrane potential coupled to strictly compartmentalized oscillations in intracellular Ca2+ levels. This predicted soma signalosome comprises the pacemaker channel Orco as a central hub that is controlled via changes in voltage, Ca2+, and cAMP levels that can be affected by multiple parallel pathways in the ORN. We propose a parsimonious model of the PTFL clock, generating multiscale membrane potential- and Ca2+ oscillations linked to cAMP oscillations. Since it is a PTFL clock, it does not require circadian degradation and circadian transcription/translation of its central constituents. Nevertheless, we predict coupling between the multiscale PTFL membrane clock and the TTFL circadian clock in the ORN’s nucleus to obtain stably synchronized circadian rhythms. As likely mechanism of coupling we predict that Ca2+- and cAMP-dependent kinases interlink both types of circadian clocks, thereby obtaining robust and at the same time flexible interlinked cellular rhythms (Figure 10). This system of interlinked circadian clocks are hypothesized to determine changing physiological setpoints of the ORN during circadian sleep-wake cycles (Stengl and Schneider, 2024).

Circadian and ultradian rhythms in spontaneous ORN activity resemble the physiologically relevant time scales for active sensing

Moths have daily rhythms in pheromone sensitivity and temporal resolution controlled via antennal circadian clocks (Flecke et al., 2010; Flecke and Stengl, 2009; Merlin et al., 2007, 2006; Schendzielorz et al., 2015; Schuckel et al., 2007; Silvegren et al., 2005). The increased pheromone sensitivity and temporal resolution during the activity phase enhances the male’s chance to detect and locate a calling female. In response to brief pheromone pulses, independent of Orco, pheromone concentration is encoded in the frequency of the first, phasic burst of spikes of the characteristic triphasic spiking sequence of ORNs (Dolzer et al., 2003; Nolte et al., 2013; Schneider et al., 2025). Because blocking Orco does not affect the ultradian gamma frequency band of spontaneous activity that correlates with the frequency ranges of the stimulus concentration-encoding the first, phasic pheromone response, the current finding further confirms that Orco is not the primary transduction channel for pheromone detection (Dolzer et al., 2003; Nolte et al., 2016, 2013). Nevertheless, Orco partakes in pheromone transduction during the later and longer-lasting time window after pheromone stimulation. Mediated through the cAMP-dependent increase in Orco conductance during the activity phase, the ORN membrane potential would oscillate pheromone-dependently at higher ultradian frequencies and with larger amplitude from a lower negative baseline resting potential. Thereby, the ORN’s pheromone sensitivity and sampling frequency of pheromone pulses could be maximized while males fly upwind in search for females.

The upregulation of neural activity during wakefulness/alertness seems to be ubiquitous among different species. For example, the synchronously recorded neurons in human EEG show fast ultradian oscillations in the alpha, beta, and gamma frequency ranges during alertness and slow ultradian oscillations in the delta and theta ranges during rest and sleep. Similarly, in the lower frequency band, the spontaneous spiking activity of hawkmoth ORNs changes from below delta during rest to theta/alpha ranges during their activity phase. While the gamma frequency band comprises spike frequency within a burst and matches the first phasic pheromone response (Dolzer et al., 2003), the spontaneous spiking activity in the theta/alpha range matches the maximal temporal resolution of pheromone pulses of up to 10 Hz (Marion-Poll and Tobin, 1992), being ideally suited to tune ORNs for active sensing. Even though oscillations in the field potential of electroantennogram responses to odor stimulation can reach 100 Hz in different insects (Szyszka et al., 2014), these frequencies most likely do not encode the temporal resolution of physiological odor responses. Apparently, they reflect the astounding adaptive ability of antennal ORNs to synchronize their endogenous membrane potential oscillations across the antenna to a broad range of stimulation frequencies. Pheromone pulse resolution of about 100 Hz was never reported in single olfactory sensilla recordings, and in behavioral experiments male moths stop their upwind mate search if pheromone pulses exceed 30 Hz, apparently due to adaptation (Baker et al., 1985; Bau et al., 2005, 2002; Tripathy et al., 2010).

In summary, Orco-dependent circadian oscillations in the ORN membrane potential are ideally suited to autonomously, fast, and flexibly tune sensitivity and temporal resolution of the pheromone-sensitive ORNs to the hawkmoth’s rest-activity cycle and to acute physiological requirements, synchronizing male and female mating behavior. Here, we have demonstrated how modulation of the leaky pacemaker channel Orco can achieve these circadian oscillations. Whether cAMP levels are under control of the TTFL clock or oscillate based on Ca2+ - and/or octopamine concentration oscillations, being part of a stand-alone PTFL clock, remains to be studied.

Stochastic ion channel dynamics and circadian modulation of Orco conductance explain modeled ORN firing patterns

Our modeling results provide mechanistic insight into the irregular spontaneous bursting activity of pheromone-sensitive ORNs in M. sexta, highlighting stochastic ion channel dynamics and circadian modulation of ion conductance that together shape spiking patterns. By employing a Langevin approximation of the stochastic HH formalism, we demonstrate that the intrinsic noise arising from the random gating of ion channels is sufficient to generate both isolated spikes and bursts of spikes, as observed experimentally. Our model effectively recapitulates the spike distributions found in vivo and accounts for the role of channel noise as a critical determinant of ORN firing patterns.

A key contribution of our model is the incorporation of the Orco ion channel as a pacemaker component, with a conductance that varies linearly with cAMP concentration. We show that post-translational circadian modulation of Orco conductance alone is sufficient to produce the observed time-of-day dependent changes in spike distribution. This finding suggests that the rhythmic expression of cAMP levels across the circadian and/or daily cycle may act as a primary driver of circadian and diurnal changes in ORN excitability, without requiring additional external inputs or modulation of other ion channels.

Importantly, our results reveal that the same set of ion channels can support the emergence of both single action potentials and complex bursting behaviors purely through the stochastic variability of their gating kinetics. This unifying explanation reduces the need for invoking separate biophysical mechanisms or circuit-level inputs to explain these distinct firing modes. For simplicity, the circadian variation in cAMP concentration was modeled using a sinusoidal function. While this approximation is sufficient for capturing the qualitative features of the rhythmic modulation, future work could enhance the biological realism of the model by coupling Orco conductance dynamics to an endogenous circadian oscillator, such as a Goodwin-type TTFL. Incorporating this layer would allow for a more detailed exploration of how endogenous clock components regulate cAMP synthesis and degradation. Additionally, further extensions could include downstream regulatory or coupling mechanisms such as phosphorylation and dephosphorylation events mediated by kinases and phosphatases under circadian control, potentially contributing to fine-tuned modulation of ion channel function over the daily cycle. Together, these findings demonstrate the value of a biophysically grounded, noise-inclusive modeling framework for understanding complex temporal patterns in sensory neuron activity and open the door for future integrative models linking different types of endogenous circadian clocks to cellular excitability.

eLife Assessment

This study uses a technically challenging long-term in vivo recordings and computational modeling to investigate whether hawkmoth olfactory receptor neurons show circadian modulation of spontaneous firing. The study provides solid evidence that ORNs have a circadian pattern of spontaneous firing that is Orco-dependent and yet that Orco transcript abundance does express a circadian rhythmicity and that cAMP can modulate all Orco-dependent activity. Together with computational work, the work is valuable in proposing a provocative hypothesis that an Orco-centred post-translational feedback loop generates the circadian rhythm.

Joint Public Review:

The revised manuscript is much clearer, and the additional analyses address several of the original concerns. RAIN analyses (Rhythmicity Analysis Incorporating Nonparametric methods) now detects circadian rhythmicity in 7/11 recordings under light-dark conditions and 8/12 recordings in constant darkness, compared with 2/12 following treatment with the Orco antagonist. This supports circadian modulation of spontaneous firing and a role for Orco in its normal expression. The expanded qPCR analysis of Orco also supports the conclusion that Orco transcript abundance is not circadian, and the cAMP experiment shows that cAMP can modulate Orco-dependent activity.

The remaining issue concerns the mechanistic interpretation. The lack of rhythmic Orco transcript abundance does not distinguish an autonomous post-translational feedback-loop (PTFL) clock from a model in which the canonical transcriptional-translational (TTFL) clock acts upstream through cAMP, calcium, kinases, phosphatases, channel trafficking, or related pathways to regulate Orco.

Similarly, the new Figure 10 provides a useful representation of the authors' hypothesis, but the proposed delayed feedback and coupling mechanisms are not experimentally demonstrated.

We do not think any further experiments are necessary for the present study. Instead, we recommend that the manuscript should clearly distinguish between what the data show and what remains proposed. The data support circadian modulation of ORN firing a role for Orco in its normal expression, non-circadian Orco transcript abundance, and cAMP-sensitive modulation of Orco-dependent activity. The proposal that an Orco-centred membrane feedback loop generates the rhythm is intriguing and may remain a hypothesis generated through this study that needs formal testing in the future. This should be explicitly stated. While this has been done in the discussion section, elsewhere, including in the abstract and elsewhere, the original claim remains.

Author response:

The following is the authors’ response to the previous reviews.

The revised manuscript is much clearer, and the additional analyses address several of the original concerns. RAIN analyses (Rhythmicity Analysis Incorporating Nonparametric methods) now detects circadian rhythmicity in 7/11 recordings under light-dark conditions and 8/12 recordings in constant darkness, compared with 2/12 following treatment with the Orco antagonist. This supports circadian modulation of spontaneous firing and a role for Orco in its normal expression. The expanded qPCR analysis of Orco also supports the conclusion that Orco transcript abundance is not circadian, and the cAMP experiment shows that cAMP can modulate Orco-dependent activity.

The remaining issue concerns the mechanistic interpretation. The lack of rhythmic Orco transcript abundance does not distinguish an autonomous post-translational feedback-loop (PTFL) clock from a model in which the canonical transcriptional-translational (TTFL) clock acts upstream through cAMP, calcium, kinases, phosphatases, channel trafficking, or related pathways to regulate Orco.

Similarly, the new Figure 10 provides a useful representation of the authors' hypothesis, but the proposed delayed feedback and coupling mechanisms are not experimentally demonstrated.

We do not think any further experiments are necessary for the present study. Instead, we recommend that the manuscript should clearly distinguish between what the data show and what remains proposed. The data support circadian modulation of ORN firing a role for Orco in its normal expression, non-circadian Orco transcript abundance, and cAMP-sensitive modulation of Orco-dependent activity. The proposal that an Orco-centred membrane feedback loop generates the rhythm is intriguing and may remain a hypothesis generated through this study that needs formal testing in the future. This should be explicitly stated. While this has been done in the discussion section, elsewhere, including in the abstract and elsewhere, the original claim remains.

We thank the referees and the editors for their efforts and their positive feedback. As requested, we clarified throughout our manuscript that our proposal of an Orco-centered post-transcriptionally controlled feedback loop in the plasma membrane, which we term PTFL clock, is a novel hypothesis that needs to be tested in future experiments.

Version 4

Abstract

The mating behavior of nocturnal Manduca sexta hawkmoths is under strict temporal control. It is orchestrated via circadian and ultradian oscillations in sex-pheromone stimuli as social Zeitgeber. The extremely sensitive pheromone-detecting olfactory receptor neurons (ORNs) that innervate the long trichoid sensilla on the male’s antennae are peripheral circadian clocks. They express the transcriptional-translational feedback loop (TTFL) circadian clockwork, best characterized in Drosophila melanogaster. In hawkmoths, it is still unknown whether or how the ORN TTFL clockwork regulates the daily rhythms in pheromone sensitivity and in temporal resolution of ultradian pheromone pulses as prerequisites to the temporal regulation of hawkmoth mating behavior. We hypothesize that, rather than the slow TTFL clock, a more rapidly adaptive post-translational feedback loop (PTFL) clockwork, assembled in a signalosome in the ORN plasma membrane, allows for temporal control of pheromone detection via generation of multiscale endogenous membrane potential oscillations. The potential oscillations of the PTFL clock could rapidly synchronize to oscillations of pheromone stimuli at different time scales, thus enabling the prediction of stimulus patterns as a mechanism for active sensing. With in vivo long-term tip recordings of long trichoid sensilla of male hawkmoths, we analyzed the spontaneous spiking activity indicative of the ORNs’ endogenous membrane potential oscillations. Consistent with our hypothesis of a multiscale PTFL clock in hawkmoth ORNs, spontaneous spiking was modulated on ultradian and circadian time scales, with maximum activity at night. When we blocked the evolutionarily conserved olfactory receptor coreceptor (Orco), the circadian modulation was abolished but the ultradian frequency modulation of the spontaneous activity remained. Consistent with PTFL control, Orco was not under the transcriptional control of the TTFL clock, but its modulation of spontaneous spiking activity was dependent on cAMP. We could replicate the experimental data in a conductance-based computational model of an ORN. In this model, Orco conductance changed as a function of fluctuating 2nd messenger levels. This study demonstrates that a PTFL clock is sufficient to impose a circadian pattern on ORN sensitivity. Future experiments are necessary to further challenge our novel hypothesis.

Introduction

Biological clocks generate endogenous physiological and behavioral oscillations on multiple timescales that persist in the absence of any external timing cues under constant conditions (Lamont and Amir, 2010). Such clocks are essential to generate internal rhythms and to permit their synchronization to external Zeitgebers at time scales from milliseconds to years. These synchronized endogenous multiscale clocks thus provide both a continuous time frame for physiology and behavior - the organism’s “present time” - and enable organisms to predict and respond to regular, periodic environmental changes (Stengl and Schneider, 2024). Even though endogenous clocks are widespread in prokaryotes and eukaryotes (Géron et al., 2024; Hall and Rosbash, 1993; Millar, 2016; Thaiss et al., 2014), current knowledge about how multiscale clocks are linked over widely diverging time scales remains sparse.

Insect circadian clock neurons, which are tightly connected to visual brain circuits, are amongst the best studied biological clocks (Hall and Rosbash, 1993; Nishiitsutsuji-Uwo and Pittendrigh, 1968; Reinhard et al., 2024). They regulate physiological and behavioral oscillations that underlie sleep-wake rhythms tied to the 24 h light-dark (LD) cycles on Earth (Dubowy and Sehgal, 2017). Circadian clock neurons possess a molecular clockwork of transcriptional-translational feedback loops (TTFLs). Entrained to the LD cycle, this TTFL clock generates endogenous oscillations of clock gene mRNA and protein levels with a cycle period of ∼24 h via negative feedback by proteins that inhibit their own transcription (Hardin, 2011). It is assumed for both insects and mammals that this TTFL comprises the master clockwork that regulates all physiological and behavioral rhythms in the circadian time range through genetically controlled output pathways (Hastings et al., 2020).

Molecular TTFL clockworks were not only found in central brain clock neurons but also in sensory neurons, such as olfactory receptor neurons (ORNs) in insect antennae, where they play a role in odor-dependent circadian behaviors like feeding, pollination, reproduction, and navigation (Flecke et al., 2010; Flecke and Stengl, 2009; Krishnan et al., 1999; Merlin et al., 2009, 2006; Page and Koelling, 2003; Plautz et al., 1997; Sauman and Reppert, 1998; Schendzielorz et al., 2015; Schuckel et al., 2007; Tanoue et al., 2004; Zhou et al., 2005). Little is known about how the TTFL clockwork of insect ORNs regulates circadian rhythms in chemosensory sensitivity that underlie the circadian control of these different behaviors.

At night, male nocturnal Manduca sexta hawkmoths are very sensitive to the sex pheromone blend that their conspecific females release in an intermittent pattern (Itagaki and Conner, 1988). The intermittency is a prerequisite to maintain the male’s search flight, which is suspended at pulse frequencies >30 Hz (Baker et al., 1985; Baker and Vogt, 1988; Stengl, 2010). The pheromone pulses are resolved of up to 3 Hz by pheromone-specific ORNs in the antenna’s long trichoid sensilla (Lee and Strausfeld, 1990; Marion-Poll and Tobin, 1992). Pheromone-specificity is provided by odor receptor (OR) subunits in the ciliary membranes of ORNs that heteromerize with the conserved olfactory receptor co-receptor Orco (Nakagawa and Vosshall, 2009; Stengl and Funk, 2013; Wicher and Miazzi, 2021). The insect odor transduction cascade is still under debate and seems to involve both ionotropic and metabotropic pathways (Sato et al., 2008; Schneider et al., 2025; Wicher et al., 2008). While there is no general agreement on all roles of Orco, it is known that in heterologous expression systems Orco homo- and heteromers form a slow, leaky, non-specific cation channel, which continuously depolarizes the cells (Jones et al., 2011; Sargsyan et al., 2011; Sato et al., 2008; Wicher, 2018; Wicher et al., 2008). In vivo tip recordings of long trichoid sensilla in M. sexta confirmed that Orco acts as leaky cation channel, acting as a pacemaker channel, because it promotes ultradian membrane potential oscillations that underlie the spontaneous action potential activity (Nolte et al., 2013). These potential oscillations turn the ORNs into ultradian temporal filters that are sensitive to the frequency of the pheromone pulses they encounter. Pheromone pulses that arrive at a similar frequency as the endogenous sub-threshold membrane potential oscillations would synchronize these oscillations and would be more likely to elicit action potential trains. Therefore, this pacemaker channel-dependent mechanism would allow for active sensing if endogenous membrane potential oscillations are in the physiological range that is necessary for tracking intermittent pheromone pulses.

It is not known whether the Orco-dependent membrane potential oscillations also occur at circadian time scales, and whether or how they are linked to ultradian potential oscillations. One hypothesis is that the molecular TTFL clockwork of ORNs exclusively controls endogenous circadian rhythms of the membrane potential via circadian expression of pacemaker channels like Orco. As alternative hypothesis, the plasma membrane of ORNs could by itself comprise a signalosome that constitutes an autonomous clock that is entrained by, but does not rely on, the TTFL clock. Constant levels of colocalized ion channels with antagonistic effects that do not require daily degradation and daily transcription via the TTFL clock, but with modulation of conductance by post-translational modifications, would constitute a more economical and faster adjustable mechanism to generate endogenous membrane potential oscillations (Stengl and Schneider, 2024). This would facilitate active sensing by signalosome-anchored receptors, rapidly tuning odorant/pheromone detection to behaviorally relevant Zeitgeber signals in the insect’s environment, other than the LD cycle that is linked to the TTFL clock.

Here, we performed a combination of experimental and computational modeling studies to investigate whether Orco plays a role in active sensing at the circadian and/or ultradian time scale via generation of endogenous membrane potential oscillations. We hypothesize that Orco is part of a post-translational feedback loop (PTFL) clockwork, assembled in a signalosome in the ORN somatic plasma membrane, rather than being TTFL controlled. To capture naturally occurring, physiologically relevant processes in insect ORNs we performed in vivo long-term tip recordings of the spontaneous spiking activity of pheromone-sensitive trichoid sensilla of intact but restrained hawkmoth males.

We corroborated the experimental data with a Hodgkin-Huxley (HH) based computational ORN model. In general, the widely used deterministic HH models do not include circadian control of ion channels. Moreover, the highly dynamic, bursting firing patterns of ORNs cannot be described by the traditional deterministic modeling of neuronal activity. This constitutes a drawback because stochasticity plays a crucial role in shaping firing patterns which are influenced not only by external stimuli but also by the inherent noise involved in channel gating. Therefore, to address the above points realistically, we developed a Langevin formulation of the HH model, which explicitly accounts for the noise introduced by the random opening and closing of ion channels and circadian modulation of the Orco channel conductance. By integrating channel noise, we could explore how circadian regulation of ionic currents contributes to the variability of firing patterns and how this noise interacts with external stimuli to influence olfactory signal processing.

Our model aims to highlight the advantages of using a stochastic framework for modeling the circadian activity of ORNs. The combined experimental and modeling efforts generate a more complete picture and predictions concerning these autonomous processes in sensory neurons, which are required for future exploration of physiologically relevant anticipatory behaviors that are driven by biological clocks.

Methods

Animals

Nocturnal M. sexta hawkmoths were bred and raised from eggs in the rearing facility at the University of Kassel. Males and females were separated during pupal stages. Males were housed isolated from the females to avoid exposure to female pheromones. All experiments were performed with adult males reared in a long-day photoperiod (L:D 17:7 h to prevent diapause) at 25 °C with relative humidity of about 55%. Larvae were fed with an artificial diet modified after Bell and Joachim (1976); adult moths could feed on sugar solution with added vitamins (Roth) ad libitum (Riffell et al., 2008).

While all animals were exposed to the same lighting regime as Zeitgeber to entrain them up until the start of the experiments, we also attempted pheromone exposure as additional circadian Zeitgeber to phase-align spontaneous ORN activity in a subset of males (Ghosh et al., 2024). Without allowing direct contact, virgin males were exposed at Zeitgeber time (ZT) 16 to females and naturally occurring pheromone blends in closed mating cages for 30 mins. Males were then moved back into isolation before tip recordings (see below) were started at ZT 17 on the same day. Since we focused on spontaneous spiking activity rather than on pheromone-dependent responses and because repetitive or long pheromone exposure changes the composition of 2nd messenger modulated ion channels in ORNs (Gawalek and Stengl, 2018), we did not employ any further pheromone synchronization protocols.

Solutions

Ringer compositions were taken from (Pézier et al., 2007) and contained (in mM): 6.4 KCl, 12 MgCl2, 1 CaCl2, 12 NaCl, 10 HEPES, 340 glucose for hemolymph ringer (HLR) at 450 mosmol/kg, and 172 KCl, 3 MgCl2, 1 CaCl2, 25 NaCl, 10 HEPES, 22.5 glucose for sensillum lymph ringer (SLR) at 475 mosmol/kg, both at pH 6.5.

The Orco antagonist OLC15 (N-(4-butylphenyl)-2-((4-ethyl-5-(2-pyridinyl)-4H-1,2,4-triazol-3-yl)thio)acetamide (Chen and Luetje, 2012); kindly provided by Dr. D. Wicher, Max Planck Institute for Chemical Ecology, Jena, Germany) was dissolved in DMSO to a concentration of 100 mM (stock solution stored at 4-8 °C) before being diluted in SLR to a final concentration of 50 µM to be used within one week (stored at 4 °C). We determined in previous studies that, compared to other Orco antagonists, OLC15 is the most specific (Chen and Luetje, 2012; Nolte et al., 2016, 2013; Pask et al., 2013). It is effective at 50 µM, according to dose-response curves, and dose-dependently antagonizes the action of the Orco agonist VUAA1 in intact hawkmoths (Nolte et al., 2016, 2013). DMSO percentage was 0.05% in the final solution.

The cAMP analog 8-bromo-cAMP acetoxymethyl ester (8-Br-cAMP-AM; Biolog Life Science Institute GmbH C Co. KG) was dissolved in DMSO to a concentration of 10 mM (stock solution stored at −20°C) and diluted to a final concentration of 100 µM in either SLR or SLR + 50 µM OLC15.

Electrophysiology

The spontaneous electrical activity of the unstimulated pheromone-sensitive ORNs in a single long trichoid sensillum was recorded extracellularly as in vivo tip recording (Kaissling et al., 1987). 1-2-d-old male M. sexta that were raised and entrained to Zeitgebers as described above were fixed in a custom-made Teflon holder with adhesive tape. The right antenna used for recordings was oriented with its dorsal surface facing up and immobilized near the base with dental wax (Boxing wax strips, KerrHawe SA). Electrodes (Ag/AgCl wire in a glass pipette filled with SLR for the recording electrode and HLR for the reference electrode) were pulled (DMZ-Universal Puller, Zeitz Instruments) from borosilicate capillaries (OD:1.56 mm, ID: 1.17 mm, without filament; Science Products) to a tip diameter of about 2 μm. All recordings were made at a distance of 2/3 to 3/4 of the antenna length towards its distal end because pheromone-sensitive long-trichoid sensilla (sensilla trichodea type 1) occur in a characteristic, homogenous pattern along the hawkmoth antenna (Kaissling et al., 1989; Kalinová et al., 2001; Lee and Strausfeld, 1990). Furthermore, all ORNs in long trichoid sensilla express the same Orco staining pattern (Nolte et al., 2016), share largely overlapping ion channel populations, and obvious differences in the activity of ORNs from different annuli have not been reported (c.f., Dolzer et al., 2003; Gawalek and Stengl, 2018; Schneider et al., 2025). Thus, it is very likely that pheromone encoding mechanisms are not locally restricted along the flagellum. Recording and reference electrodes were located on the same annulus: After cutting off the distal 20 annuli of the antenna with micro scissors, the reference electrode was inserted to a depth of 2 annuli into the antennal lumen from the cut end. The cut was covered with electrode gel (GE Medical Systems Information Technology) to prevent desiccation. Under microscopic control, some tips of the long trichoid sensilla of the annulus were cut off with sharpened Dumont forceps, and the recording electrode was slipped over a single sensillum of an upper sensillar row. Electrodes were moved with manual micromanipulators (MMJ, Märzhäuser Wetzlar; and Leitz, Leica Microsystems) or with motorized micromanipulators (Microstar, Scientifica) with a joystick controller (Scientifica).

ORN activity was amplified 200-fold (custom built amplifier with 1012 Ω input impedance, or ELC-01 MX in combination with DPA-2FX, npi, or EXT 10-2F, npi), low pass filtered at 1.3 kHz, digitized at 20 kHz with a Digidata (1550A or 1550B, Molecular Devices), and saved for offline analysis with Clampex 10 (pCLAMP suite, Molecular Devices).

All tip recording experiments were performed at room temperature (21-27 °C) and a relative humidity of 35-60%. These environmental fluctuations across experiments were based on seasonal influences. Throughout each experiment, conditions were stable. To ensure that neither electrode reattachment to the same trichoid sensillum nor application of DMSO influenced spiking activity, we performed two sets of paired experiments at ZT 1-3 in which we recorded the spontaneous activity before and after removing and reattaching the SLR-filled recording electrode to the same sensillum, or recorded the spontaneous activity of the same sensillum in SLR and SLR + 0.15% DMSO. Both treatments did not affect spontaneous spiking activity (Figure 1B).

Attributes of pheromone-sensitive ORN spiking patterns.

(A) Spikes were identified by their prominence in the raw ORN recording. The inter-spike interval (ISI) was defined as the time from one spike to the next. A train with ≥2 spikes with ISI ≤ 50 ms was considered a burst. The inter-burst interval (IBI) was the ISI between the last spike of the first burst and the first spike of the subsequent burst, omitting any individual spikes in between. Burst duration was the time between the first and last spike of a burst. The inter-event interval (IEI) included the ISI between bursts and individual spikes. The number of action potentials (spikes) per burst, and percentage of total action potentials as part of bursts were computed. The intra-burst interval was the ISI within a burst. Frequency of spikes, bursts, events and intra-bursts were determined as inverse of ISI, IBI, IEI and intra-burst interval. Detaching and reattaching the recording electrode to a long trichoid sensillum (Bi; paired t-test t(9) = −2.184; p = 0.057; N = 10) or adding DMSO to the sensillum lymph ringer (SLR) (Bii; Wilcoxon signed-rank test Z = 0.153; p = 0.922; N = 10) did not affect the mean spiking frequency.

We recorded ORN activity either in long-day conditions (17:7 LD) or constant darkness (DD) to examine the free-running peripheral ORN clock. For LD experiments, the light regime continued as in the rearing facility during the experiment. For DD and OLC15 experiments, the lights were turned off at the beginning of the recording, i.e., all animals were exposed to the same light Zeitgeber until the beginning of the experiment. To examine the effect of Orco on the spontaneous action potential (AP) activity pattern, we used the recording electrode to infuse OLC15 in DD conditions. To examine the effect of cAMP on Orco gating, we recorded the spontaneous ORN activity at the end of the moth’s activity phase (ZT 1-3) in SLR as control, and either 8-Br-cAMP-AM, or 8-Br-cAMP-AM + OLC15, infused through the recording electrode.

ORNs generate spontaneous action potentials. However, high spiking activity during control recordings is indicative of stress, whereas low/no activity could be indicative of damage. Therefore, we only considered experiments in which the average spiking frequency was between 2 Hz and 0.008 Hz in control conditions.

Data analysis

To characterize the spontaneous ORN spiking activity, we first detected spikes with a self-written program in Python. Using the scipy.signal package and find_peaks function, we used a moving mean with 100 samples (5 ms) window size for baseline correction, followed by applying a 8th order low-pass Bessel filter (2000 Hz cutoff frequency) to the raw voltage trace to facilitate spike detection within the following conditions: spike full width (3-15 ms), threshold search for amplitude of spike (0.5-2.5 mV) and threshold prominence (0.1). Two or more consecutive spikes with intervals ≤ 50 ms were considered a burst (Dolzer et al., 2001). We then calculated the following attributes in bins of either 10 min or 1 h for further analysis: instantaneous spike frequency (ISI, 1 / inter-spike interval), inter-burst interval (IBI, time between two consecutive bursts with no single spikes in between), inter-event interval (IEI, time between two events, i.e. two consecutive single spikes, two consecutive bursts, a burst and a consecutive spike, and vice versa), burst duration (time from first to last spike in a burst), percentage of all spikes in the experiment that are members of burst, and mean number of spikes per burst (Figure 1A). All statistics were done with those binned values.

Hawkmoth behavioral and physiological rhythms desynchronize across the population in the absence of Zeitgebers like circadian rhythms of pheromone stimulation and daily light-dark cycles. Therefore, we first used RAIN analysis (Thaben and Westermark, 2014) to identify circadian rhythms (24 ± 4 h) in the attributes above for each animal and condition (LD, DD, OLC15) in the 1 h-binned data. Data was downsampled to 2 h resolution and thus included every other 1 h bin to speed up computational time. To further quantify the circadian changes in attributes across animals, it was then necessary to phase-align the recordings obtained in DD and OLC15. To achieve this, we used an optimization procedure to calculate the optimal temporal lags for each dataset, aiming to maximize the overall cross-correlation between the calculated binned spike frequency series. The optimization was performed using simulated annealing, which provided a set of time shifts (lags) for each series. These lags were applied to shift the frequency data in a manner that maximized the synchronization between the different subjects’ data, resulting in a “subjective ZT” with subjective ZT 0 at the first maximum of the binned ISI for each individual. The objective function used for optimization was the negative sum of pairwise correlations between all series after applying the shifts. The resulting alignment allowed us to compare the frequency patterns across the different conditions.

Once the data were aligned, we computed the mean frequency across all subjects for each bin and each condition. Then, the first high activity period was identified by finding the maximum in the mean frequency series. Subsequently, we identified the second high activity period as the time 24 hours after the first one and the first low activity period was found by locating the minimum in the mean frequency series between the two high activity periods. Besides the mean frequency, we also searched for significant oscillations in IBI, IEI, burst duration, percentage of spikes that are part of bursts, and mean number of spikes per burst. For this we found the maximum value of these statistics within a 3 h range centered in the first low activity period and compared it with the values at the second high activity period.

Continuous wavelet and Fourier analysis were performed using PyBoat (version 0.9.12) (Mönke et al., 2020) to examine the time series of neuronal firing frequency. The continuous wavelet method allows us to detect and characterize rhythmic patterns within the data by identifying frequency components that change over time. Specifically, we used wavelet analysis to determine whether a neuron exhibited circadian and ultradian rhythms. Unlike traditional Fourier analysis, which provides an overall frequency composition but assumes stationarity, wavelet analysis is particularly useful for biological signals because it can reveal how these frequencies evolve dynamically. By applying this approach, we could assess not only the presence of ultradian rhythms but also whether their strength or occurrence fluctuated in a circadian manner, indicating modulation by the 24-hour cycle. For this analysis, we used the 10 min-binned data to have sufficiently high resolution to detect ultradian rhythms. Periods were determined as local maxima in the Fourier power spectra.

Real-time quantitative polymerase chain reaction (qPCR)

We followed the qPCR methodology detailed in (Schneider et al., 2025). Briefly, male moths were isolated and cultured until the second day after eclosion, and the antennae from four individuals were collected every 4 hours for one day, starting at ZT 0, using liquid nitrogen, with four biological replicates per time point. Total RNA was extracted using TRIzol® Reagent (Invitrogen) and quantified for concentration and purity with the NanoDrop ND-1000 spectrophotometer (Thermo Fisher Scientific). First-strand complementary DNA (cDNA) was synthesized using the PrimeScript™ RT reagent Kit with gDNA Erase (Takara Bio Inc.). Primers for Orco (F: GCTCGCTACCACCAAATTGC, R:TCGTGACCCAACTGACAACA; Gene ID: LOC115452348) and for timeless (tim) (F: TTAAGCCGACCGTAGTGCTG, R: CGTCTTCCGTCCATGTGTCT; Gene ID: LOC115446009) were designed using the Primer-BLAST online program of NCBI. The qPCR was conducted following the protocol of TB Green® Premix Ex Taq™ II (Tli RNase H Plus) (Takara Bio Inc.), with the program: 95 °C for 30 s, followed by 40 cycles of 95 °C for 5 s, and 60 °C for 30 s. Relative expression levels of Orco were calculated using the 2-ΔΔCt method (Livak and Schmittgen, 2001).

Statistics

All statistics were done with self-written python codes, SigmaPlot 12 (Systat Software), or JASP (versions 0.19.03 and 0.95.4, University of Amsterdam) with a significance level α = 0.05. Unless noted otherwise, data were checked for normal distribution with Anderson-Darling or Shapiro-Wilk tests and for equal variance with Levene’s test. If one of these assumptions failed, we used non-parametric tests, otherwise we used parametric tests.

To assess the statistical significance of observed differences in neuronal activity of the aligned data between ZTs, we applied the Wilcoxon signed-rank test to pairwise comparisons between the first low activity period and the subsequent high activity period and discarded the transient high activity at the beginning of most recordings.

For paired recordings (electrode reattachment, addition of DMSO, or cAMP/cAMP+OLC15 in the same animal, we used either paired student-t test or Wilcoxon signed-rank test, depending on normal distribution of data.

For other data, we used an ANOVA on ranks with either Tukey’s or Dunn’s post-hoc test for multiple comparisons as indicated.

ORN Model

In this study, we introduce a single-compartment Hodgkin-Huxley (HH) type model of the spontaneous activity of M. sexta ORNs (available on GitHub (Forlino et al., 2025)). The model contains 5 sets of ion channels that were previously identified in electrophysiological recordings (Dolzer et al., 2021): the fast sodium (Na+) and potassium (K+) channels responsible for the generation of spikes, low voltage-activated calcium channels (LVA) responsible for the initiation of bursts, calcium-gated potassium channels (BK) responsible for the termination of bursts, and the Orco channel, which is a voltage-independent non-specific, cAMP-gated channel. A schematic of the model is provided in Figure 2A. We modeled Orco as a ZT-dependent conductance to represent the effect of the oscillations in cAMP concentration that were identified in both hawkmoth and cockroach antennae (Schendzielorz et al., 2014, 2015, 2012) (Figure 2B, C). For simplicity we chose to use a sine function to directly represent the oscillations in Orco conductance, but similar results can be achieved by modeling it as an output of a Goodwin model of the circadian clock via cAMP concentration.

The conductance-based model of a simplified hawkmoth ORN.

(A) Schematic of the model containing a voltage gated sodium channel (Na+), a voltage gated potassium channel (K+), a low voltage activated calcium channel (LVA), a calcium and voltage activated potassium channel (BK), a leak channel, and a cAMP-gated leaky non-specific cation channel (Orco). The Orco channel is represented as a linear conductance whose value oscillates with a sinusoidal shape as a function of the ZT. The circadian rhythm in the Orco conductance is hypothesized to result from a cAMP-dependent increase in channel open-time probability (as found in Drosophila Orco: (Getahun et al., 2013)) via a circadian oscillation of antennal cAMP concentrations (Schendzielorz et al., 2015). This oscillation is driven by the circadian clock and serves as an input; the internal mechanism of the clock is not modeled explicitly. Arrows indicate the flow of ions through the channels. (B) Parametrized circadian oscillation in Orco conductance introduced as an external input to our model, representing the gating of Orco by cAMP whose concentration oscillates on a circadian time scale. The times of minimum and maximum cAMP concentration are indicated by colored arrows. (C) Parametrized IV curve of the Orco channel showing a linear behavior with the slope depending on the cAMP concentration. Line colors correspond to the arrows in B. (D) Markov-chain representing all possible states and transitions for each type of ion channel included in the model. The per capita transition rates (α and β) depend on membrane potential and in the case of the BK channel also on the intracellular Ca2+-concentration. The “N” state vectors contain the population of channels in each state “n”. Directed edges are numbered in red. The only conducting state of each ion channel, representing the condition where all its gates are open, is shown in green.

Each type of ion channel possesses one or more gates that can open and close individually with transition rates α and β for opening and closing, respectively, of activation gates (n, m) and inactivation gates (h). As a result, each possible combination of open and closed gates defines a possible state in which each ion channel can be found. A schematic of the Markov chain depicting all possible states and transitions for each ion channel is shown in Figure 2D. Each ion channel can conduct current only in the state where all its gates are open.

First, we constructed a deterministic 22-dimensional model to serve as the mean field of the Langevin model following the methodology proposed by Pu and Thomas (2020). Here, one dimension tracks the membrane potential, another the intracellular Ca2+ concentration, and the remaining 20 dimensions track the number of channels that are in a specific state of the Markov chains. Considering the conservation of the total number of ion channels, each type of ion channel required one less dimension than the total possible states.

The complete Langevin model adds to the mean field an independent noise source for each transition in the Markov chain. These noise sources are biophysically motivated since the transition between ion channel states is a stochastic process, with the probability of transition determined by the kinetics of each kind of ion channel. The variability in the spiking behavior of a neuron emerges from this non-deterministic gating of individual ion channels. The detailed mathematical description of the model can be found in the Supplementary Material.

Results

We tested the hypothesis that sensory neurons, such as insect ORNs, perform active, anticipatory sensing based upon an endogenous plasma membrane clock with the pacemaker channel Orco as the core element. We focused on pheromone-sensitive long trichoid sensilla of male M. sexta hawkmoth antennae which express daily rhythms in pheromone sensitivity and temporal resolution that are controlled by a circadian clock. We performed minimally invasive in vivo tip recordings over the course of several days to search for daily rhythms in spontaneous spiking activity as a measure of endogenous membrane potential oscillations.

Pheromone-sensitive hawkmoth ORNs exhibit Orco-dependent circadian rhythms in spontaneous spiking activity

The ORNs that innervate the pheromone-sensitive long trichoid sensilla on male M. sexta antennae display spontaneous spiking patterns of irregular bursts with interspersed single action potentials in short-term recordings (Dolzer et al., 2001). In the present study we documented the same pattern in long-term recordings over the course of 2-7 days. This enabled us to search for daily and circadian rhythmicity in spontaneous ORN spiking activity in the absence of pheromone stimulation.

RAIN analysis of the spontaneous spiking patterns revealed significant daily rhythms in 7 of 11 animals under long-day conditions (17:7 h LD) (Figure 3A). The maxima in spontaneous activity occurred mostly during the dark phase when nocturnal hawkmoths are active (Figure 3A (top), Bi light blue arrow), and minima occurred during the light phase when hawkmoths rest/sleep (Figure 3A (bottom), Bi dark blue arrow). High spiking activity in the dark phase had a mean maximum frequency of ∼3 Hz (delta range: 0.5 – 4 Hz) (Figure 3Bii). In contrast, the minimum spiking frequencies during the hawkmoth’s resting phase were below 0.5 Hz (Figure 3A (bottom)).

Spontaneous activity of pheromone-sensitive olfactory receptor neurons (ORNs) of male M. sexta shows Orco-dependent circadian modulation.

(A) Example tip recordings of the spontaneous activity of one long trichoid sensillum during a period of high (top, average frequency ∼ 3 Hz) and low (bottom, average frequency ∼ 0.2 Hz) activity. (Bi) Spontaneous spike frequency of one long-term ORN recording during 17:7 light-dark cycles (same experiment as in A, colored arrows indicate respective Zeitgeber times (ZT)), indicated by the black-white bar at the top, increased during each activity phase and was low during the resting phases. Spike frequencies were averaged for each 1 h bin. (Bii) Mean spontaneous ORN spike frequency across all LD animals in 1 h bins. Data from the individual in Bi is highlighted in red. Recordings had different lengths, therefore the number of recordings used for the element-wise mean for each 1 h bin decreased with the time since the start of the recording, indicated in the top panel. (Biii) Mean spontaneous ORN spike frequency across all LD animals in 1 h bins during the first 48 hours revealed circadian activity. (Ci) The spontaneous spike frequency of one long-term ORN recording in constant darkness (DD, the expected times of lights on are represented by grey bars above) exhibited a circadian pattern. The peak activity shifted in DD due to the animal’s endogenous, free-running circadian period τ = 21.27 h. (Cii) Mean spontaneous ORN spike frequency across all DD animals in 1 h bins. Data from the individual in Ci is highlighted in red. (Ciii) Mean spontaneous ORN spike frequency across all DD animals in 1 h bins during the first 48 hours after aligning the time to the first maximum in spontaneous activity (see Methods) revealed circadian activity. (Di) The spontaneous spike frequency of one long-term ORN recording in constant darkness with infusion of the Orco antagonist OLC15 (orange frame) dissipated the circadian rhythm of spontaneous activity. (Dii) Mean spontaneous ORN spike frequency across all OLC15 animals in 1 h bins. Data from the individual in Di is highlighted in red. In contrast to LD and DD conditions, the spike frequency decreases over time with prolonged exposure to OLC15. (Diii) Mean spontaneous ORN spike frequency across all OLC15 animals in 1 h bins (mean ± SD) during the first 48 hours after aligning the time to the first maximum in spontaneous activity (see Methods). In contrast to LD and DD conditions, the circadian change in spontaneous spiking frequency disappeared. (E) The slopes of the linear fits to the binned spiking activity of each individual animal in the three different conditions (see Methods). Each dot indicates the slope for one experiment. The line indicates the mean. The slope for OLC15 is significantly different from control LD.

Since ORN firing patterns had a clear daily rhythm in LD both in individual animals (Figure 3Bi) and averaged across animals (Figure 3Bii, Biii), we examined whether these rhythms depended on cycling light cues or whether they persisted in constant darkness (DD), as expected for an endogenous, circadian clock-driven rhythm. In DD, the spontaneous ORN firing pattern remained rhythmic (RAIN: 8 of 12 animals), albeit with less robustness: The single peak of high spiking activity that was observed at the same ZT across several days in LD divided into multiple bouts of high spiking activity in the activity phase in 8 of 12 animals (Figure 3Ci). Also, in DD conditions, the high spiking activity phase shifted into the subjective day due to the individual endogenous circadian periods of about 23.5 ± 2.8 h (21.27 h in Figure 3Ci). Using pheromone exposure as additional Zeitgeber also did not synchronize the animals (Figure 3 – Supplement). Thus, averaging the instantaneous spiking activity across all DD animals did not depict clear circadian rhythms (Figure 3Cii). Therefore, we phase-aligned the mean frequency of spontaneous activity (see Methods). After alignment, circadian rhythms were visible across DD animals (Figure 3Ciii), comparable to the daily rhythms in LD (cf. Figure 3 Biii and Ciii).

Next, we examined whether Orco, a leaky, non-specific cation channel, is the dominant depolarizing pacemaker current of ORNs that drives circadian rhythms in spontaneous spiking activity. Infusion of the Orco antagonist OLC15 into the sensillum lymph obliterated circadian rhythms and attenuated the spontaneous activity in several, but not all experiments (RAIN: 2 of 12 animals expressed circadian rhythms) (Figure 3Di). This attenuation resulted in a linear decrease in spiking activity over several days (Figure 3Dii, Diii). Even after phase-aligning the individual experiments, no circadian rhythm was present in the average across all OLC15-treated animals (Figure 3Diii).

Because OLC15 is not membrane-permeable on its own it was infused with low concentrations (0.05%) of DMSO through the recording electrode into the sensillum lymph and, therefore, effectiveness increased over time. Hence, we fit the spiking activity of each animal and in each condition with linear regression lines over the whole recording time to compare the slopes of the fits (Figure 3E). In LD, the mean ± SD slope was 0.004 ± 0.006 Hz/h, indicating that the average spiking activity remained similar throughout the duration of the recording. In DD, the slope was −0.003 ± 0.018 Hz/h and was not different from LD. In contrast, the mean slope during OLC15 treatment was −0.007 ± 0.022 Hz/h, significantly smaller than in LD (ANOVA on ranks with Dunn’s post-hoc test; H(2) = 7.681, p = 0.021). The negative slope corroborated the finding that spiking activity decreased over time in the presence of the Orco blocker.

Activity rhythms between hawkmoths desynchronize and have only weak phase coupling in the absence of pheromone stimulation, even in LD cycles (Ghosh et al., 2024). Therefore, we first determined the percentage of animals that expressed circadian rhythmicity (period of 24 ± 4 h) for each attribute with RAIN (Figure 4A). In LD, all attributes except #AP per burst and intra-burst frequency showed daily modulation in most animals. Circadian rhythms were still present in more than half of the animals in DD for mean spiking frequency, mean burst frequency, and mean event frequency. In contrast, the attributes of only a few animals remained rhythmic with the Orco channel blocker OLC15.

Blocking Orco removed circadian modulation of spontaneous ORN spiking patterns.

RAIN analyses demonstrated significant circadian rhythmicity for most animals in most attributes of the spontaneous spiking activity in LD and DD conditions but not in OLC15 (A). Values and background colors indicate the percentage of animals that expressed significant circadian rhythmicity. Circadian differences of attributes were further quantified between the time windows of low vs. high activity (Wilcoxon signed-rank test, α = 0.05; B-H). Significant differences occurred mostly in LD (white boxes) and DD (gray boxes) conditions, but never when Orco was blocked with OLC15 (orange boxes). In both LD and DD, the mean spiking frequency was increased during high activity (Bi, Bii). Only in LD but not in DD, the mean burst frequency (Ci, Cii) increased, whereas the relative number of spikes belonging to a burst (Di, Dii) decreased significantly during the high activity period. Both in LD and DD the mean event frequency (Fi,Fii) increased significantly. Mean burst duration (Ei, Eii), mean number of spikes per bust (Gi,Gii), and mean intra-burst intervals (Hi, Hii) did not differ between low vs. high spiking activity in LD and DD. When blocking Orco, the attributes for low vs. high activity did not differ for any of the attributes tested (Biii-Hiii). In DD and OLC15conditions, activity phases were aligned as described in the Methods; thus, low activity in DD was at subjective ZT 19 and in OLC15 at subjective ZT 13, high activity was in both cases at subjective ZT 24. ZTs for low and high activity in LD were 10 and 0, respectively.

To further quantify the circadian differences, especially of the desynchronized animals, we compared different attributes of spontaneous activity between periods of high and low activity for DD conditions after phase alignment (see Methods) instead of comparing fixed ZT or circadian time (CT) intervals. Despite the high variability in ORN spiking between individuals, the mean frequency of spontaneous ORN activity was significantly higher during the high activity phase than during low activity in both LD and DD recordings (Figure 4Bi, Bii). In LD, but not in DD, the frequency of bursting increased significantly during the high activity phase (Figure 4Ci, Cii). The percentage of spikes that are part of bursts was less variable and lower during the high activity phase in LD but not significantly different in DD (Figure 4Di, Dii). The mean burst duration was less variable during the high than the low activity phase in LD, but not significantly shorter in any condition (Figure 4Ei, Eii). Furthermore, both in LD and DD, the mean inter-event frequencies were significantly higher during the high activity phase (Figure 4Fi, Fii), indicating a significantly higher overall spontaneous spiking activity during the hawkmoth’s activity phase. The mean number of spikes per burst, and intra-burst frequencies did not differ significantly between high or low activity periods in both LD and DD conditions (Figure 4Gi, Gii, Hi, Hii). The addition of Orco antagonist OLC15 deleted any significant differences of spontaneous activity found in LD and DD (Figure 4Biii-Hiii). In conclusion, the mean frequency of spontaneous spiking and the frequency of bursting expressed circadian modulation, and are both most likely controlled via a circadian clock that involves Orco.

Orco imposes circadian modulation on the ultradian rhythms in spontaneous ORN activity

After establishing that Orco influences the circadian changes in ORN spiking activity between the animal’s resting and activity phase, we examined if Orco also influences ORN spiking rhythms on the largely different ultradian time scales in the frequency range between 0.01-1000 Hz. In all conditions, two broad bands of ultradian instantaneous frequencies were evident: one upper band of >10 to ∼100 Hz (beta/gamma range) and a lower band between 0.1 to ≤10 Hz (delta/theta/alpha range) (Figure 5). The upper band primarily represents the frequencies of spikes within a burst, while the lower band represents single spikes and frequencies between bursts.

Blocking Orco removed circadian regulation of ultradian rhythms in the spontaneous ORN spiking pattern.

Heat map of the instantaneous frequencies (1/ISI) over two consecutive days of long-term tip recordings of one pheromone-sensitive long trichoid sensillum under different conditions (LD (A), DD (B), OLC15 (C)) in each panel. Representative recordings from one animal in each panel. Pixel color indicates the counts of instantaneous frequencies in that respective bin. The points fall mostly in a band of high frequencies (>10 – ∼100 Hz) and low frequencies (0.01 – ∼10 Hz), with the high-frequency band representing the instantaneous frequencies of spikes within a burst and the low-frequency band the instantaneous frequency between bursts. (A, B) The high frequency band indicates daily and circadian modulation of frequency prevalence. In addition, the low frequency band (<0.01 Hz to ∼ 10 Hz) also displayed circadian modulation of the frequency composition. (C) Infusion of OLC15 deleted the circadian modulation in both frequency bands but not the ultradian rhythm of frequency prevalence.

In LD, two types of daily/circadian modulation of the ultradian frequencies occurred (Figure 5A). The upper frequency band was mostly modulated in prevalence (its frequency of occurrence) but not in its frequency range. The lower band was modulated both in prevalence and frequency range, which resulted in a sinusoidal change in the mean of the frequency band over 24 h. During the hawkmoth’s activity phase, higher frequencies of up to ∼10 Hz occurred, while frequencies dropped below 1 Hz during the resting phase. Because of this circadian increase in the lower frequency band, it merged with the high frequency band when both bands had maximum prevalence during the activity phase around ZT 0.

The circadian modulation of both frequency bands was maintained in constant darkness (Figure 5B) but phase-shifted with respect to the projected ZT due to the endogenous circadian period of each individual. Again, this indicates that a clock is modulating the spontaneous ORN spiking activity over the course of a day. The addition of the Orco antagonist OLC15 deleted any circadian modulation of either frequency band. Interestingly, the instantaneous frequencies in both bands still showed correlated changes in prevalence and merging, but now at ultradian periods only.

Additional Fourier analysis (see Methods) of spontaneous spiking activity over the course of several days revealed faster ultradian and slower infradian rhythms, ranging from 0.6 h to 38 h, in addition to the circadian (24 ± 4 h) rhythm (Figure 6Ai, Bi, Ci). Some rhythms (factorials of 24 h, like 12 h, 6 h, etc.) appeared to be harmonics of the circadian rhythm, which were not excluded in our analysis. Most animals expressed ultradian rhythms of 2-4 h in LD as well as DD conditions. Fourier confirmed the decrease of animals with circadian spiking activity in OLC15 (58.3% in DD, 33.3% in OLC15). Please note that discrepancies in the number of animals with circadian rhythms between RAIN and Fourier/wavelet analyses are due to the fundamentally different sensitivities in methodology. In brief, Fourier analysis assumes a stationary sinusoidal component, whereas RAIN as a rank-based umbrella test detects consistent raise-fall patterns, including asymmetric rhythms. Nonetheless, both methods confirmed a dramatic decrease in animals with circadian spiking patterns in OLC15. In addition, 33.3% of animals expressed infradian spiking patterns (up from 0% in DD) with OLC15. Ultradian periods could be detected in all animals in all conditions, but their distribution did not differ between conditions (1-way ANOVA; F(2, 117) = 0.87, p = 0.421). These results were corroborated by continuous wavelet analysis, which provided time-resolved spectral power across different rhythmic periods (Figure 6Aii, Bii, Cii). In control LD and DD conditions, wavelet power for ultradian periods increased during times of elevated firing rate and decreased during periods of low activity (Figure 6Aii and Bii, middle and bottom panels). This dynamic pattern indicates that ultradian rhythms were not constant over time but were instead modulated with a circadian rhythm, becoming more prominent during circadian peaks in neuronal activity. Such circadian gating of ultradian power was clearly visible in the wavelet spectrograms as alternating bands of higher and lower ultradian power aligned with daily activity cycles. Following Orco blockade with OLC15, this modulation was disrupted (Figure 6Cii middle panel): although ultradian components persisted (Figure 6Cii bottom panel), they no longer showed systematic variation with the circadian cycle, suggesting a decoupling of ultradian rhythms from circadian control.

Blocking Orco affected ultradian and infradian frequencies in the spontaneous ORN firing patterns.

The Fourier analysis of spontaneous spiking activity revealed rhythms with ultradian (<20 h; dark grey), circadian (20-28 h; white), and infradian periods (>28 h; light grey) in LD (Ai), DD (Bi), and OLC15 (Ci). Each line represents one animal where each triangular marker is at the local maximum in the frequency spectra obtained for that animal. The histograms above illustrate how often specific periods of spiking activity rhythms occurred averaged over all animals. Circadian rhythms were detected in LD (N = 5 of 11), DD (N = 7 of 10), and OLC15 (N = 4 of 12). Wavelet analysis of LD (Aii), DD (Bii), and OLC15 (Cii) confirmed the occurrence of multiscale periods. Example plots for one animal each, the same animals as in Figure 5. For each panel, the top plot depicts the mean firing frequency, the middle plot the wavelet power in the same period range as panels i, and the bottom plot the wavelet power for infradian periods up to 7 h on a log scale to highlight infradian time scales. Gray dashed lines indicate cone of uncertainty.

Orco expression is not under the control of the molecular TTFL clockwork

So far, our results indicated a prominent role of Orco in the circadian modulation of ORN spiking activity. Since it is the general opinion that the master circadian clock is the TTFL clock that dominates all physiological and behavioral circadian rhythms via transcriptional control, we examined whether Orco transcript abundance shows a daily rhythm, as previously found for clock genes in hawkmoth antennae (Schneider et al., 2025; Schuckel et al., 2007). However, in contrast to the circadian clock gene tim, qPCR of whole male hawkmoth antennae across different ZTs revealed no significant daily rhythm of Orco transcript abundance (Figure 7; ANOVA on ranks, H(5) = 6.91, p = 0.227). Thus, we suggest that Orco function is controlled via a post-translational circadian clockwork mechanism that is associated directly with the plasma membrane (Stengl and Schneider, 2024; Stengl and Schröder, 2021).

Orco is not under control of the TTFL-based molecular circadian clock.

Relative expression levels of Orco and tim mRNA from male hawkmoth antennae raised under LD conditions. Orco expression levels did not differ significantly between ZTs (ANOVA on ranks, H(5) = 6.91, p = 0.227: N = 4 per ZT). tim expression levels changed significantly throughout the day (1-way ANOVA, F(5, 18) = 6.215, p = 0.002) Red lines indicate mean, tim served as positive control.

Circadian changes in Orco conductance are sufficient to modulate spiking activity in a computational model of a hawkmoth ORN

The maximum conductance of Drosophila Orco as a leaky ion channel is modulated by cAMP (Getahun et al., 2013; Martín et al., 2001; Sargsyan et al., 2011; Wicher et al., 2008). In hawkmoth antennae, cAMP levels oscillate throughout the day and are involved in sensitizing the ORNs to pheromone (Dolzer et al., 2021; Flecke et al., 2010; Flecke and Stengl, 2009; Schendzielorz et al., 2014, 2015). Hence, cAMP modulation of Orco conductance is a putative mechanism for the Orco-dependent circadian modulation of ORN spiking activity, integrating Orco in a circadian PTFL-based membrane clockwork (Stengl and Schneider, 2024).

To test the hypothesis that Orco is a cAMP-dependent, leaky ion channel, we constructed an ORN model (Methods and Supplementary Material) and tuned the parameters to provide a quantitative match with the biological data. The developed model provides a unified description of experimental observations. For instance, the ORN model yields irregular spiking activity with a bimodal distribution, consisting of both bursts and isolated spikes, successfully reproducing the experimental recordings (Figure 8A). This suggests a unified mechanism underlying the initiation of both phenomena: If the pacemaker-driven depolarization reaches the Na+ threshold, a firing process begins, and at this point randomness determines if other slower ion channels responsible for bursting are sufficiently activated to sustain a burst of several spikes.

The model adequately reproduces the firing pattern of the biological ORN.

(A) The model activity (bottom trace) displays both isolated spikes and bursts of variable length as in the original recording (top trace). The simulated trace shows the membrane potential (an intracellular variable), while the biological data were obtained with extracellular electrodes. (B) Semi-logarithmic plot of the distribution of spikes per burst of an experimental LD recording (top) and simulated model (bottom) of spontaneous action potential activity of pheromone-sensitive neurons. The experimental distribution corresponds to the 1-hour period of highest firing frequency in the recording; the simulated distribution was obtained from a separate 1-hour simulation at fixed high Orco conductance. Both the model and the recordings display exponential decrease of the count’s density as a function of the burst length, shown as a linear decrease in the logarithmic scale. (C) Comparison of experimentally recorded and model-predicted circadian modulation of spiking attributes. Top panels show mean spike frequency (Ci, experimental; Cii, model), whereas bottom panels show the percentage of spikes belonging to bursts (Ciii, experimental; Civ, model). The simulated model reproduced the circadian dynamics observed in the experimental recordings, demonstrating good agreement between model and biological data. (D) Heat map of the simulated model, x-axis shows time (days), y-axis shows instantaneous frequency (Hz), and pixel color indicates count density on a logarithmic scale. The instantaneous frequencies can be roughly divided in a band of high frequencies (top, ∼50 – 80 Hz) and low frequencies (bottom, ∼0.1 – 5 Hz), with the high-frequency band representing the instantaneous frequencies of spikes within a burst and the low-frequency band the instantaneous frequency between bursts. (Di) Simulated spikes with circadian regulation by the Orco channel; the inter-burst frequencies varied widely with circadian rhythmicity while the frequencies within burst remained approximately constant. The simulated results also reproduced the “merging” effect between the two bands when the bottom one approaches the top one. (Dii) Simulated spikes without circadian regulation of Orco had no circadian oscillations.

The model successfully reproduces the circadian modulation of spiking activity: the maximum firing rate occurs when the Orco conductance is maximal, whereas the maximum percentage of spikes that are part of bursts correlates with the lowest Orco conductance, resulting in antiphase oscillations, as were observed in the experimental recordings (Figure 8C).

The frequency prevalences in the model over several days clustered in two distinct frequency bands: an upper band that corresponds to the intra-burst frequencies and a lower band, corresponding to inter-burst frequencies (Figure 8D), in agreement with the biological experiments (Figure 5A, B). When circadian oscillations in Orco conductance were included, inter-burst frequencies showed pronounced circadian variations, whereas intra-burst frequencies remained relatively constant, reminiscent of the biological data. Notably, the simulation captured the experimentally observed “merging” effect, where the lower frequency band approaches the upper band during high-activity periods (Figure 8D).

This merging of spiking and bursting frequencies suggests that the system approached a bifurcation condition coming from the distribution of spikes per burst (Figure 8B), computed from the 1 h period of highest firing frequency in an experimental recording and from a separate 1 h simulation at fixed high Orco conductance. In both cases, the distribution followed a decreasing exponential trend – in contrast to the constant value or bounded range expected from a standard bursting neuron model – consistent with a system operating near a bifurcation point. The observation that the circadian modulation of spiking activity requires only one parameter, namely changes in the Orco conductance, to describe the circadian effects on the ORN spiking pattern indicates the crucial role of the Orco channel in circadian control.

cAMP regulates the circadian function of Orco

Our computational model predicted that the daily/circadian oscillations in cAMP levels (Schendzielorz et al., 2015) control the circadian modulation of spontaneous spiking activity of hawkmoth ORNs. To test this prediction in hawkmoth ORNs, we increased cAMP levels by infusing 8-Br-cAMP-AM through the recording electrode into the sensillar lymph (Figure 9) at ZT 1-3, when naturally occurring cAMP levels are lowest (Schendzielorz et al., 2015) to maximize the effect. This significantly increased the mean frequency of the spontaneous spiking activity (Figure 9A; Wilcoxon signed-rank test Z = −2.701; p = 0.004). To check whether this increase in mean frequency was caused by direct or indirect interaction of cAMP with the Orco channel, we infused the Orco antagonist OLC15 together with 8-Br-cAMP-AM through the same recording electrode. Blocking Orco prevented a significant change in the spontaneous spiking frequency via the cAMP analogue (Figure 9B; paired t-test t(9) = −1.446; p = 0.182). This further corroborates cAMP-dependent modulation of the circadian pacemaker channel Orco.

cAMP modulates open-time probability of Orco in M. sexta.

Perfusion of 8-Br-cAMP-AM significantly increased the spontaneous spiking activity at the end of the activity phase (ZT 1-3; A; Wilcoxon signed-rank test Z = −2.701; p = 0.004). However, activity did not increase when 8-Br-cAMP-AM was added together with the Orco blocker OLC15 (B; paired t-test t(9) = −1.446; p = 0.182). Electrodes were filled with sensillum lymph ringer containing 0.1% and 0.15% DMSO as control in A and B, respectively. Data from individual animals are connected by lines.

Discussion

Searching for daily and circadian rhythms of Orco-dependent membrane potential control in olfactory receptor neurons (ORNs) of M. sexta, in vivo long-term tip recordings of pheromone-sensitive long trichoid sensilla of male hawkmoth antennae were performed. We detected interlinked circadian and ultradian frequencies in the spontaneous spiking activity, indicative of underlying membrane potential oscillations in ORNs. We found that the olfactory receptor co-receptor (Orco) acts as a pacemaker channel that controls the daily/circadian modulation of the spiking activity without affecting the ultradian gamma frequency range that is relevant for the primary fast, phasic pheromone response (Dolzer et al., 2003; Nolte et al., 2016; Schneider et al., 2025). Furthermore, since OLC15-dependent block of Orco did not delete ultradian action potential activity, other additional pacemaker channels must be present in ORNs that possibly control different frequency ranges of ultradian membrane potential oscillations. Since qPCR experiments at different ZTs showed that Orco transcription was not directly controlled by the circadian TTFL clock present in hawkmoth ORNs, we concluded that the daily/circadian control of Orco function is obtained at the post-translational level. Our computational model predicted that cAMP-dependent post-translational modulation of Orco conductance is sufficient to account for our electrophysiological findings if cAMP levels vary in a daily/circadian rhythm. Daily rhythms in cAMP levels are confirmed in hawkmoth antennae (Schendzielorz et al., 2015). Furthermore, here we provided evidence that cAMP analogs Orco-dependently control spontaneous activity of hawkmoth ORNs. Thus, we suggest that, colocalized in a membrane signalosome in the ORN soma compartment, constitutively depolarizing Orco homomeric ion channels generate ultradian membrane potential oscillations via delayed activation of hyperpolarizing K+ channels (Figure 10). The ultradian membrane potential oscillations express a daily/circadian rhythm with higher frequencies during the hawkmoth’s activity phase at night compared to the rest phase during the day. Thus, as novel hypothesis we suggest that the soma signalosome constitutes an endogenous, plastic, post-translational feedback loop (PTFL) clockwork, controlling membrane potential oscillations with Orco as 2nd messenger-dependent control hub that is coupled to, but not commanded by, the circadian TTFL clock in the nucleus.

A signalosome assembling an endogenous multiscale PTFL membrane clock in the soma of hawkmoth olfactory receptor neurons (ORNs) with 2nd messenger-dependent Orco as central control hub is proposed to mediate membrane potential rhythms.

Schematic of an antennal pheromone-sensitive ORN that is strictly compartmentalized by tight junctions (thick black borders) into three functionally separated membrane compartments: cilium, soma/dendrite, and axon. Soma/dendrite and axon are wrapped by supporting cells (grey background), ensuring common extracellular ionic concentrations for both membrane compartments. Thus, only the cilium is exposed to the sensillum lymph that contains high (∼200 mM) K+ concentration. The ciliary membrane comprises a specialized membrane signalosome devoted to highly sensitive pheromone detection. In contrast, a signalosome in the soma/dendrite controls the membrane potential and intracellular Ca2+ concentration. Sustained depolarization via opening of cation-permeable, homomeric Orco pacemaker channels in the soma are counteracted by delayed voltage-gated hyperpolarizing K+-channels, together generating ultradian oscillations in the membrane potential with interlinked oscillations in intracellular Ca2+. Circadian/daily rhythms in cAMP concentrations mediating Orco conductance, next to Ca2+, superimpose a circadian rhythm on prevalence of ultradian oscillations. Orco is not under direct circadian transcriptional control by the circadian TTFL clock in the nucleus that is entrained to the LD cycle, thus being part of a PTFL membrane clock entrained to other Zeitgebers. PTFL- and TTFL-clocks are coupled by not-yet identified mechanisms. LD: light dark cycle; OR: odor receptor, Orco: olfactory receptor coreceptor, PTFL: post-translational feedback loop clock; TTFL: transcriptional-translational feedback loop clock; Vm: membrane potential; sine waves: oscillations.

The expression of a 2nd messenger-dependent leaky pacemaker channel in ORNs, which does not require daily transcription, would constitute an economic and rapidly adjustable mechanism of active sensing. As part of a PTFL membrane clock, Orco could orchestrate endogenous membrane potential oscillations that are tuned to detect and resonate with behaviorally relevant temporal patterns of Zeitgeber signals, such as the circadian and ultradian pheromone fluctuations in the insect’s environment that are detected in the cilia of ORNs (Stengl and Schneider, 2024; Stengl and Schröder, 2021). The present study provides first evidence for a systems view of biological timing that involves multiscale PTFL membrane clocks linked to the TTFL clock in insect sensory neurons, instead of the prevailing hierarchical view that assumes all endogenous circadian oscillations are outputs of the genetic circadian TTFL clock (Stengl and Schneider, 2024; Stengl and Schröder, 2021). Future studies will challenge our novel hypothesis.

Orco in the ORN soma compartment acts as a pacemaker channel that controls the membrane potential oscillations underlying spontaneous spiking activity

The resting membrane potential of most silent neurons lies close to the negative equilibrium potential for K+ ions, mainly due to potassium leak channels and the driving force by the 10x higher intracellular K+ concentration (Goldstein et al., 2001; Hille, 2001; Patel and Honoré, 2001; Talley et al., 2001). Thus, spontaneously active neurons require the expression of Na+ and/or Ca2+ permeable leak or pacemaker channels that open at the hyperpolarized resting membrane potential, thus depolarizing the neuron via influx of cations along their concentration gradients (Bose et al., 2014; Cochet-Bissuel et al., 2014; Das et al., 2016; Golowasch et al., 2017; Lüthi and McCormick, 1998; Ratliff et al., 2021; Robinson and Siegelbaum, 2003; Sharma et al., 2023). When the pacemaker-driven depolarization reaches spike threshold, voltage-gated Na+ and K+ channels in the axon promote spiking. Furthermore, voltage-dependent Ca2+ channels allow for depolarization-dependent Ca2+ influx. Voltage- and Ca2+-gated hyperpolarizing K+ channels then repolarize the neuron to negative membrane potentials, re-starting the cycle. Furthermore, different fine-tuned mechanisms of homeostasis re-establish intracellular low (pico- to nanomolar range) Ca2+ concentrations, preventing necrosis. Thus, membrane potential oscillations with interlinked oscillations in intracellular Ca2+ levels drive spontaneous spiking activity, which depends on the type of pacemaker channel and antagonistic ion channels present.

In insect ORNs, at least two types of pacemaker channels exist: the highly conserved, slow, leaky Orco, which we focus on in this study, and an inversely voltage-dependent HCN type channel (Butterwick et al., 2018; Dolzer et al., 2021; Kodirov, 2022; Lee and MacKinnon, 2017; Mandala et al., 2025; Nolte et al., 2016, 2013; Sato et al., 2008; Stengl and Funk, 2013; Stengl and Schneider, 2024; Stengl and Schröder, 2021; Wicher et al., 2008; Wicher and Miazzi, 2021). When expressed in vitro in heterologous expression systems, Drosophila as well as Manduca Orco homo- and heteromers assemble into leaky, non-specific cation channels with a reversal potential around 0 mV, being constitutively permeable for Na+, K+, and Ca2+ (Nolte et al., 2016, 2013; Sato et al., 2008; Stengl and Funk, 2013; Wicher et al., 2008; Wicher and Miazzi, 2021). Thus, while this study did not aim to unravel the overall generation of spontaneous activity in ORNs, here, we examined whether the leak channel Orco in hawkmoth ORNs is the core pacemaker channel responsible for membrane potential oscillations that drive spontaneous circadian activity (Stengl, 2010; Stengl and Schneider, 2024).

Orco homo- or heteromers are likely to have different localizations and functions in the distinct, sealed-off membrane compartments of ORNs that face different ionic environments: the cilium (also termed “outer dendrite”), and the soma / (“inner”) dendrite (Stengl and Funk, 2013; Stengl and Schröder, 2021). We propose for Manduca, that homomeric Orco pacemaker channels, as core element of a PTFL membrane clock, are restricted to the soma/dendrite compartment of ORNs, where they face the same extracellular cation concentrations as other brain neurons, with low K+, high Na+, and high Ca2+ concentrations (Stengl and Schneider, 2024; Stengl and Schröder, 2021). In contrast, we expect Manduca Orco heteromers to serve as obligatory chaperons to localize ORs to cilia and maintain them in the ciliary membranes, as proven in Drosophila (Benton et al., 2006). Since the ciliary compartment faces very high concentrations of K+ in the sensillum lymph, matching intracellular K+ levels (Kaissling and Thorson, 1980; Steinbrecht, 1992), we propose that Orco channels in the ciliary membrane remain closed. If Orco homo- or heteromers in cilia would constitute leaky pacemaker channels, their K+ permeability would clamp the ciliary membranes at 0 mV, due to the high extracellular K+ concentration, strongly reducing the driving force of pheromone-dependent receptor currents, severely compromising pheromone sensitivity (Stengl and Funk, 2013). Accordingly, it was shown previously in hawkmoth ORNs that Ca2+ channels, most likely of the TRP ion channel family, serve as transduction channels for the primary events of the highly sensitive pheromone transduction, and not a non-specific cation channel like Orco (Gawalek and Stengl, 2018; Nolte et al., 2016, 2013; Stengl, 1994, 1993).

In support of the specific localization of Orco homo- or heteromers, Orco antisera labeled Orco in two of the three sealed off ORN membrane compartments in both Drosophila (Benton et al., 2006) and Manduca (Nolte et al., 2016): in the membrane of the soma/dendrite compartment and in the membrane of the cilium, but not in the axons. In contrast, antisera against an OR in Drosophila detected ORs accumulated only in ciliary membranes but not in the other compartments of ORNs (Benton et al., 2006). Thus, the localization of OR-Orco heteromers in cilia confirm that Orco is an obligatory coreceptor and chaperone of ORs localizing and maintaining the OR-Orco complexes in the ciliary membrane (Benton et al., 2006). Since antisera against Manduca ORs are not available, we do not know for hawkmoths whether the pheromone-sensitive OR-Orco heteromers are located to ciliary membranes, while Orco homomers accumulate in the soma/dendrite compartment (Fandino et al., 2019). Nevertheless, Orco is likely to assemble as homomeric complexes in wild-type hawkmoth ORNs because Orco expression starts during development, coinciding with the beginning of spontaneous activity, well before ORs are present. Furthermore, the Orco agonist VUAA1 activates Orco in primary cell cultures of maturing ORNs before they become pheromone sensitive (Nolte et al., 2016; Schweitzer et al., 1976).

In summary, we propose that Orco homomers are specifically located to a signalosome in the soma/dendrite compartment of ORNs (Figure 10), controlling the ORN’s membrane potential. Because the Orco-dependent ultradian and circadian membrane potential oscillations are maintained under constant conditions, they are generated by an endogenous multiscale clock. Thus, we predict that a distinct signalosome with Orco homomers in the soma, but not in the cilia, serves as endogenous self-assembling multiscale PTFL membrane clock for the circadian control of the membrane potential. We hypothesize that this distinctly located PTFL membrane clock is coupled to, but not dominated by, the TTFL clock in the hawkmoth ORNs to allow for fast, plastic regulation of the membrane potential (Schuckel et al., 2007; Stengl and Schneider, 2024).

Orco is hypothesized to be a second messenger-dependent control hub of a PTFL membrane clock that regulates circadian membrane potential oscillations in hawkmoth ORNs

Focusing on the signalosome in the ORN soma compartment, our novel hypothesis predicts that Orco is a 2nd messenger-dependent control hub of a somatic multiscale PTFL clock. In accordance with this hypothesis, different 2nd messenger cascades regulate Drosophila Orco via a Ca2+/calmodulin binding site and via phosphorylation/de-phosphorylation at multiple sites that are conserved between species (Cao et al., 2016; Getahun et al., 2016; Guo and Smith, 2017; Mukunda et al., 2016, 2014; Sargsyan et al., 2011). In contrast to our current focus of Orco homomer function in Manduca, focus and data interpretation in Drosophila lies on the function of Orco heteromers. In Drosophila, Orco’s Ca2+/calmodulin binding site controls the localization of the OR-Orco complex in the ciliary membrane of ORNs, provides for Ca2+-dependent negative feedback control of Orco conductance, and depends on previous odor stimulation (Bahk and Jones, 2016; Cao et al., 2016; Guo and Smith, 2022; Jain et al., 2021). After phosphorylation of Orco by protein kinase C (PKC), Drosophila Orco directly or indirectly becomes cyclic nucleotide-dependent, with cGMP and cAMP increasing Orco-dependent depolarization and Ca2+ influx (Getahun et al., 2013; Wicher et al., 2008). However, it is not yet resolved in Drosophila in which of the different sensilla compartments and at what time of day Orco homo- or heteromers are controlled via 2nd messengers and how this control is regulated.

Also, in Manduca the 2nd messenger-dependent modulation of Orco is not yet understood. However, due to Orco’s highly conserved structure we assume Ca2+/calmodulin-, PKC-, and cyclic nucleotide-dependent control of hawkmoth Orco localization and/or open-time probability, likely similar as described for Drosophila Orco. It is very likely that the conserved structure of Orco with its manifold 2nd messenger-modulation sites designate it as a hub for intersecting 2nd messenger control in different insect species. Since blocking Orco in M. sexta decreased the spontaneous spiking activity in ORNs and deleted its circadian modulation, Orco is apparently the main target for circadian clock control modulating circadian membrane potential rhythms. Furthermore, the ultradian rhythms in spontaneous activity appear to be controlled by additional pacemaker channels that are not focus of the current study. Since Orco transcript abundance was constant throughout the day, Orco is not controlled directly by the circadian TTFL clock. Rather, open-time probability and/or membrane localization of Orco appear to be controlled in a circadian manner via post-translational mechanisms. Because daily rhythms in cAMP levels were found in hawkmoth antennae (Schendzielorz et al., 2015), our modeling predicted that the most parsimonious explanation for Orco’s circadian control of the membrane potential is Orco’s regulation via daily rhythms of cAMP levels. In support of this hypothesis, we demonstrated that the significant increase of spontaneous spiking activity caused by 8-Br-cAMP application is canceled with co-application of the Orco blocker OLC15. Furthermore, our modeling studies showed that this post-translational circadian control of Orco via daily/circadian levels of cAMP is sufficient to explain our physiological findings. Thus, we hypothesize that a PTFL membrane clock (Figure 10), linked to the TTFL nuclear clock, determines the circadian oscillations in pheromone sensitivity and temporal resolution of ORNs via cAMP- and Orco-dependent membrane potential control (Dolzer et al., 2021, 2008; Flecke et al., 2010, 2006; Redkozubov, 2000; Schendzielorz et al., 2015, 2012; Ziegelberger et al., 1990).

Our novel hypothesis predicts that a signalosome in the soma membrane of hawkmoth ORNs assembles a circadian PTFL membrane clock controlling daily/circadian rhythms in the membrane potential and Ca2+ concentration, comprising Orco as central control hub. Via unknown mechanisms that still need to be resolved, circadian oscillations in cAMP levels are generated in ORNs that are linked to Orco. As is characteristic for a clock (Stengl and Schneider, 2024), the PTFL membrane clock comprises positive feedforward elements that upregulate the ORN’s depolarization and intracellular Ca2+ levels, which are then sequentially downregulated via delayed negative feedback elements/mechanisms. Thus, the PTFL clock generates oscillations in the membrane potential coupled to strictly compartmentalized oscillations in intracellular Ca2+ levels. This predicted soma signalosome comprises the pacemaker channel Orco as a central hub that is controlled via changes in voltage, Ca2+, and cAMP levels that can be affected by multiple parallel pathways in the ORN. We propose a parsimonious model of the PTFL clock, generating multiscale membrane potential- and Ca2+ oscillations linked to cAMP oscillations. Since it is a PTFL clock, it does not require circadian degradation and circadian transcription/translation of its central constituents. Nevertheless, we predict coupling between the multiscale PTFL membrane clock and the TTFL circadian clock in the ORN’s nucleus to obtain stably synchronized circadian rhythms. As likely mechanism of coupling we predict that Ca2+- and cAMP-dependent kinases interlink both types of circadian clocks, thereby obtaining robust and at the same time flexible interlinked cellular rhythms (Figure 10). This system of interlinked circadian clocks are hypothesized to determine changing physiological setpoints of the ORN during circadian sleep-wake cycles (Stengl and Schneider, 2024).

Circadian and ultradian rhythms in spontaneous ORN activity resemble the physiologically relevant time scales for active sensing

Moths have daily rhythms in pheromone sensitivity and temporal resolution controlled via antennal circadian clocks (Flecke et al., 2010; Flecke and Stengl, 2009; Merlin et al., 2007, 2006; Schendzielorz et al., 2015; Schuckel et al., 2007; Silvegren et al., 2005). The increased pheromone sensitivity and temporal resolution during the activity phase enhances the male’s chance to detect and locate a calling female. In response to brief pheromone pulses, independent of Orco, pheromone concentration is encoded in the frequency of the first, phasic burst of spikes of the characteristic triphasic spiking sequence of ORNs (Dolzer et al., 2003; Nolte et al., 2013; Schneider et al., 2025). Because blocking Orco does not affect the ultradian gamma frequency band of spontaneous activity that correlates with the frequency ranges of the stimulus concentration-encoding the first, phasic pheromone response, the current finding further confirms that Orco is not the primary transduction channel for pheromone detection (Dolzer et al., 2003; Nolte et al., 2016, 2013). Nevertheless, Orco partakes in pheromone transduction during the later and longer-lasting time window after pheromone stimulation. Mediated through the cAMP-dependent increase in Orco conductance during the activity phase, the ORN membrane potential would oscillate pheromone-dependently at higher ultradian frequencies and with larger amplitude from a lower negative baseline resting potential. Thereby, the ORN’s pheromone sensitivity and sampling frequency of pheromone pulses could be maximized while males fly upwind in search for females.

The upregulation of neural activity during wakefulness/alertness seems to be ubiquitous among different species. For example, the synchronously recorded neurons in human EEG show fast ultradian oscillations in the alpha, beta, and gamma frequency ranges during alertness and slow ultradian oscillations in the delta and theta ranges during rest and sleep. Similarly, in the lower frequency band, the spontaneous spiking activity of hawkmoth ORNs changes from below delta during rest to theta/alpha ranges during their activity phase. While the gamma frequency band comprises spike frequency within a burst and matches the first phasic pheromone response (Dolzer et al., 2003), the spontaneous spiking activity in the theta/alpha range matches the maximal temporal resolution of pheromone pulses of up to 10 Hz (Marion-Poll and Tobin, 1992), being ideally suited to tune ORNs for active sensing. Even though oscillations in the field potential of electroantennogram responses to odor stimulation can reach 100 Hz in different insects (Szyszka et al., 2014), these frequencies most likely do not encode the temporal resolution of physiological odor responses. Apparently, they reflect the astounding adaptive ability of antennal ORNs to synchronize their endogenous membrane potential oscillations across the antenna to a broad range of stimulation frequencies. Pheromone pulse resolution of about 100 Hz was never reported in single olfactory sensilla recordings, and in behavioral experiments male moths stop their upwind mate search if pheromone pulses exceed 30 Hz, apparently due to adaptation (Baker et al., 1985; Bau et al., 2005, 2002; Tripathy et al., 2010).

In summary, Orco-dependent circadian oscillations in the ORN membrane potential are ideally suited to autonomously, fast, and flexibly tune sensitivity and temporal resolution of the pheromone-sensitive ORNs to the hawkmoth’s rest-activity cycle and to acute physiological requirements, synchronizing male and female mating behavior. Here, we have demonstrated how modulation of the leaky pacemaker channel Orco can achieve these circadian oscillations. Whether cAMP levels are under control of the TTFL clock or oscillate based on Ca2+ - and/or octopamine concentration oscillations, being part of a stand-alone PTFL clock, remains to be studied.

Stochastic ion channel dynamics and circadian modulation of Orco conductance explain modeled ORN firing patterns

Our modeling results provide mechanistic insight into the irregular spontaneous bursting activity of pheromone-sensitive ORNs in M. sexta, highlighting stochastic ion channel dynamics and circadian modulation of ion conductance that together shape spiking patterns. By employing a Langevin approximation of the stochastic HH formalism, we demonstrate that the intrinsic noise arising from the random gating of ion channels is sufficient to generate both isolated spikes and bursts of spikes, as observed experimentally. Our model effectively recapitulates the spike distributions found in vivo and accounts for the role of channel noise as a critical determinant of ORN firing patterns.

A key contribution of our model is the incorporation of the Orco ion channel as a pacemaker component, with a conductance that varies linearly with cAMP concentration. We show that post-translational circadian modulation of Orco conductance alone is sufficient to produce the observed time-of-day dependent changes in spike distribution. This finding suggests that the rhythmic expression of cAMP levels across the circadian and/or daily cycle may act as a primary driver of circadian and diurnal changes in ORN excitability, without requiring additional external inputs or modulation of other ion channels.

Importantly, our results reveal that the same set of ion channels can support the emergence of both single action potentials and complex bursting behaviors purely through the stochastic variability of their gating kinetics. This unifying explanation reduces the need for invoking separate biophysical mechanisms or circuit-level inputs to explain these distinct firing modes. For simplicity, the circadian variation in cAMP concentration was modeled using a sinusoidal function. While this approximation is sufficient for capturing the qualitative features of the rhythmic modulation, future work could enhance the biological realism of the model by coupling Orco conductance dynamics to an endogenous circadian oscillator, such as a Goodwin-type TTFL. Incorporating this layer would allow for a more detailed exploration of how endogenous clock components regulate cAMP synthesis and degradation. Additionally, further extensions could include downstream regulatory or coupling mechanisms such as phosphorylation and dephosphorylation events mediated by kinases and phosphatases under circadian control, potentially contributing to fine-tuned modulation of ion channel function over the daily cycle. Together, these findings demonstrate the value of a biophysically grounded, noise-inclusive modeling framework for understanding complex temporal patterns in sensory neuron activity and open the door for future integrative models linking different types of endogenous circadian clocks to cellular excitability.